---
title: "Concurrent IBE and IBR in terradish"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Concurrent IBE and IBR in terradish}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4.5, warning = FALSE, message = FALSE)
```

## What the joint model estimates

**Isolation by resistance (IBR)** associates genetic structure with resistance
through a landscape graph. **Isolation by environment (IBE)** associates genetic
structure with environmental differences between sampled sites. Fitting both
estimates each association conditional on the other predictors. It does not
allocate genetic variation to distinct causal mechanisms.

This vignette builds one pairwise design containing geographic distance and
scaled environmental differences, then uses it in both measurement families.
A small synthetic example keeps the complete workflow runnable. See
`vignette("getting-started", package = "terradish")` for preparing real rasters
and `vignette("wishart-covariance", package = "terradish")` for genotype summaries.

The two families connect the same pairwise differences to their responses:

| Component | MLPE distance regression | Wishart contrast covariance |
|----|----|----|
| Landscape | Resistance `R(theta)` with slope `beta` | Graph covariance `E(theta)` with weight `tau` |
| Pairwise difference `D_k` | Additive mean term `gamma_k * D_k` | Centered kernel `K_k = -H D_k H / 2` with weight `lambda_k` |
| Environmental constraint | `gamma_k` can have either sign | `lambda_k >= 0` |
| Remaining variation | Intercept, residual precision, shared-site correlation | Common covariance nugget |
| Effective information | Determined by the fitted MLPE likelihood | Analyst supplies `nu` |

Here `H` centers sites. For the corresponding covariance-derived distances,
the Wishart model implies the additive mean
`tau * R + sum(lambda_k * D_k) + 2 * exp(sigma)`. This makes the distance
interpretation comparable to MLPE, although the likelihoods and constraints
differ. The default `absdiff` kernel is not an outer product of site values;
the `sqdiff` transform gives that special case.

## 1. Build a landscape and one pairwise design

```{r landscape}
library(terradish)
library(terra)

# A small projected grid with one continuous landscape covariate.
r <- rast(nrows = 8, ncols = 8, xmin = 0, xmax = 8,
          ymin = 0, ymax = 8, crs = "EPSG:3857")
values(r) <- sin(seq_len(64) / 5) + seq_len(64) / 64
names(r) <- "x"
coords <- xyFromCell(r, c(1, 5, 8, 12, 20, 27, 33, 37, 45, 53, 57, 64))
surface <- conductance_surface(r, coords, directions = 8, saveStack = TRUE)

# Environmental values are site attributes. They need not be graph covariates.
environment <- data.frame(climate = sin(seq_len(nrow(coords))))
environment_pairs <- pairwise_endpoint_covariates(
  environment, transform = "absdiff", scale = TRUE
)
z <- pairwise_covariates(environment_pairs, geographic = dist(coords))
head(z)
```

**How to read the output.** Each row is an unordered site pair in R's
lower-triangle order. The climate column is the absolute difference after
standardizing site values; one unit is one site-level standard deviation of
difference. Geographic distance retains the coordinate units, meters here.
`pairwise_covariates()` does not scale either input again. In real studies,
use a projection suitable for the study region and retain the same site order
in the response, graph, and pairwise design.

The supported transforms are `absdiff`, `sqdiff`, `euclidean`, and `manhattan`.
Choose the transform before fitting and retain it across families. A Wishart
kernel must be positive semidefinite in site contrasts; the constructor rejects
incompatible pairwise geometry rather than silently correcting it.

```{r measurement-models}
g_mlpe <- mlpe_covariates(z)
g_wishart <- wishart_covariates(z, model = "wishart_covariance")
```

These are measurement-model factories. Their coefficients are estimated when
the resistance model is fitted. Both retain the original pairwise design so
site subsets do not independently restandardize environmental values.

## 2. Create covariance and matching distance responses

For an interpretable comparison between families, use the same covariance
information represented in two forms. This example draws from a declared
Wishart model with landscape, climate, and geographic components. It does
not imply that a generic genetic distance or F~ST~ follows that distribution.

```{r response}
nu_demo <- 1000
base <- simulate_covariance_response(c(x = 0.8), ~x, surface,
  tau = 1, sigma = 0.1, nu = nu_demo, seed = 301)
kernels <- attr(g_wishart, "kernel_covariates")
Sigma <- base$Sigma + 0.1 * kernels[, , 1] + 0.05 * kernels[, , 2]
set.seed(301)
S_cov <- stats::rWishart(1, df = nu_demo, Sigma = Sigma)[, , 1] / nu_demo
S_dist <- dist_from_cov(S_cov)
check_distance_response(S_dist)$admissible
```

The diagnostic checks distance geometry. Here the generating model also
supplies a sampling justification. For real data, assess that justification
separately. `nu_demo` is an illustrative simulation setting, not a recommendation
to use the number of SNPs as effective degrees of freedom.

## 3. Fit the candidate models within each family

The uniform surface still implies resistance among sites. Adding pairwise
environmental terms to it is therefore **IBD + IBE (uniform surface)**, not
an IBE-only model. Geographic distance is an additional explicit measurement
predictor; uniform-graph resistance is generally not Euclidean distance.

```{r fit-models}
ctrl <- NewtonRaphsonControl(maxit = 100, verbose = FALSE)
fit_ibd <- terradish(S_dist ~ 1, surface,
  measurement_model = mlpe, control = ctrl)
fit_ibe <- terradish(S_dist ~ 1, surface,
  measurement_model = g_mlpe, control = ctrl)
fit_ibr <- terradish(S_dist ~ x, surface,
  measurement_model = mlpe, control = ctrl)
fit_joint <- terradish(S_dist ~ x, surface,
  measurement_model = g_mlpe, control = ctrl)

fit_w_ibd <- terradish(S_cov ~ 1, surface,
  measurement_model = wishart_covariance, nu = nu_demo, control = ctrl)
fit_w_ibe <- terradish(S_cov ~ 1, surface,
  measurement_model = g_wishart, nu = nu_demo, control = ctrl)
fit_w_ibr <- terradish(S_cov ~ x, surface,
  measurement_model = wishart_covariance, nu = nu_demo, control = ctrl)
fit_w_joint <- terradish(S_cov ~ x, surface,
  measurement_model = g_wishart, nu = nu_demo, control = ctrl)

fits <- list(mlpe_ibd = fit_ibd, mlpe_uniform_ibe = fit_ibe,
  mlpe_ibr = fit_ibr, mlpe_joint = fit_joint,
  wishart_ibd = fit_w_ibd, wishart_uniform_ibe = fit_w_ibe,
  wishart_ibr = fit_w_ibr, wishart_joint = fit_w_joint)
vapply(fits, function(fit) fit$convergence$code, numeric(1))
```

Interpret code 0 as numerical convergence under the stopping criteria.
Inspect nonzero codes, gradients, inner nuisance convergence, and active
bounds before comparing fits. A no-resistance boundary makes the conductance
surface unidentified; a zero environmental weight alone does not.

## 4. Read conditional conductance and environmental associations

```{r conditional-coefficients}
rbind(MLPE_without_pairs = coef(fit_ibr), MLPE_with_pairs = coef(fit_joint),
      Wishart_without_pairs = coef(fit_w_ibr), Wishart_with_pairs = coef(fit_w_joint))
summary(fit_joint)
summary(fit_w_joint)
```

**How to read the output.** A positive `x` coefficient means greater relative
conductance as `x` increases, conditional on the included pairwise terms.
The estimated coefficient can change after adding those terms because the
model can represent the same spatial pattern in more than one way. Neither
surface is a unique causal map of landscape effects.

In MLPE, an environmental `gamma` gives the conditional change in genetic
distance per unit of pairwise difference. It is unconstrained, so a negative
estimate is possible. In Wishart, `lambda` is a nonnegative kernel weight.
An association in the opposite direction is represented at the zero bound.
The Wishart summary marks that boundary and gives an approximate one-sided
upper limit rather than a symmetric interval.

```{r response-change}
mlpe_response_change(fit_joint)
```

This reports the MLPE response change over the stored 10th-to-90th-percentile
pairwise contrast. The nuisance inference includes uncertainty in the fitted
conductance parameters. It remains conditional on the selected model and
measurement assumptions.

### Compare IBE:IBR ratios

```{r ratios}
ratios <- rbind(terradish_ibe_ratio(fit_joint),
                terradish_ibe_ratio(fit_w_joint))
ratios
```

The ratio is `gamma / beta` for MLPE and `lambda / tau` for Wishart. It is
the resistance-distance equivalent of one unit of the pairwise difference.
These ratios share an interpretation because the response is the same
covariance-derived distance and both families use the same pairwise transform
and scaling. Their absolute coefficients are not comparable across likelihoods.
Each fit also estimates its own conductance surface, so its resistance scale
can differ. The ratio is undefined when the resistance coefficient is zero.
The geographic row has the same equivalent-distance interpretation but is
an explicit IBD term, not an environmental mechanism.

An environmental association can arise from local adaptation, historical
structure, correlated geography, sampling, or another omitted predictor.
The fitted term does not distinguish these mechanisms. A precise ratio alone
does not establish causal IBE.

## 5. Compare predictive support and sensitivity

Information criteria are computed separately within each likelihood family:

```{r information-criteria}
labels <- c("IBD", "IBD + IBE (uniform surface)", "IBR", "IBR + pairwise terms")
aic_table(list(fit_ibd, fit_ibe, fit_ibr, fit_joint), mod_names = labels)
aic_table(list(fit_w_ibd, fit_w_ibe, fit_w_ibr, fit_w_joint), mod_names = labels)
```

Lower AIC balances fit and parameter count within the declared candidate set.
Wishart AIC can over-select environmental kernels when `nu` overstates effective
information. AICc does not repair that problem. Cross-family AIC comparisons
have no common likelihood basis.

### Nested tests and boundaries

```{r nested-tests}
anova(fit_ibr, fit_joint)
anova(fit_w_ibr, fit_w_joint)
```

Adding unconstrained MLPE slopes uses the ordinary chi-square reference under
its asymptotic assumptions. Adding only nonnegative Wishart kernel weights
uses a chi-bar-square reference with binomial weights. This reference is exact
for information-orthogonal kernels and approximate otherwise. The climatic
and geographic kernels here need not be orthogonal. Wishart test magnitudes
also depend on supplied `nu`; treat them as descriptive at that value.
The uniform-surface IBE model and the IBR model are not nested in each other.

```{r nu-sensitivity}
fit_w_lower_nu <- terradish_rescale_nu(fit_w_joint, nu = 50)
rbind(primary = coef(fit_w_joint), sensitivity = coef(fit_w_lower_nu))
terradish_ibe_ratio(fit_w_lower_nu)
```

Rescaling retains coefficient estimates and broadens uncertainty when `nu`
decreases. Rescale every candidate to the same value for a model comparison.
This is a sensitivity calculation, not estimation of effective information.

### Fixed-nuisance cross-validation

```{r fixed-cv, eval = FALSE}
# Use the identical site folds for the plain and extended measurement models.
folds <- terradish_folds(coords, k = 3, method = "random", seed = 42)
cv_mlpe <- terradish_cv_folds(surface,
  formulas = list(IBR = S_dist ~ x, joint = S_dist ~ x), folds = folds,
  model = list(IBR = mlpe, joint = g_mlpe), nuisance = "fixed", control = ctrl)
cv_wishart <- terradish_cv_folds(surface,
  formulas = list(IBR = S_cov ~ x, joint = S_cov ~ x), folds = folds,
  model = list(IBR = wishart_covariance, joint = g_wishart),
  nu = nu_demo, nuisance = "fixed", control = ctrl)
cv_mlpe$summary
cv_wishart$summary
```

All nuisance coefficients, including pairwise terms, are learned from training
sites and held fixed during scoring. Compare candidates within each family,
using paired differences on common successful folds. Reprofiling on test sites
would instead assess performance after local recalibration. For spatial
transfer, use defensible spatial blocks and more sites than this teaching
example. Repeated folds expose partition sensitivity without creating new
independent observations.

## 6. Visualize the joint model

```{r fit-plot, fig.cap = "Observed and fitted pairwise distances for the joint MLPE model."}
plot(fit_joint, type = "fit")
```

Agreement describes the fitted response summary. It does not show that the
two spatial predictors identify distinct processes.

```{r surface-plot, fig.cap = "Relative conductance conditional on the pairwise climate and geographic terms."}
plot(fit_joint, type = "surface", data = surface)
```

The map and its intervals concern the conductance component after including
pairwise terms. Compare it with the fit omitting those terms before attributing
the spatial pattern to landscape `x`.

```{r marginal-plot, fig.cap = "Conditional conductance association with landscape x."}
plot(fit_joint, type = "marginal", data = surface)
```

For weakly sampled tails, compare unclamped predictions with
`support = "focal"`. Clamping uses focal-cell covariate limits and changes
prediction, not the fitted likelihood or the evidence for a mechanism.

## Quick reference

```{r quick-reference, eval = FALSE}
# 1. Preserve identical transforms and site ordering across families.
z <- pairwise_covariates(
  pairwise_endpoint_covariates(environment, transform = "absdiff", scale = TRUE),
  geographic = dist(coords)
)
g_mlpe <- mlpe_covariates(z)
g_wishart <- wishart_covariates(z, model = "wishart_covariance")

# 2. Use covariance-derived distances when comparing ratio interpretations.
S_dist <- dist_from_cov(S_cov)
fit_m <- terradish(S_dist ~ x, surface, measurement_model = g_mlpe)
fit_w <- terradish(S_cov ~ x, surface, measurement_model = g_wishart, nu = nu_demo)

# 3. Inspect convergence, conditional surface changes, and ratio uncertainty.
summary(fit_m)
summary(fit_w)
rbind(terradish_ibe_ratio(fit_m), terradish_ibe_ratio(fit_w))

# 4. Assess predictive support and sensitivity to effective information.
terradish_ibe_ratio(terradish_rescale_nu(fit_w, nu = 50))
```

## Summary of key functions

| Function | Purpose |
|----|----|
| `pairwise_endpoint_covariates()` | Transform site attributes into pairwise differences |
| `pairwise_covariates()` | Combine those differences with geographic or other pairwise matrices |
| `mlpe_covariates()`, `wishart_covariates()` | Use the same design in two likelihood families |
| `terradish_ibe_ratio()` | Report resistance-equivalent ratios with joint uncertainty |
| `mlpe_response_change()` | Express an MLPE pairwise association on the response scale |
| `terradish_rescale_nu()` | Assess Wishart precision sensitivity without refitting |
| `terradish_cv_folds()` | Evaluate held-out prediction on a fixed landscape graph |

## See also

- `vignette("getting-started", package = "terradish")`: graph preparation and conductance interpretation.
- `vignette("wishart-covariance", package = "terradish")`: covariance construction and effective information.
- `vignette("model-comparison", package = "terradish")`: nesting, repeated folds, and paired comparisons.
- `vignette("simulation-design", package = "terradish")`: recovery and coverage under declared scenarios.
