---
title: "Simulating Covariance Data to Plan a Study"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Simulating Covariance Data to Plan a Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4.5,
  warning  = FALSE,
  message  = FALSE
)
```

## Why simulate before you sample?

Landscape-genetic studies are expensive: every additional sampling site means
field time and cost. Before
committing to a design it is worth asking, in advance:

- **How many sampling sites** do I need to detect the conditional conductance
  association specified in my scenario?
- **How much approximately independent marker information** (represented by
  the supplied `nu`) do I need?
- **Where should the sites go**: spread evenly across the landscape, or placed
  some other way?
- If I run the analysis, **how often will I recover the generating parameters
  under this model**, and how often will sampling variation obscure them?

`terradish` answers these questions by simulation. Because the Wishart
covariance model has a tractable generative form, you can draw synthetic
covariance matrices from a *known* conductance surface, refit the model, and
measure how reliably the generating parameters are recovered, all without any forward-time
genetic simulation. This vignette shows how.

This is a **design tool**, not a replacement for individual-based or
coalescent simulation. It assumes the fitted model is correctly specified and
asks: given that model and this set of scenarios, how much data do I need? The
answer is conditional on the graph, parameters, candidate set, likelihood,
optimizer, and sampling strategies evaluated here.

---

## The generative model

A covariance-based `terradish` fit (see
`vignette("wishart-covariance", package = "terradish")`) assumes the observed
genetic covariance matrix `S` is a noisy realization of a structured
covariance:

$$\Sigma = \tau \, E(\theta) + \sigma\, I, \qquad
  \nu\, S \sim \mathrm{Wishart}(\nu, \Sigma)$$

where:

- $E(\theta)$ is the **conductance-implied covariance**: the generalized
  inverse of the graph Laplacian at conductance parameters $\theta$. This is
  the part that carries the landscape signal.
- $\tau \ge 0$ scales that signal.
- $\sigma \ge 0$ is a diagonal residual-variance term for covariance variation
  not represented by $E(\theta)$. It is not a direct estimate of drift,
  population size, or a Gaussian smoothing scale.
- $\nu$ is the supplied **effective Wishart degrees of freedom**, not a fitted
  parameter. It can be interpreted as the effective number of approximately
  independent marker contributions. Large $\nu$ gives draws tightly clustered
  around $\Sigma$; small $\nu$ gives noisier draws.

`simulate_covariance_response()` draws from exactly this model, and
`covariance_response_power()` automates the simulate-refit-summarize loop across
many designs. Fitted `wishart_covariance()` models store the nugget internally
on the log scale, but both simulation helpers accept $\sigma$ as the
nonnegative variance added to the diagonal.

For a fixed landscape template, $\tau / \sigma$ is a convenient
**signal-to-nugget ratio**. Larger values place more weight on the
landscape-implied covariance relative to diagonal noise. The ratio should be
interpreted within a fixed template because rescaling $E(\theta)$ also changes
its meaning.

> **Keep `nu` front of mind.** As explained in
> `vignette("wishart-covariance", package = "terradish")`, `nu` acts as an
> *effective sample size*. It does not change the *point estimates* recovered
> from a given covariance matrix, but it controls their precision (standard
> errors scale as $1/\sqrt{\nu}$) and therefore the **statistical power** to
> detect a nonzero coefficient under the simulated model. Sample size and
> marker design are important study-design levers, but the effective `nu` may
> be smaller than the raw marker count because of linkage and shared
> population history. No marker-count rule automatically establishes it.
> Distinguish the value that generates variation (`nu`) from the value
> assumed when fitting (`nu_fit`).

---

## Set up a landscape template

We need a landscape to define the conductance surface. Any
`conductance_surface()` works; here we reuse the bundled *Melipona* rasters as
a realistic template. Build the surface with `saveStack = TRUE` so the
simulation tools can subset focal sites efficiently.

```{r template}
library(terradish)
library(terra)

data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

covariates <- c(melip.altitude, melip.forestcover)
names(covariates) <- c("altitude", "forestcover")
covariates <- scale_covariates(covariates)

surface <- conductance_surface(covariates, melip.coords,
                               directions = 8, saveStack = TRUE)

cat("Candidate focal sites:", length(surface$demes), "\n")
```

We also fix the **generating parameters** we want to recover: a moderate
positive forest-cover coefficient and a moderate negative altitude coefficient
for log conductance.
These are illustrative simulation settings, not estimates from the bundled
*Melipona* genetic response. The bundled rasters provide a realistic landscape
template, while the synthetic covariance response is generated from the values
below.

```{r truth}
theta_true <- c(forestcover = 0.6, altitude = -0.4)
tau_true   <- 2   # strength of the landscape signal
sigma_true <- 1   # nugget (non-landscape variance)
```

---

## A single simulated data set

`simulate_covariance_response()` returns one covariance matrix (plus the
ingredients used to build it). It has the same matrix form as a genotype
covariance summary, but its sampling distribution is generated by the model.
Real genotype summaries need not follow that distribution.

```{r single-sim}
sim <- simulate_covariance_response(
  theta             = theta_true,
  formula           = ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  tau               = tau_true,
  sigma             = sigma_true,
  nu                = 100,     # 100 effective independent marker contributions
  seed              = 1
)

dim(sim$covariance)        # one row/column per focal site
round(sim$covariance[1:4, 1:4], 3)
```

Refitting estimates the parameters used to generate the data. One draw may
depart substantially from the generating values:

```{r single-fit}
fit <- terradish(
  sim$covariance ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = wishart_covariance,
  nu                = 100
)

rbind(true      = theta_true,
      estimated = round(coef(fit), 3))
```

A single replicate only tells you what happened *once*. To characterize a
design you must repeat the draw many times.

---

## Separate generating information from assumed information

`nu` controls variation among simulated covariance matrices. `nu_fit` controls
the precision assumed by the fitted likelihood; its default is `nu`. Raising
`nu_fit` while keeping `nu` fixed does not create more informative data. It
can make intervals narrower and increase nominal detections while reducing
coverage. Evaluate that mismatch explicitly:

```{r nu-fit-sensitivity, eval = FALSE}
# These runs use the same seeds and therefore the same simulated responses.
# Increase nsim after checking runtime and convergence on a small pilot.
design <- list(
  theta = theta_true, formula = ~ forestcover + altitude, data = surface,
  sample_sizes = 20, strategies = "spacefill",
  tau = tau_true, sigma = sigma_true, nu = 100,
  nsim = 100, seed = 905,
  control = NewtonRaphsonControl(maxit = 100)
)
matched <- do.call(covariance_response_power, design)
overstated <- do.call(covariance_response_power,
                     c(design, list(nu_fit = 2000)))
rbind(matched = matched$parameter_summary,
      overstated = overstated$parameter_summary)
```

**How to read the output.** Compare fit rate, coverage, and power together.
Similar estimates with narrower intervals under larger `nu_fit` reflect the
assumed precision, not better recovery of the generating process. A high
detection rate accompanied by poor coverage is not a successful design.
Failed and nonconverged fits count as non-detections; inspect their frequency
before interpreting power. Coverage and bias summaries require checking which
replicates supplied valid estimates.

These are runnable study-design recipes rather than stored results. Long
power sweeps are not run during vignette builds, and no hidden precomputed
files are needed to render this guide. Save your configuration, seeds, full
assessment objects, package versions, and convergence diagnostics when running
them. Report Monte Carlo uncertainty: for a proportion `p` estimated from
`B` independent replicates, its approximate Monte Carlo standard error is
`sqrt(p * (1 - p) / B)`.

For covariance simulation, the generating `nu` must be at least the covariance
dimension required by `stats::rWishart()`. That simulation constraint does not
make raw marker counts appropriate fitted effective degrees of freedom.

## Sampling sites and placement

Compare both site count and placement under the same generating scenarios.
More sites can improve recovery, but redundant sites or weak covariate coverage
may add little information. Hold generating and fitting `nu` explicit in each
comparison, include weak-signal scenarios, and assess whether the fitted model
is numerically identifiable. Study-design conclusions remain internal to the
simulated model. Additional validation against misspecified, coalescent, or
individual-based processes is needed before translating them into field targets.

---

## Comparing sampling strategies and candidate models

`covariance_response_power()` automates the same loop across **sampling
designs**: it subsets the focal sites in the surface to a target sample size,
optionally using different placement strategies, simulates responses, refits
candidate models, and summarizes recovery. The graph is reused across designs, but repeated fitting can still be costly.

Here we compare three sample sizes under two placement strategies, and we fit
two candidate models per replicate: the true two-covariate model and a reduced
altitude-only model. This lets us measure both **parameter recovery** and
**model-selection** performance.

```{r power-run, eval = FALSE}
power <- covariance_response_power(
  theta             = theta_true,
  formula           = ~ forestcover + altitude,
  data              = surface,
  sample_sizes      = c(10, 20, 34),
  strategies        = c("spacefill", "random"),
  conductance_model = loglinear_conductance,
  fit_models = list(
    full          = list(formula = ~ forestcover + altitude),
    altitude_only = list(formula = ~ altitude)
  ),
  tau        = tau_true,
  sigma      = sigma_true,
  nu         = 100,
  nu_fit     = 100,   # change separately to assess overstated precision
  nsim       = 20,    # increase for smoother estimates in a real study
  n_designs  = 2,     # random designs evaluated per sample size
  seed       = 1,
  control    = NewtonRaphsonControl(maxit = 100, verbose = FALSE)
)

power
```


### Scenario-level summary

The `summary` table has one row per (strategy, sample size, model):

```{r power-summary, eval = FALSE}
power$summary[, c("strategy", "sample_size", "model",
                  "fit_rate", "conductance_power",
                  "mean_conductance_cor", "selected_AIC_rate")]
```

Key columns:

- **`fit_rate`**: fraction of replicates that converged. Values below 1 flag a
  design where the model is hard to fit (often too few sites).
- **`conductance_power`**: fraction of replicates whose fitted conductance
  surface has a Pearson correlation with the generating surface above
  `conductance_cor_threshold` (default 0.9). This measures recovery of the
  relative spatial pattern, not its absolute scale or an ecological mechanism.
  It is useful for spline models whose basis coefficients are not individually
  interpretable.
- **`mean_conductance_cor`**: average correlation between the fitted and true
  conductance surfaces.
- **`selected_AIC_rate`**: how often each model had the lowest AIC within the
  specified, comparable candidate set. For the `full` model, this is the rate
  at which the generating formula ranked first among those candidates.

### Parameter-level summary

`parameter_summary` reports per-coefficient power, bias, RMSE, and confidence
interval coverage:

```{r power-params, eval = FALSE}
power$parameter_summary[
  power$parameter_summary$model == "full",
  c("strategy", "sample_size", "parameter",
    "power", "bias", "rmse", "coverage")
]
```

What to look for:

- **`power`**: proportion of all simulated replicates in which the coefficient
  had the generating sign and a two-sided Wald p-value below `alpha`. Failed or
  boundary fits count as non-detections. Choose a target such as 0.8 before the
  assessment rather than treating it as a universal threshold.
- **`bias`**: average estimate minus the generating value. Persistent bias can
  reflect model misspecification, boundary behavior, optimization failures, or a
  design too small for the asymptotic approximation.
- **`rmse`**: typical estimation error; should shrink as sample size grows.
- **`coverage`**: fraction of 95% confidence intervals that contain the truth.
  Well-calibrated intervals give ~0.95. Coverage well below 0.95 means the
  reported uncertainty is too optimistic for that design.


---

## Interpreting and acting on the results

A design assessment usually answers one of three questions:

| Question | Column(s) to read | Decision rule |
|----------|-------------------|---------------|
| Enough sites? | `parameter_summary$power`, `conductance_power` | Pick the smallest sample size that reaches your predeclared recovery target under conservative scenarios. |
| Enough marker information? | power-vs-`nu` curves | Identify the smallest plausible effective `nu` past the power elbow under conservative scenarios. |
| Which site placement? | compare `strategy` rows | Prefer the strategy with higher power / surface correlation at equal cost. |
| Trustworthy intervals? | `parameter_summary$coverage` | Require coverage near 0.95 before trusting reported CIs. |

A few practical notes:

- **Increase `nsim`.** The recipes use small pilot counts; no power sweep is executed at build time.
  For a real assessment use at least 100-500 simulations per design so that
  power and coverage estimates are stable.
- **Match the simulation to your expectations.** The answer depends on
  `theta`, `tau`, and `sigma`. If you are unsure, run the assessment across a
  grid of plausible effect sizes and $\tau / \sigma$ ratios, then report the
  range.
- **Use and report defensible `nu` values.** Markers in linkage disequilibrium
  are not independent, so the effective `nu` can be smaller than the raw marker
  count. Planning with an optimistic `nu` overstates precision and power. Use
  a justified range and report sensitivity to both generating `nu` and
  assumed `nu_fit`.
- **Spline and Gaussian models.** For non-linear conductance models the
  individual coefficients are not interpretable; lean on
  `conductance_power` and `mean_conductance_cor` rather than per-parameter
  power. Pass the relevant factory via `conductance_model` and `fit_models`.

---

## Quick reference

```{r quick-reference, eval = FALSE}
library(terradish)
library(terra)

# 1. Build a landscape template (saveStack = TRUE enables site subsetting)
surface <- conductance_surface(covariates, coords,
                               directions = 8, saveStack = TRUE)

# 2. One simulated covariance matrix from specified generating parameters
sim <- simulate_covariance_response(
  theta = c(x1 = 0.6, x2 = -0.4), formula = ~ x1 + x2,
  data = surface, tau = 2, sigma = 1, nu = 100, seed = 1
)

# 3. Refit to check recovery
fit <- terradish(sim$covariance ~ x1 + x2, data = surface,
                 conductance_model = loglinear_conductance,
                 measurement_model = wishart_covariance, nu = 100)
coef(fit)

# 4. Full design assessment across sample sizes / strategies / models
power <- covariance_response_power(
  theta = c(x1 = 0.6, x2 = -0.4), formula = ~ x1 + x2,
  data = surface, sample_sizes = c(10, 20, 30),
  strategies = c("spacefill", "random"),
  tau = 2, sigma = 1, nu = 100, nu_fit = 100, nsim = 200, seed = 1
)
power$summary            # fit rate, conductance recovery, model selection
power$parameter_summary  # power, bias, RMSE, coverage per coefficient
```


## Summary of key functions

| Function | Purpose |
|----------|---------|
| `conductance_surface()` | Build the landscape template; `saveStack = TRUE` enables site subsetting |
| `simulate_covariance_response()` | One covariance matrix simulated from known parameters |
| `terradish()` | Refit a simulated data set to check parameter recovery |
| `covariance_response_power()` | Full design sweep across sample sizes, strategies, and models |
| `$summary` | Per-scenario fit rate, conductance recovery, and model-selection rate |
| `$parameter_summary` | Per-coefficient power, bias, RMSE, and interval coverage |

## See also

- `vignette("wishart-covariance", package = "terradish")`: the covariance
  workflow and a full explanation of `nu`.
- `vignette("getting-started", package = "terradish")`: the core modeling
  workflow.
- `vignette("model-comparison", package = "terradish")`: AIC, likelihood
  ratio tests, and cross-validation for real data.
