---
title: "Directional gene flow: non-reversible conductance"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Directional gene flow: non-reversible conductance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 4.5, warning = FALSE, message = FALSE)
suppressPackageStartupMessages(library(terradish))
res <- tryCatch(readRDS("vignette-directional.rds"), error = function(e) NULL)
have_res <- !is.null(res)
if (have_res) {
  DIM <- res$example$DIM
  r <- terra::rast(nrows = DIM, ncols = DIM, xmin = 0, xmax = DIM,
                   ymin = 0, ymax = DIM)
  covs <- c(
    terra::setValues(r, res$example$v1_values),
    terra::setValues(r, res$example$elev_values)
  )
  names(covs) <- c("v1", "elev")
  coords <- terra::xyFromCell(r, res$example$focal_cells)
  surface <- conductance_surface(covs, coords, directions = 8)
  dir_cov <- edge_gradient(covs[["elev"]], surface)
  dir_cov$d <- cbind(elev = dir_cov$d)
  edge_rates <- directed_rates(res$fit, data = surface, directional = dir_cov)
}
```

## Why directional conductance?

Standard isolation-by-resistance treats movement as **symmetric**: the cost of
moving from cell *a* to cell *b* equals the cost from *b* to *a*. But gene flow is
frequently **directional**: downstream in a riverscape, downhill, downwind, or
along a source-sink gradient. Accounting for asymmetric movement is one of the
most-requested capabilities in landscape genetics, and symmetric resistance
distance cannot represent it (Lundgren & Ralph 2019; Thomaz et al. 2019).

`terradish_directed()` models movement as a **directed, non-reversible** Markov
generator whose edge rates depend on directional covariates:

\deqn{\log G_{a\to b} = \tfrac12(\eta_a + \eta_b) + d_{ab}^\top \gamma, \qquad \eta_k = x_k^\top \theta,}

where \eqn{\theta} are the usual (symmetric) conductance effects and \eqn{\gamma}
are directional effects on **antisymmetric** edge covariates \eqn{d_{ab} = -d_{ba}}
(elevation drop, flow accumulation, wind). The genetic data are linked through the
**commute time** of the chain, which is symmetric, so every `terradish`
measurement model is used unchanged. With \eqn{\gamma = 0} the generator is
reversible and the model reduces to isolation by resistance.

> **Scope.** This is the tractable directional model (a directed commute time
> reducing to resistance distance when reversible), not the full structured
> coalescent (intractable on fine rasters). It is also a deme/coarse-graph tool
> in its current pure-R form (see "Performance and scope" below).

This vignette uses a small precomputed example so the document can show real
tables and figures without fitting the directed model during package checks. The
vignette rebuilds the tiny example graph, but the fitted model itself is read
from `vignettes/vignette-directional.rds`. To refresh the cached fit, run
`dev/make_vignette_directional.R` from the source tree.

## Directional edge covariates

`edge_gradient()` turns a spatial layer (e.g. elevation) into the antisymmetric
per-directed-edge covariate \eqn{d_{ab} = x_a - x_b}:

```{r edges, eval = FALSE}
library(terradish)
library(terra)

# A coarse/deme-scale conductance surface (directional fitting is for small or
# coarse graphs in the current R engine; see Performance and scope).
surface <- conductance_surface(covariates, coords, directions = 8)

# Directional covariate from an elevation layer (downhill = positive drop)
dir_cov <- edge_gradient(elevation, surface)

# Optional, but helpful: name the directional covariate so the coefficient
# appears as gamma_elev rather than gamma1 in printed output.
dir_cov$d <- cbind(elev = dir_cov$d)
```

## Fitting a directional model

```{r fit, eval = FALSE}
fit <- terradish_directed(
  S ~ habitat,                    # symmetric conductance covariate(s)
  data              = surface,
  directional       = dir_cov,    # directional covariate(s)
  measurement_model = wishart_covariance,   # or generalized_wishart / mlpe
  nu                = nu
)
summary(fit)        # theta (symmetric) + gamma (directional), z-tests, AIC
```

The fitted object carries the full S3 suite: `print`, `summary`, `coef`,
`logLik`, `AIC`, `vcov`, `confint`, and `plot`. The cached example below was
simulated from a known directional model on a 7 by 7 lattice, with six focal
sites, 156 undirected graph edges, \eqn{\theta = 0.5}, and
\eqn{\gamma = 0.6}.

```{r cached-summary, eval = have_res}
summary(res$fit)
```

**How to read the `summary()` output:** The `theta` row is the reversible
conductance effect. It is on the log-conductance scale, so a one-unit increase
in `v1` multiplies the symmetric conductance by `exp(theta)`. The `gamma` row is
the directional effect. It is also on a log-rate scale, but it applies to an
antisymmetric edge covariate: positive `gamma_elev` means edges with positive
`elev` gradient have higher fitted rate in the positive-gradient direction and
lower fitted rate in the reverse direction. The z table is a Wald approximation;
use it as evidence for detectability, then use the edge-rate plots below to see
where that detectable directional mechanism is expressed on the graph.

A likelihood-ratio test of the directional effect compares the model against its
reversible (${\gamma = 0}$) special case:

```{r lrt, eval = FALSE}
# Refit the same model with the directional coefficient fixed at zero.
fit_reversible <- terradish_directed(
  S ~ habitat,
  data              = surface,
  directional       = dir_cov,
  measurement_model = wishart_covariance,
  nu                = nu,
  gamma_bound       = 0
)

lrt_stat <- 2 * (as.numeric(logLik(fit)) -
                   as.numeric(logLik(fit_reversible)))

# The test df is the number of constrained directional coefficients.
# Set this explicitly because the reduced object still carries a gamma slot
# fixed at zero.
lrt_df <- length(fit$gamma)
lrt_p  <- pchisq(lrt_stat, df = lrt_df, lower.tail = FALSE)

data.frame(statistic = lrt_stat, df = lrt_df, p_value = lrt_p)
```

```{r lrt-table, echo = FALSE, eval = have_res}
knitr::kable(
  res$lrt_table,
  digits = 3,
  caption = "Likelihood-ratio comparison of the directed model against the reversible special case."
)
```

**How to read the likelihood-ratio test:** The reduced model keeps the same
symmetric conductance structure but fixes \eqn{\gamma = 0}. The test statistic
is twice the log-likelihood gain from allowing directionality. The degrees of
freedom are the number of directional coefficients fixed at zero in the reduced
model. A small p-value means the directed generator fits better than its
reversible special case for this response.

## Visualizing directional results

Use the coefficient table and the map together. The coefficient table answers
whether the directional mechanism is supported; the map shows where that fitted
mechanism has the strongest edge-level expression on the graph.

```{r recovery-table, echo = FALSE, eval = have_res}
knitr::kable(
  res$coef_table,
  digits = 3,
  caption = "Directed-model recovery on simulated data."
)
```

**How to read the coefficient table:** The `truth` column is the parameter value
used to generate the synthetic response. The `estimate` column is what
`terradish_directed()` recovered from that response. Here the fitted symmetric
conductance effect and the directional effect are both close to the generating
values, which is what this controlled example is meant to demonstrate.

```{r directional-visuals, eval = FALSE}
# One row per undirected graph edge, with fitted rates in both directions.
edge_rates <- directed_rates(fit, data = surface, directional = dir_cov)

# Arrows point from the lower-rate endpoint toward the higher-rate endpoint.
# Width and color increase with |log(rate_ab / rate_ba)|.
plot(fit, type = "directional", data = surface, directional = dir_cov)

# Overlay directional arrows on the fitted reversible conductance surface.
plot(fit, type = "combined", data = surface, directional = dir_cov)
```

```{r edge-summary, echo = FALSE, eval = have_res}
knitr::kable(
  res$edge_summary,
  digits = 3,
  caption = "Summary of fitted directional edge-rate bias in the cached example."
)
```

The key edge-scale quantity is
${\log(G_{a\to b} / G_{b\to a})}$. A value of 0 means the fitted rate is the
same in both directions along that edge. A value of 0.69 means the favored
direction has about twice the rate of the reverse direction, because
${\exp(0.69) \approx 2}$.

```{r directional-plot, echo = FALSE, fig.width = 6, fig.height = 5, eval = have_res}
plot(res$fit, type = "directional", data = surface, directional = dir_cov,
     min_abs_log_ratio = 0.05)
```

**How to read the directional map:** Each arrow points from the lower-rate
endpoint toward the higher-rate endpoint on an undirected graph edge. Color and
line width increase with ${|\log(G_{a\to b} / G_{b\to a})|}$. The map is an
edge-level diagnostic for the fitted generator, not a direct observation of
migration.

```{r combined-plot, echo = FALSE, fig.width = 6, fig.height = 5, eval = have_res}
plot(res$fit, type = "combined", data = surface, directional = dir_cov,
     min_abs_log_ratio = 0.05)
```

**How to read the combined map:** The background raster is the reversible
conductance component, `exp(x theta)`. The arrows show the directional bias on
top of that surface. Read the two layers separately first: the background says
where movement is generally easy or difficult, and the arrows say which
direction is favored along each edge after accounting for the symmetric
component.

The directional map is inspired by effective-migration surface displays such as
EEMS and FEEMS, but the interpretation is model-specific. In
`terradish_directed()`, arrows show fitted **directional edge-rate bias** in the
directed generator. They should not be described as directly observed migration
rates. For presentation figures, consider filtering very small arrows:

```{r directional-filter, eval = FALSE}
plot(fit, type = "combined", data = surface, directional = dir_cov,
     min_abs_log_ratio = 0.6)
```

### Worked example (simulated from the directed model)

To create the cached result, we simulated a covariance response from the
directed model on a small lattice with a positive directional effect and refit
the model:

```{r recovery-show, eval = FALSE}
fit <- terradish_directed(S ~ v1, data = surface, directional = dir_cov,
                          measurement_model = generalized_wishart, nu = nu)
coef(fit)           # theta (~truth) and gamma (~truth, positive)
```

```{r grad-check, echo = FALSE, results = "asis", eval = have_res}
cat(sprintf("The reverse-mode (transpose-solve) gradient matches numerical differentiation to **%.0e**, and gamma is recovered with the correct (positive) sign.\n", res$grad_err))
```

## Identifiability: can direction be recovered from symmetric data?

This is the central question, and the honest answer has two parts.

**In principle: yes.** The genetic response (a distance/covariance) is symmetric,
so only a symmetric function of the directed generator (the commute time) is
observable. One might fear this erases direction. It does for perfectly
symmetric/periodic geometry, but **landscape boundaries break that symmetry**, so
direction *does* leave a signature in the symmetric commute time. On a bounded
lattice the commute-time pattern at \eqn{+\gamma} and \eqn{-\gamma} are
distinguishable, and \eqn{\gamma} (sign and magnitude) is recovered from
model-simulated data.

**In practice: it is hard, and requires strong spatial structure.** Real genetic
data are generated by a coalescent process, not by the commute-time model, and
the directional component is a perturbation on top of (much larger) symmetric
isolation by distance. Detecting \eqn{\gamma} therefore needs:

- **strong differentiation** (low effective migration / high \eqn{F_{ST}}); when
  populations are well mixed the spatial signal, and with it any directional
  signal, is weak, and the resistance scale itself collapses
  (\eqn{\tau \to 0});
- **many markers and sites**, since the directional signature is second-order.

Treat a non-significant \eqn{\gamma} as "not detectable here," not "no
directionality," and report \eqn{\gamma} with its uncertainty and a sensitivity
check. Where direction is the primary question and differentiation is modest,
data summaries that carry directional information directly (allele-frequency
clines, identity-by-descent asymmetries) are more powerful than any symmetric
distance. This is a fundamental limit, not a limitation of this implementation.

## Performance and scope

The current engine factorizes one sub-generator per focal site per likelihood
evaluation. This is exact and fine for **small or coarse graphs** (deme-scale
analyses, coarsened rasters), but a *fit* on a full-resolution raster (hundreds of
thousands of cells) is slow, and strongly heterogeneous rates can ill-condition
the solves. A non-symmetric sparse-LU C++ backend (one factorization reused for
the forward and transpose solves) is planned for full-resolution use. Until then:

- coarsen the raster (`terra::aggregate`) or work at the deme scale;
- keep covariates scaled and directional effects moderate;
- check that estimates are stable across resolutions.

## Quick-reference

```{r quickref, eval = FALSE}
library(terradish)
surface <- conductance_surface(covariates, coords, directions = 8)  # coarse/deme-scale
dir_cov <- edge_gradient(elevation, surface)                        # directional covariate
fit <- terradish_directed(S ~ habitat, data = surface, directional = dir_cov,
                          measurement_model = wishart_covariance, nu = nu)
summary(fit)                       # theta (symmetric) + gamma (directional)
plot(fit, data = surface)          # symmetric conductance surface
edge_rates <- directed_rates(fit, data = surface, directional = dir_cov)
head(edge_rates)                   # fitted edge rates and log-rate ratios
plot(fit, type = "combined", data = surface, directional = dir_cov)
```
