---
title: "Non-reversible conductance and directional edge-rate bias"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Non-reversible conductance and directional edge-rate bias}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 4.5, warning = FALSE, message = FALSE)
suppressPackageStartupMessages(library(terradish))
res <- tryCatch(readRDS("vignette-directional.rds"), error = function(e) NULL)
have_res <- !is.null(res)
if (have_res) {
  DIM <- res$example$DIM
  r <- terra::rast(nrows = DIM, ncols = DIM, xmin = 0, xmax = DIM,
                   ymin = 0, ymax = DIM)
  covs <- c(
    terra::setValues(r, res$example$v1_values),
    terra::setValues(r, res$example$elev_values)
  )
  names(covs) <- c("v1", "elev")
  coords <- terra::xyFromCell(r, res$example$focal_cells)
  surface <- conductance_surface(covs, coords, directions = 8)
  dir_cov <- edge_gradient(covs[["elev"]], surface)
  dir_cov$d <- cbind(elev = dir_cov$d)
  edge_rates <- directed_rates(res$fit, data = surface, directional = dir_cov)
}
```

## Why directional conductance?

Standard isolation-by-resistance uses symmetric edge weights. That structure
cannot represent a fitted rate from cell *a* to cell *b* that differs from the
reverse rate. Such asymmetry may be relevant to hypotheses about rivers,
elevation gradients, wind, or source-sink structure.

`terradish_directed()` fits a **directed, non-reversible** Markov generator whose
edge rates depend on directional covariates:

$$
\log G_{a\to b} = \tfrac12(\eta_a + \eta_b) + d_{ab}^\top \gamma,
\qquad \eta_k = x_k^\top \theta.
$$

where $\theta$ are symmetric conductance coefficients and $\gamma$ are
edge-rate coefficients on **antisymmetric** edge covariates
$d_{ab} = -d_{ba}$
(elevation drop, flow accumulation, wind). The genetic data are linked through the
**commute time** of the chain, which is symmetric, so every `terradish`
measurement model is used unchanged. With $\gamma = 0$ the generator is
reversible and the model reduces to isolation by resistance.

> **Scope.** This is the tractable directional model (a directed commute time
> reducing to resistance distance when reversible), not the full structured
> coalescent or a direct migration-rate model. It is most useful at deme scale or
> on coarse rasters; full-resolution use should start with the SparseLU backend
> and careful sensitivity checks (see "Performance and scope" below).

This vignette uses a small precomputed example so the document can show real
tables and figures without fitting the directed model during package checks. The
vignette rebuilds the tiny example graph, but the fitted model itself is read
from `vignettes/vignette-directional.rds`. To refresh the cached fit, run
`dev/make_vignette_directional.R` from the source tree.

## Directional edge covariates

`edge_gradient()` turns a spatial layer (e.g. elevation) into the antisymmetric
per-directed-edge covariate $d_{ab} = x_a - x_b$:

```{r edges, eval = FALSE}
library(terradish)
library(terra)

# A coarse/deme-scale conductance surface is the recommended starting point.
surface <- conductance_surface(covariates, coords, directions = 8)

# Directional covariate from an elevation layer (downhill = positive drop)
dir_cov <- edge_gradient(elevation, surface)

# Optional, but helpful: name the directional covariate so the coefficient
# appears as gamma_elev rather than gamma1 in printed output.
dir_cov$d <- cbind(elev = dir_cov$d)
```

## Fitting a directional model

```{r fit, eval = FALSE}
fit <- terradish_directed(
  S ~ habitat,                    # symmetric conductance covariate(s)
  data              = surface,
  directional       = dir_cov,    # directional covariate(s)
  measurement_model = wishart_covariance,   # match this to the response type
  nu                = nu,
  solver            = "sparse_lu_cpp"
)
summary(fit)        # theta (symmetric) + gamma (directional), z-tests, AIC
```

The response requirements do not change in the directed extension.
`wishart_covariance` requires a positive semidefinite covariance matrix,
`generalized_wishart` requires an admissible squared-distance response
coherently related to a centered positive semidefinite covariance matrix, and
`mlpe` or `leastsquares` accepts a pairwise distance response. Use
`check_distance_response()` before a generalized-Wishart fit;
`terradish_directed()` performs the same check automatically.

The fitted object carries the full S3 suite: `print`, `summary`, `coef`,
`logLik`, `AIC`, `vcov`, `confint`, and `plot`. The cached example below was
simulated from a known directional model on a 7 by 7 lattice, with six focal
sites, 156 undirected graph edges, $\theta = 0.5$, and $\gamma = 0.6$.

```{r cached-summary, eval = have_res}
summary(res$fit)
```

**How to read the `summary()` output:** The `theta` row is the reversible
conductance coefficient. It is on the log-conductance scale, so a one-unit increase
in `v1` multiplies the symmetric conductance by `exp(theta)`. The `gamma` row is
the directional edge-rate coefficient. It is also on a log-rate scale, but it applies to an
antisymmetric edge covariate: positive `gamma_elev` means edges with positive
`elev` gradient have higher fitted rate in the positive-gradient direction and
lower fitted rate in the reverse direction. The z table is a Wald approximation;
use it as conditional evidence for that specified covariate, then use the
edge-rate plots below to see where the fitted component is expressed on the
graph. Neither the coefficient nor the arrows directly observe migration.

A likelihood-ratio test of the directional term compares the model against its
reversible (${\gamma = 0}$) special case:

```{r lrt, eval = FALSE}
# Refit the same model with the directional coefficient fixed at zero.
fit_reversible <- terradish_directed(
  S ~ habitat,
  data              = surface,
  directional       = dir_cov,
  measurement_model = wishart_covariance,
  nu                = nu,
  gamma_bound       = 0,
  solver            = "sparse_lu_cpp"
)

lrt_stat <- 2 * (as.numeric(logLik(fit)) -
                   as.numeric(logLik(fit_reversible)))

# The test df is the number of constrained directional coefficients.
# Set this explicitly because the reduced object still carries a gamma slot
# fixed at zero.
lrt_df <- length(fit$gamma)
lrt_p  <- pchisq(lrt_stat, df = lrt_df, lower.tail = FALSE)

data.frame(statistic = lrt_stat, df = lrt_df, p_value = lrt_p)
```

```{r lrt-table, echo = FALSE, eval = have_res}
knitr::kable(
  res$lrt_table,
  digits = 3,
  caption = "Likelihood-ratio comparison of the directed model against the reversible special case."
)
```

**How to read the likelihood-ratio test:** The reduced model keeps the same
symmetric conductance structure but fixes $\gamma = 0$. The test statistic
is twice the log-likelihood gain from allowing directionality. The degrees of
freedom are the number of directional coefficients fixed at zero in the reduced
model. A small p-value indicates stronger conditional evidence for the specified
directed component than for its reversible special case. For Wishart fits, the
strength of this evidence depends on the supplied `nu`.

## Visualizing directional results

Use the coefficient table and the map together. The coefficient table summarizes
conditional support for the specified edge-rate term; the map shows where that
fitted term has the strongest edge-level expression on the graph.

```{r recovery-table, echo = FALSE, eval = have_res}
knitr::kable(
  res$coef_table,
  digits = 3,
  caption = "Directed-model recovery on simulated data."
)
```

**How to read the coefficient table:** The `truth` column is the parameter value
used to generate the synthetic response. The `estimate` column is what
`terradish_directed()` recovered from that response. Here the fitted symmetric
conductance coefficient and directional coefficient are both close to the generating
values, which is what this controlled example is meant to demonstrate.

```{r directional-visuals, eval = FALSE}
# One row per undirected graph edge, with fitted rates in both directions.
edge_rates <- directed_rates(fit, data = surface, directional = dir_cov)

# Arrows point from the lower-rate endpoint toward the higher-rate endpoint.
# Width and color increase with |log(rate_ab / rate_ba)|.
plot(fit, type = "directional", data = surface, directional = dir_cov)

# Overlay directional arrows on the fitted reversible conductance surface.
plot(fit, type = "combined", data = surface, directional = dir_cov)
```

```{r edge-summary, echo = FALSE, eval = have_res}
knitr::kable(
  res$edge_summary,
  digits = 3,
  caption = "Summary of fitted directional edge-rate bias in the cached example."
)
```

The key edge-scale quantity is
${\log(G_{a\to b} / G_{b\to a})}$. A value of 0 means the fitted rate is the
same in both directions along that edge. A value of 0.69 means the favored
direction has about twice the rate of the reverse direction, because
${\exp(0.69) \approx 2}$.

```{r directional-plot, echo = FALSE, fig.width = 6, fig.height = 5, eval = have_res}
plot(res$fit, type = "directional", data = surface, directional = dir_cov,
     min_abs_log_ratio = 0.05)
```

**How to read the directional map:** Each arrow points from the lower-rate
endpoint toward the higher-rate endpoint on an undirected graph edge. Color and
line width increase with ${|\log(G_{a\to b} / G_{b\to a})|}$. The map is an
edge-level diagnostic for the fitted generator, not a direct observation of
migration.

```{r combined-plot, echo = FALSE, fig.width = 6, fig.height = 5, eval = have_res}
plot(res$fit, type = "combined", data = surface, directional = dir_cov,
     min_abs_log_ratio = 0.05)
```

**How to read the combined map:** The background raster is the reversible
conductance component, `exp(x theta)`. The arrows show the directional bias on
top of that surface. Read the two layers separately first: the background says
where fitted relative conductance is high or low, and the arrows show which
direction the fitted generator favors along each edge after accounting for the
symmetric component. Neither layer is a direct movement map.

The directional map is inspired by effective-migration surface displays such as
EEMS and FEEMS, but the interpretation is model-specific. In
`terradish_directed()`, arrows show fitted **directional edge-rate bias** in the
directed generator. They should not be described as directly observed migration
rates. For presentation figures, consider filtering very small arrows:

```{r directional-filter, eval = FALSE}
plot(fit, type = "combined", data = surface, directional = dir_cov,
     min_abs_log_ratio = 0.6)
```

### Worked example (simulated from the directed model)

To create the cached result, we simulated a covariance response from the
directed model on a small lattice with a positive directional coefficient and
refit the model with the covariance Wishart likelihood:

```{r recovery-show, eval = FALSE}
fit <- terradish_directed(S ~ v1, data = surface, directional = dir_cov,
                          measurement_model = wishart_covariance, nu = nu,
                          solver = "sparse_lu_cpp")
coef(fit)           # theta (~truth) and gamma (~truth, positive)
```

```{r grad-check, echo = FALSE, results = "asis", eval = have_res}
cat(sprintf("The reverse-mode (transpose-solve) gradient matches numerical differentiation to **%.0e**, and gamma is recovered with the correct (positive) sign.\n", res$grad_err))
```

## Identifiability: can direction be recovered from symmetric data?

This is the central question, and the honest answer has two parts.

**In principle: yes.** The genetic response (a distance/covariance) is symmetric,
so only a symmetric function of the directed generator (the commute time) is
observable. One might fear this erases direction. It does for perfectly
symmetric/periodic geometry, but **landscape boundaries break that symmetry**, so
direction *does* leave a signature in the symmetric commute time. On a bounded
lattice the commute-time patterns at $+\gamma$ and $-\gamma$ can differ. The
controlled example above recovers the generating sign and magnitude, but this
does not establish unique identification for other graphs or responses.

**In practice: it is hard, and requires strong spatial structure.** Real genetic
data are generated by a coalescent process, not by the commute-time model, and
the directional component is a perturbation on top of (much larger) symmetric
isolation by distance. Detecting $\gamma$ is therefore more plausible with:

- a response with a detectable spatial signal; when the fitted resistance scale
  approaches $\tau = 0$, neither symmetric nor directional conductance is
  identified;
- enough sites and effective marker information to resolve the weak commute-time
  signature, while recognizing that more data do not guarantee uniqueness.

Treat an interval for $\gamma$ that includes zero as "not distinguishable here,"
not "no biological directionality," and report $\gamma$ with its uncertainty
and a sensitivity check. Where direction is the primary question and
differentiation is modest, data summaries that carry directional information
directly (allele-frequency clines, identity-by-descent asymmetries) are more
powerful than any symmetric distance. This is a fundamental limit, not a
limitation of this implementation.

## Performance and scope

The directed engine factorizes one sub-generator per focal site per likelihood
evaluation. This is exact and fine for **small or coarse graphs** (deme-scale
analyses, coarsened rasters). For larger graphs, use
`solver = "sparse_lu_cpp"` to call the non-symmetric Eigen SparseLU backend. It
caches one factorization per focal absorber and reuses it for the forward
hitting-time solve and the transpose adjoint solve. Even with that backend,
full-resolution raster fits (hundreds of thousands of cells) can be slow, and
strongly heterogeneous rates can ill-condition the solves. Practical guidance:

- coarsen the raster (`terra::aggregate`) or work at the deme scale;
- keep covariates scaled and directional coefficients moderate;
- check that estimates are stable across resolutions.

## Quick-reference

```{r quickref, eval = FALSE}
library(terradish)
surface <- conductance_surface(covariates, coords, directions = 8)  # coarse/deme-scale
dir_cov <- edge_gradient(elevation, surface)                        # directional covariate
fit <- terradish_directed(S ~ habitat, data = surface, directional = dir_cov,
                          measurement_model = wishart_covariance, nu = nu,
                          solver = "sparse_lu_cpp")
summary(fit)                       # theta (symmetric) + gamma (directional)
plot(fit, data = surface)          # symmetric conductance surface
edge_rates <- directed_rates(fit, data = surface, directional = dir_cov)
head(edge_rates)                   # fitted edge rates and log-rate ratios
plot(fit, type = "combined", data = surface, directional = dir_cov)
```

## Summary of key functions

| Function | Purpose |
|----------|---------|
| `conductance_surface()` | Build the weighted graph; use a coarse/deme-scale grid here |
| `edge_gradient()` | Turn a raster into a signed directional edge covariate |
| `terradish_directed()` | Fit the non-reversible model, estimating theta and gamma |
| `summary()` | Coefficient table; `gamma = 0` is the reversible case for the specified edge covariates |
| `anova()` | Likelihood ratio test of the directional model against `gamma = 0` |
| `directed_rates()` | Per-edge forward and backward rates and their log ratio |
| `plot(type = "combined")` | Directional arrows drawn over the conductance surface |

## See also

- **`vignette("getting-started", package = "terradish")`**, the core workflow
  this vignette assumes: scaling covariates, building a surface, fitting and
  reading a symmetric model.
- **`vignette("wishart-covariance", package = "terradish")`**, the covariance
  measurement model used here and how to choose `nu`.
- **`vignette("model-comparison", package = "terradish")`**, likelihood-ratio
  tests and AIC under their comparability conditions.
- **`vignette("large-landscapes", package = "terradish")`**, why the directed
  engine costs more per evaluation and how far it scales.
