---
title: "Directional gene flow: non-reversible conductance"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Directional gene flow: non-reversible conductance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 4.5, warning = FALSE, message = FALSE)
res <- tryCatch(readRDS("vignette-directional.rds"), error = function(e) NULL)
have_res <- !is.null(res)
```

## Why directional conductance?

Standard isolation-by-resistance treats movement as **symmetric**: the cost of
moving from cell *a* to cell *b* equals the cost from *b* to *a*. But gene flow is
frequently **directional**: downstream in a riverscape, downhill, downwind, or
along a source-sink gradient. Accounting for asymmetric movement is one of the
most-requested capabilities in landscape genetics, and symmetric resistance
distance cannot represent it (Lundgren & Ralph 2019; Thomaz et al. 2019).

`terradish_directed()` models movement as a **directed, non-reversible** Markov
generator whose edge rates depend on directional covariates:

\deqn{\log G_{a\to b} = \tfrac12(\eta_a + \eta_b) + d_{ab}^\top \gamma, \qquad \eta_k = x_k^\top \theta,}

where \eqn{\theta} are the usual (symmetric) conductance effects and \eqn{\gamma}
are directional effects on **antisymmetric** edge covariates \eqn{d_{ab} = -d_{ba}}
(elevation drop, flow accumulation, wind). The genetic data are linked through the
**commute time** of the chain, which is symmetric, so every `terradish`
measurement model is used unchanged. With \eqn{\gamma = 0} the generator is
reversible and the model reduces to isolation by resistance.

> **Scope.** This is the tractable directional model (a directed commute time
> reducing to resistance distance when reversible), not the full structured
> coalescent (intractable on fine rasters). It is also a deme/coarse-graph tool
> in its current pure-R form (see "Performance and scope" below).

## Directional edge covariates

`edge_gradient()` turns a spatial layer (e.g. elevation) into the antisymmetric
per-directed-edge covariate \eqn{d_{ab} = x_a - x_b}:

```{r edges, eval = FALSE}
library(terradish)
library(terra)

# A coarse/deme-scale conductance surface (directional fitting is for small or
# coarse graphs in the current R engine; see Performance and scope).
surface <- conductance_surface(covariates, coords, directions = 8)

# Directional covariate from an elevation layer (downhill = positive drop)
dir_cov <- edge_gradient(elevation, surface)
```

## Fitting a directional model

```{r fit, eval = FALSE}
fit <- terradish_directed(
  S ~ habitat,                    # symmetric conductance covariate(s)
  data              = surface,
  directional       = dir_cov,    # directional covariate(s)
  measurement_model = wishart_covariance,   # or generalized_wishart / mlpe
  nu                = nu
)
summary(fit)        # theta (symmetric) + gamma (directional), z-tests, AIC
```

The fitted object carries the full S3 suite: `print`, `summary`, `coef`,
`logLik`, `AIC`, `vcov`, `confint`, and `plot` (the symmetric conductance
surface). A likelihood-ratio test of the directional effect compares the model
against its reversible (\eqn{\gamma = 0}) special case:

```{r lrt, eval = FALSE}
# directed vs reversible (gamma = 0) on the same commute-time engine
AIC(fit)            # vs the reversible reduced fit
```

### Worked example (simulated from the directed model)

To show the machinery recovering a known directional effect, we simulate a
covariance response from the directed model on a small lattice with a positive
(downhill-biased) \eqn{\gamma} and refit:

```{r recovery-show, eval = FALSE}
fit <- terradish_directed(S ~ v1, data = surface, directional = dir_cov,
                          measurement_model = generalized_wishart, nu = nu)
coef(fit)           # theta (~truth) and gamma (~truth, positive)
```

```{r recovery-table, echo = FALSE, eval = have_res}
knitr::kable(res$coef_table, digits = 3,
             caption = "Directed-model recovery on simulated data (true theta = 0.5, gamma = 0.6).")
```

```{r grad-check, echo = FALSE, results = "asis", eval = have_res}
cat(sprintf("The reverse-mode (transpose-solve) gradient matches numerical differentiation to **%.0e**, and gamma is recovered with the correct (positive) sign.\n", res$grad_err))
```

## Identifiability: can direction be recovered from symmetric data?

This is the central question, and the honest answer has two parts.

**In principle: yes.** The genetic response (a distance/covariance) is symmetric,
so only a symmetric function of the directed generator (the commute time) is
observable. One might fear this erases direction. It does for perfectly
symmetric/periodic geometry, but **landscape boundaries break that symmetry**, so
direction *does* leave a signature in the symmetric commute time. On a bounded
lattice the commute-time pattern at \eqn{+\gamma} and \eqn{-\gamma} are
distinguishable, and \eqn{\gamma} (sign and magnitude) is recovered from
model-simulated data.

**In practice: it is hard, and requires strong spatial structure.** Real genetic
data are generated by a coalescent process, not by the commute-time model, and
the directional component is a perturbation on top of (much larger) symmetric
isolation by distance. Detecting \eqn{\gamma} therefore needs:

- **strong differentiation** (low effective migration / high \eqn{F_{ST}}); when
  populations are well mixed the spatial signal, and with it any directional
  signal, is weak, and the resistance scale itself collapses
  (\eqn{\tau \to 0});
- **many markers and sites**, since the directional signature is second-order.

Treat a non-significant \eqn{\gamma} as "not detectable here," not "no
directionality," and report \eqn{\gamma} with its uncertainty and a sensitivity
check. Where direction is the primary question and differentiation is modest,
data summaries that carry directional information directly (allele-frequency
clines, identity-by-descent asymmetries) are more powerful than any symmetric
distance. This is a fundamental limit, not a limitation of this implementation.

## Performance and scope

The current engine factorizes one sub-generator per focal site per likelihood
evaluation. This is exact and fine for **small or coarse graphs** (deme-scale
analyses, coarsened rasters), but a *fit* on a full-resolution raster (hundreds of
thousands of cells) is slow, and strongly heterogeneous rates can ill-condition
the solves. A non-symmetric sparse-LU C++ backend (one factorization reused for
the forward and transpose solves) is planned for full-resolution use. Until then:

- coarsen the raster (`terra::aggregate`) or work at the deme scale;
- keep covariates scaled and directional effects moderate;
- check that estimates are stable across resolutions.

## Quick-reference

```{r quickref, eval = FALSE}
library(terradish)
surface <- conductance_surface(covariates, coords, directions = 8)  # coarse/deme-scale
dir_cov <- edge_gradient(elevation, surface)                        # directional covariate
fit <- terradish_directed(S ~ habitat, data = surface, directional = dir_cov,
                          measurement_model = wishart_covariance, nu = nu)
summary(fit)                       # theta (symmetric) + gamma (directional)
plot(fit, data = surface)          # symmetric conductance surface
```
