---
title: "Joint Gaussian Scale Optimization in terradish"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Joint Gaussian Scale Optimization in terradish}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 6,
  fig.height = 4.5,
  warning   = FALSE,
  message   = FALSE
)
```

## Why optimize sigma inside the model?

The original `radish` workflow for scale of effect usually worked in two steps:

1. Smooth each raster at several candidate scales.
2. Fit a resistance model to each smoothed raster surface and compare the fits.

That approach is still useful, but it treats scale as an **outer** tuning
parameter. The new native Gaussian conductance system in `terradish` moves the
scale parameter into the conductance model itself, so the optimizer can update:

- the conductance coefficient(s), and
- the smoothing scale(s)

in the same likelihood calculation.

This vignette shows the current implementation of that idea using
`gaussian_smoothed_loglinear_conductance()`.

---

## What sigma means here

For a raster covariate `x`, the Gaussian scale-aware model first smooths the
raster with a Gaussian kernel and then uses the smoothed values in the usual
log-linear conductance model.

Conceptually:

$$C_i = \exp(\beta \cdot x_{i,\sigma})$$

where:

- $x_{i,\sigma}$ is the Gaussian-smoothed raster value at cell *i*
- $\beta$ is the conductance coefficient
- $\sigma$ controls how broad the smoothing kernel is

Small `sigma` means a very local scale of effect. Large `sigma` means the
covariate is averaged over a broader neighborhood.

### Are there limits on sigma?

Yes. `terradish` now estimates `sigma` directly on its natural distance scale
and fits it within explicit lower and upper bounds.

By default, those bounds are data-derived:

- the lower bound is half of the mean raster cell width
- the upper bound is the diagonal length of the retained raster extent

You can override those defaults with `sigma_lower =` and `sigma_upper =` when
you create the conductance model.

Internally, the optimizer now works on a converted sigma scale by default,
using approximate cell-width units. The fitted coefficients and summaries are
still reported back in map units for users, so the conversion helps the
optimization without changing the interpretation target.

Two practical caveats are worth keeping in mind:

- factor-valued raster layers are rejected for this Gaussian conductance model
  rather than being treated as scalable predictors
- leverage / `partial_X` diagnostics are still unavailable, because the current
  diagnostic machinery assumes a one-to-one match between conductance
  parameters and design-matrix covariates

---

## Worked example

We will fit a small staged example using the `melip` data. To keep the vignette
fast and stable, we will:

- use only eight sampling sites
- start with one raster covariate (`forestcover`)
- use `mlpe` as the measurement model, because the response matrix contains
  non-independent pairwise genetic distances

This is enough to show the practical workflow without making the vignette slow.
The main goal here is not to maximize model complexity. Instead, we will use
the staged workflow that is usually safest for real data:

1. fit a one-raster Gaussian scale model
2. compare it to a standard fixed-raster model
3. only then add a second raster and inspect whether the fitted scales still
   behave sensibly

```{r load}
library(terradish)
library(terra)

data(melip)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.altitude <- terra::unwrap(melip.altitude)
melip.coords <- terra::unwrap(melip.coords)

keep <- 1:8
melip.Fst <- melip.Fst[keep, keep, drop = FALSE]
coords <- melip.coords[keep]
```

### Step 1 — Build a conductance surface with the original raster

Unlike the standard fixed-raster workflow, we do **not** pre-scale the raster
with `scale_covariates()` here. The Gaussian conductance model smooths the raw
raster first, then standardizes the active-cell values internally at each
parameter evaluation.

```{r surface}
covariates <- melip.forestcover
names(covariates) <- "forestcover"

surface <- conductance_surface(
  covariates,
  coords,
  directions = 8,
  saveStack = TRUE
)
```

Two details matter here:

- `saveStack = TRUE` is required, because the Gaussian conductance model needs
  the retained raster stack as its smoothing domain.
- We are fitting the scale of effect for the raster itself, so the original
  raster values need to be available.

### Step 2 — Create the Gaussian scale-aware conductance model

```{r gaussian-model}
gaussian_model <- gaussian_smoothed_loglinear_conductance(surface)
```

This conductance model will estimate two parameters:

- `forestcover`: the conductance coefficient
- `sigma.forestcover`: the Gaussian smoothing scale on the natural distance scale

The same mechanism also supports interactions and polynomial terms after
smoothing, but it is usually better to start with the simplest one-raster model
you can justify.

### Step 3 — Fit the one-raster Gaussian scale model

`optimizer = "auto"` currently resolves to BFGS for this conductance model.
That is intentional: the native Gaussian scale path supports analytic
gradients, and BFGS is the stable first optimizer for the joint
`beta + sigma` parameter vector.

```{r fit}
fit_gaussian <- terradish(
  melip.Fst ~ forestcover,
  data = surface,
  conductance_model = gaussian_model,
  measurement_model = mlpe,
  optimizer = "auto",
  leverage = FALSE,
  control = NewtonRaphsonControl(maxit = 12, verbose = FALSE)
)

fit_gaussian
```

### Step 4 — Inspect the fitted one-raster scale

```{r coef}
coef(fit_gaussian)
```

The conductance coefficient tells us whether smoothed forest cover is
associated with higher or lower conductance. The scale parameter is already on
its natural scale, so we can summarize it directly:

```{r sigma}
coef(fit_gaussian)[["sigma.forestcover"]]
```

`gaussian_scale_summary()` converts the fitted sigma values into:

- the native raster distance scale
- the internal optimization scale and its conversion factor
- raster cell widths
- one-dimensional Gaussian half-widths
- two-dimensional radial extents that contain a chosen amount of kernel mass

```{r sigma-summary}
gaussian_scale_summary(fit_gaussian)
```

That table is often the quickest way to interpret the scale of effect. For
example, `radial_95` is the radius that contains approximately 95% of the
Gaussian kernel mass, while `sigma_cells_x` and `sigma_cells_y` show how broad
the effect is relative to the raster resolution.

You can also visualize the fitted kernel directly. The `sigma` plot shows the
Gaussian weight curve over distance for each fitted scale parameter, with a
vertical dashed line at the fitted effective distance containing 90% of the
one-dimensional kernel mass and a shaded uncertainty band for that distance.
Each panel is also annotated with:

- the fitted `sigma`
- the fitted effective 90% distance
- the corresponding confidence interval bounds

```{r sigma-plot, fig.cap = "***Fitted Gaussian kernel with uncertainty band. The dashed line marks the distance containing 90% of the one-dimensional kernel mass.***"}
plot(fit_gaussian, type = "sigma")
```

If you already know how to convert your map units into a more interpretable
distance scale, you can relabel the plot directly. For example, if one map
unit corresponds to one metre, then multiplying by `0.001` reports distances
in kilometres:

```{r sigma-plot-km, fig.cap = "***Fitted Gaussian kernel with x-axis relabelled in kilometres.***"}
plot(
  fit_gaussian,
  type = "sigma",
  distance_per_map_unit = 0.001,
  distance_unit = "km"
)
```

### Step 5 — Compare against a standard fixed-raster fit

The scale-aware fit is easiest to interpret when it is compared directly to the
same model without an estimated `sigma`.

```{r fixed-fit}
fit_fixed <- terradish(
  melip.Fst ~ forestcover,
  data = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  optimizer = "auto",
  leverage = FALSE,
  control = NewtonRaphsonControl(maxit = 12, verbose = FALSE)
)

fit_fixed
```

```{r compare-table}
data.frame(
  model = c("Gaussian scale-aware", "Fixed raster"),
  logLik = c(fit_gaussian$loglik, fit_fixed$loglik),
  AIC = c(fit_gaussian$aic, fit_fixed$aic)
)
```

At this stage, the main question is not simply "which AIC is smaller?" A more
useful question is:

- does `sigma.forestcover` stay away from the bounds?
- does the fitted scale look biologically plausible relative to the raster
  resolution?
- is the scale-aware fit at least competitive with the simpler fixed-raster
  version?

If the answer is yes, then adding another raster is more defensible.

### Step 6 — Summarize the one-raster fit

```{r summary}
summary(fit_gaussian)
```

For this conductance model, the coefficient table includes both:

- the conductance coefficient, and
- `sigma`

So the scale parameter is treated just like any other model parameter in the
joint optimization.

### Step 7 — Recover the fitted conductance surface

Once the model is fitted, the usual conductance helper still works.

```{r conductance}
fitted_conductance <- conductance(surface, fit_gaussian)
fitted_conductance
```

And we can plot the fitted conductance layer directly:

```{r conductance-plot, fig.cap = "***Fitted conductance surface from the Gaussian scale-aware model, with sampling sites shown in red.***"}
plot(fitted_conductance[[1]], main = "Fitted conductance")
points(coords, pch = 19, col = "red")
```

---

## Stage 2 — Add a second raster only after sigma looks sensible

Once the one-raster scale parameter looks stable enough to trust, we can add a
second raster and inspect whether the scale estimates remain interpretable.

```{r surface-two}
covariates_two <- c(melip.altitude, melip.forestcover)
names(covariates_two) <- c("altitude", "forestcover")

surface_two <- conductance_surface(
  covariates_two,
  coords,
  directions = 8,
  saveStack = TRUE
)
```

```{r fit-two}
fit_gaussian_two <- terradish(
  melip.Fst ~ forestcover + altitude,
  data = surface_two,
  conductance_model = gaussian_smoothed_loglinear_conductance(surface_two),
  measurement_model = mlpe,
  optimizer = "auto",
  leverage = FALSE,
  control = NewtonRaphsonControl(maxit = 12, verbose = FALSE)
)

fit_gaussian_two
```

```{r sigma-two}
gaussian_scale_summary(fit_gaussian_two)
```

This is exactly the point where the staged workflow helps. If one or more
sigma estimates now collapse to the lower bound, or drift toward the upper
bound, that is useful information. It suggests one of several things:

- the added raster may not need smoothing at all
- the data may not support estimating multiple scales simultaneously
- the smaller one-raster scale was not as stable as it first appeared

That does **not** mean the model failed. It means the scale parameters are
telling you that the data may prefer a simpler spatial structure.

In the small `melip` subset used here, that is exactly the kind of pattern to
watch for: the one-raster `forestcover` scale can be inspected on its own
before adding `altitude`. After `altitude` is added, one or both fitted scale
parameters may move toward the lower bound, which is a cue to
proceed cautiously rather than immediately adding more landscape terms.

---

## Important current limitations

This first version is intentionally focused. Right now it supports:

- additive formulas made of raw numeric raster names
- interactions among numeric raster names
- polynomial/arithmetic terms such as `I(forestcover^2)`
- one Gaussian scale parameter per selected raster
- joint optimization through analytic gradients and BFGS

It does **not yet** support:

- factor-valued raster layers
- leverage / `partial_X` diagnostics
- coarse-raster approximation for this conductance model

It does now support conditional-on-sigma marginal-effect plotting, so you can
inspect conductance and response-scale curves after fitting, but the leverage
machinery is still intentionally disabled.

---

## Where to go with your analysis

If this workflow looks reasonable for your data, practical steps are:

1. Fit a small one-raster model first and inspect `sigma`.
2. Compare it to a standard fixed-raster `loglinear_conductance` fit.
3. Add additional rasters only after the one-raster scale parameter behaves
   sensibly.
4. Treat boundary-hitting sigma estimates as a model-selection signal, not just
   an optimization nuisance.
5. Reproject rasters before making ecological distance interpretations, because
   longitude/latitude rasters report sigma in degrees.

This staged approach makes it easier to tell whether the data are truly
identifying a landscape scale of effect, rather than letting multiple scale
parameters compete in an over-complex first model.

---

## Companion vignettes

- **`vignette("getting-started", package = "terradish")`** — the core
  `terradish` workflow: scaling covariates, building a conductance surface,
  fitting and interpreting models, and the four plot types.
- **`vignette("model-comparison", package = "terradish")`** — likelihood ratio
  tests, AIC/AICc/BIC tables, likelihood surface visualization, and
  cross-validation.
- **`vignette("ibe-ibr-workflow", package = "terradish")`** — fitting isolation
  by environment (IBE) and isolation by resistance (IBR) simultaneously in the
  same model.
