---
title: "Getting Started with terradish"
author: "Bill Peterman"
output: 
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting Started with terradish}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4.5,
  warning  = FALSE,
  message  = FALSE
)

```

## What is terradish for?

`terradish` estimates conditional associations between raster covariates
(altitude, forest cover, roads, water bodies, etc.) and genetic distance or
covariance among sampled populations.

The core idea is **isolation by resistance (IBR)**: instead of assuming that
genetic distance simply tracks straight-line geographic distance (isolation by
distance, IBD), IBR represents the landscape as a heterogeneous graph. Each
cell has a fitted **conductance**, and neighboring cells define weighted edges.
Conductance is a relative model quantity, not a direct observation of movement,
migration, habitat suitability, or permeability.

If we can map conductance across the landscape as a function of measurable
covariates (rasters), we can:

1. Build a **conductance surface**: a weighted graph over the raster grid.
2. Build the weighted graph Laplacian $L(\theta)$ and its generalized inverse
   $E(\theta)=L(\theta)^+$.
3. Derive resistance distances from $E(\theta)$ when the measurement model
   requires them.
4. Link $E(\theta)$ or its resistance distances to an observed genetic matrix
   with a selected measurement likelihood.
5. Optimize the conductance and nuisance parameters by maximum likelihood.

`terradish` automates these steps and provides gradients, standard errors,
model-comparison tools, and visualization helpers.

---

## A conceptual road map

Before writing any code, it helps to see how the pieces fit together:

```
Raster covariates                   Sampling locations
(altitude, forestcover)             (GPS coordinates)
        |                                  |
        +-----> conductance_surface() <----+
                        |
                        |   (builds the weighted graph)
                        v
             conductance model -> L(theta) -> E(theta)
                         |
                         v
              measurement model  <-- distance or covariance likelihood
                        |
                        v
                   fitted model
                   coef(), summary(), fitted(),
                   conductance(), anova()
```

The **conductance model** maps raster values at each grid cell to a conductance
value. `loglinear_conductance` is the default: conductance =
exp(θ₁·altitude + θ₂·forestcover + ...). The parameters θ are what we
estimate.

The **measurement model** describes the statistical relationship between
pairwise resistance distances and pairwise genetic distances. Two choices cover
most use-cases:

- `leastsquares`: ordinary regression; fast, assumes independent errors.
- `mlpe`: "maximum likelihood population effects"; accounts for the
  non-independence among all pairwise measurements that share a sampling site.
  This is generally the preferred choice.

---

## Installation

```{r, eval = FALSE}
# Released version
install.packages("terradish")

# Development version
remotes::install_github("wpeterman/terradish")
```

No other package is required. `terradish` implements the MLPE measurement model
itself, so you do not need `corMLPE`. One optional package, `adegenet`, is used
only by `pca_dist()` for turning `genind` genotype objects into a distance
matrix; install it if you need that helper.

```{r, eval = FALSE}
install.packages("adegenet")
```

---

## Load the package and example data

```{r load}
library(terradish)
library(terra)
```

`terradish` ships with example data from a study of *Melipona* bees in Brazil. Load everything with one call:

```{r data}
data(melip)

# The rasters are stored as "packed" objects to comply with CRAN policy;
# unwrap() converts them back to usable SpatRaster / SpatVector objects.
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)
```

What do we have?

- **`melip.altitude`**: a raster of altitude across the study area, already
  rescaled to roughly the unit interval rather than left in meters.
- **`melip.forestcover`**: a raster of forest cover proportion (0-1).
- **`melip.coords`**: 37 sampling locations as a `SpatVector`.
- **`melip.Fst`**: a 37 x 37 matrix of pairwise F~ST~ (a standard measure
  of genetic differentiation; larger values mean more differentiation).

```{r inspect-data}
# How many sampling sites?
nrow(melip.coords)

# Quick look at the genetic distance matrix
# (only the first 5 x 5 block for readability)
round(melip.Fst[1:5, 1:5], 3)
```

F~ST~ is symmetric (the differentiation between site A and site B is the same
as between B and A), so only the lower triangle carries new information.

---

## Visualize the spatial data

```{r plot-covariates, fig.cap = "***Altitude raster with sampling sites.***"}
plot(melip.altitude, main = "Altitude (rescaled)")
points(melip.coords, pch = 19, cex = 0.8, col = 'red')
```

```{r plot-forestcover, fig.cap = "***Forest cover raster (0 = no forest, 1 = full cover).***"}
plot(melip.forestcover, main = "Forest cover")
points(melip.coords, pch = 19, cex = 0.8, col = 'red')
```

---

## Scale your rasters: an important step

Conductance in the loglinear model is exp(θ·x). If `x` (e.g. altitude in
meters) has a very large range, small changes in θ produce enormous changes in
conductance, making optimization numerically unstable. **Always scale continuous
covariates** before fitting.

```{r scale}
covariates <- c(melip.altitude, melip.forestcover)
names(covariates) <- c("altitude", "forestcover")
covariates <- scale_covariates(covariates)
```

By default, `scale_covariates()` subtracts each layer's mean and divides by its
standard deviation, giving every continuous covariate a mean of 0 and a
standard deviation of 1. It also stores the original-scale center and scale for
each layer, which lets `plot(fit, type = "marginal", data = surface)` label
x-axes in the original units automatically.

---

## Build the conductance surface

```{r surface}
surface <- conductance_surface(covariates, melip.coords, directions = 8)
```

`conductance_surface()` does three things:

1. Reads the raster values at every non-NA grid cell.
2. Identifies which cells correspond to the sampling locations.
3. Builds a graph where adjacent cells (here, using all 8 neighbors including
   diagonals) are connected by edges whose weights will be determined by the
   conductance model.

`directions = 8` is typical for landscape genetics; it allows diagonal movement
and is less sensitive to grid orientation than `directions = 4`.

```{r surface-check}
# How many grid cells are in the graph?
nrow(surface$x)

# The covariate data for the first few cells
head(surface$x)
```

### Optional: reduce graph size for large rasters

The example raster is small enough that we can fit directly. For larger
landscapes, graph size becomes the main driver of runtime. If your sampling
locations occupy only part of the raster, you can crop the raster to a buffered
sampling extent before the graph is constructed:

```{r crop-buffer-example, eval=FALSE}
surface_cropped <- conductance_surface(
  covariates,
  melip.coords,
  directions = 8,
  saveStack = TRUE,
  crop_buffer = 0.05
)
```

The buffer is measured in the raster's map units. In a projected raster this
might be meters; in a longitude/latitude raster it is degrees. Cropping is a
safe speed tool only when the cropped area still contains the landscape context
needed to estimate resistance distances. A good habit is to try a few buffer
widths and confirm that the fitted coefficients are stable.

Another speed option is a coarse-raster warm start. This fits the same model on
aggregated rasters first, then refines the estimate on the full-resolution
surface:

```{r coarse-raster-example, eval=FALSE}
surface_for_coarse <- conductance_surface(
  covariates,
  melip.coords,
  directions = 8,
  saveStack = TRUE
)

fit_coarse <- terradish(
  melip.Fst ~ forestcover + altitude,
  data = surface_for_coarse,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  approximation = "coarse_raster",
  approximation_control = list(
    factor = c(4, 2),
    exact_refine = TRUE
  )
)
```

With `exact_refine = TRUE`, the final reported model is still evaluated on the
original graph. The coarse fits are only used to find better starting values.

---

## Checking computational settings

For the small *Melipona* example the default settings run in seconds and no
tuning is needed. For larger landscapes (e.g., a high-resolution raster with
hundreds of thousands of graph vertices, or a spline conductance model that
expands to many parameters), the right optimizer and solver can meaningfully
reduce wall-clock time.

`terradish_assess_settings()` is an **advisory tool** for this. It:

1. **Profiles the graph**: counts vertices, edges, focal sites, and
   conductance parameters, and reports those back without running any
   optimization.
2. **Benchmarks the direct solver**: times several combinations of Cholesky
   factorization style and solve backend to find the fastest configuration for
   your specific graph.
3. **Runs short optimizer probes**: executes two Newton steps for each of
   three optimizer/line-search combinations (Newton + Hager-Zhang, BFGS +
   Hager-Zhang, BFGS + Armijo) and reports which was fastest.
4. **Optionally probes coarse-raster warm-starts**: checks whether fitting on
   an aggregated raster first produces an acceptable speedup.
5. **Returns `$recommended`**: a named list of the best settings that can be
   passed directly to `terradish()`.

The function does **not** change any package defaults and does **not** affect
parameter estimation; only the computational path differs.

```{r assess-settings-demo, eval = FALSE}
# saveStack = TRUE is required for the coarse-raster probe
surface_for_assess <- conductance_surface(
  covariates, melip.coords,
  directions = 8, saveStack = TRUE
)

assessment <- terradish_assess_settings(
  melip.Fst ~ forestcover + altitude,
  data              = surface_for_assess,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  probe_maxit       = 2,      # keep small: probes compare early-iteration cost
  coarse_probe      = TRUE,   # also time the coarse-raster warm-start path
  verbose           = TRUE    # print progress as probes run
)

# Print a full summary: graph profile, probe results, recommended settings
assessment
```

The printed output has three main parts:

```
terradish setting assessment
Graph: 22443 vertices, 88844 edges, 37 focal sites
Conductance parameters: 2

Recommended settings:
  optimizer:   newton  (fastest acceptable probe: newton_hager_zhang)
  line search: HagerZhangControl
  solver:      direct  (fastest direct solver probe: supernodal_ll + cholmod_cpp_cached)
  approximation: none  (exact full-resolution fit)

Defaults comparison: Assessment matched terradish defaults.
```

- **Graph profile**: tells you how large your problem is. A vertex count above
  ~50 000 is where solver choice starts to matter; above ~500 000 it is often
  the dominant cost.
- **Recommended settings**: the fastest combination found by the probes.
  `"(fastest acceptable probe: ...)"` shows which probe won and why.
- **Defaults comparison**: a one-line summary of whether the assessment
  recommends anything different from what `terradish()` would use by default.
  If it says `"Assessment matched terradish defaults"`, just call `terradish()`
  normally.

### Using `$recommended` in the final fit

The `$recommended` list maps directly onto `terradish()` arguments:

```{r assess-use, eval = FALSE}
fit_tuned <- terradish(
  melip.Fst ~ forestcover + altitude,
  data              = surface_for_assess,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = assessment$recommended$optimizer,
  control           = assessment$recommended$control,
  solver            = assessment$recommended$solver,
  solver_control    = assessment$recommended$solver_control,
  approximation     = assessment$recommended$approximation,
  approximation_control = assessment$recommended$approximation_control
)
```

You can also inspect and edit the list before using it. For example, to override
the recommended `maxit` without losing the solver recommendation:

```{r assess-edit, eval = FALSE}
my_control <- assessment$recommended$control
my_control$maxit <- 50          # allow more iterations than the probe default

fit_tuned2 <- terradish(
  melip.Fst ~ forestcover + altitude,
  data    = surface_for_assess,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  optimizer      = assessment$recommended$optimizer,
  control        = my_control,
  solver         = assessment$recommended$solver,
  solver_control = assessment$recommended$solver_control
)
```

### When assessment is most useful

| Situation | Assessment benefit |
|---|---|
| Graph < 10 000 vertices | Minimal; defaults usually converge in seconds |
| Graph 10 000 to 100 000 vertices | Solver benchmark often helps |
| Graph > 100 000 vertices | All three probes are worthwhile |
| Spline conductance (`df ≥ 4`) | Parameter expansion can shift optimizer recommendation from Newton to BFGS |
| Fitting many candidate models | Run assessment once, reuse `$recommended` for all fits |
| Coarse-raster approximation desired | Set `coarse_probe = TRUE` to check whether warm-start pays off |

### Troubleshooting direct-solver failures

If a fit stops with a `CHOLMOD` warning or a
`non-positive-definite reduced Laplacian` message, the direct sparse Cholesky
solver could not factorize the graph at the current conductance values. This is
usually a numerical or data-preparation problem rather than a genetic-distance
model result.

Immediate checks:

- Confirm the raster layers used by the conductance model are finite on the
  retained graph cells. Missing, infinite, or extremely large values can produce
  unusable conductances after the exponential transform.
- Scale continuous covariates with `scale_covariates()` and start conductance
  parameters near zero. Very large positive or negative starts can make some
  graph edges effectively infinite or zero.
- Re-run `terradish_assess_settings(..., verbose = TRUE)` to inspect graph size,
  direct-solver settings, and the optimizer path before launching a long fit.
- While diagnosing the surface, try `solver = "auto"` or `solver = "amg"`.
  These paths do not fix invalid conductance values, but they can help separate
  a direct Cholesky factorization issue from a broader model setup problem.

---

## Fit your first model

We will start with the simplest case: are altitude and forest cover conditionally
associated with genetic differentiation through the fitted conductance model?

```{r fit-leastsquares}
fit_ls <- terradish(
  melip.Fst ~ forestcover + altitude,  # formula: response ~ covariates
  data              = surface,          # the conductance surface
  conductance_model = loglinear_conductance,
  measurement_model = leastsquares      # ordinary regression
)
```

The formula works like a standard R formula:

- The **left-hand side** is the matrix of observed genetic distances.
- The **right-hand side** lists which raster layers parameterize conductance.

`leastsquares` is the quickest measurement model. It treats the pairwise
genetic distances as independent observations regressed on resistance distances.
That is not strictly correct (measurements sharing a site are correlated), but
it gives a useful exploratory or diagnostic fit.

```{r summary-ls}
summary(fit_ls)
```

**How to read the output:**

- **theta**: the conductance parameters. A positive θ for a covariate means
  that higher values are associated with higher fitted relative conductance,
  conditional on other terms and the graph. A negative θ means lower fitted
  relative conductance. Neither sign is a direct estimate of movement or a
  causal landscape effect.
- **phi**: nuisance parameters of the measurement model. For MLPE, these are
  the intercept, slope, log precision, and an unconstrained correlation
  parameter. The actual residual precision is `exp(tau)`, and the actual
  shared-site correlation is `plogis(rho) / 2`. These parameters are profiled
  out automatically and generally are not the primary inferential target.
- The standard errors and z-statistics help gauge parameter uncertainty.

---

## Upgrade to the MLPE measurement model

Pairwise genetic distances are not independent: the distance between sites A-B
and A-C both involve site A. `mlpe` ("maximum likelihood population effects")
accounts for this correlation structure and is the recommended choice for
inference.

```{r fit-mlpe}
fit_mlpe <- terradish(
  melip.Fst ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe
)

summary(fit_mlpe)
```

Compare the point estimates and standard errors between `fit_ls` and
`fit_mlpe`. They are often similar, but MLPE standard errors better reflect the
shared-site dependence represented by the model.

---

## Choosing a measurement model

`terradish` provides four measurement models. The right choice depends on
the form of your genetic data and how you want to handle the non-independence
of pairwise observations.

| Model | Input `S` | Requires `nu` | Correlation structure |
|---|---|---|---|
| `leastsquares` | distance matrix | no | none (independent errors) |
| `mlpe` | distance matrix | no | MLPE shared-site correction |
| `generalized_wishart` | admissible squared-distance matrix | yes | generalized Wishart |
| `wishart_covariance` | **covariance** matrix | yes | Wishart |

`leastsquares` is the fastest and simplest. It is a reasonable first pass, but
its standard errors can be too small because it ignores shared-site correlation.

`mlpe` corrects for the non-independence that arises because every pair (A-B)
and (A-C) both involve site A. It is the default recommendation for
distance-matrix data when the effective marker degrees of freedom are not
recorded.

`generalized_wishart` evaluates a generalized Wishart likelihood for an
admissible squared-distance response that is coherently related to a centered
positive semidefinite covariance matrix. It is not valid for an arbitrary
dissimilarity matrix. It requires `nu`, the effective Wishart degrees of
freedom for `S`. For biallelic SNPs this is usually
the retained polymorphic SNP count (reduced for linkage disequilibrium). For
microsatellites, use the **number of loci** as the conservative default; the
within-locus allele correlations mean that `sum(K_l - 1)` should be treated as
a larger sensitivity value, not as a known effective count; see
`vignette("wishart-covariance", package = "terradish")` for a full discussion.
The input must satisfy the model's distance-to-covariance admissibility
conditions; common labels such as F~ST~ do not establish that automatically.
Use `check_distance_response()` on the matrix that you intend to analyze.
`terradish()` performs the same check before fitting and stops rather than
silently transforming an inadmissible response.

`wishart_covariance` is the Wishart analogue for data supplied as a
**covariance** matrix (e.g. from `cov_from_biallelic()`) rather than a
distance matrix. It also requires `nu`.

### Fitting with `generalized_wishart`

```{r fit-wishart}
# Construct an admissible squared-distance response from a simulated positive
# semidefinite covariance matrix. The same nu is used for simulation and fit.
nu_gw <- 50
sim_gw <- simulate_covariance_response(
  theta   = c(forestcover = 0.4, altitude = -0.3),
  formula = ~ forestcover + altitude,
  data    = surface,
  tau     = 0.8,
  sigma   = 0.2,
  nu      = nu_gw,
  seed    = 1
)
gw_distance <- dist_from_cov(sim_gw$covariance)
gw_check <- check_distance_response(gw_distance)
gw_check[c("admissible", "relative_min_eigenvalue", "n_negative")]

fit_gw <- terradish(
  gw_distance ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = generalized_wishart,
  nu                = nu_gw
)

summary(fit_gw)
```

**How to read the response check:** `admissible = TRUE` means the centered Gram
matrix is positive semidefinite within the documented numerical tolerance.
`relative_min_eigenvalue` reports the smallest eigenvalue relative to the
largest; a tiny negative value can reflect rounding. `n_negative` counts only
violations that exceed the tolerance. This diagnostic establishes mathematical
compatibility with the likelihood, not biological suitability of the chosen
genetic summary.

The nuisance parameters for `generalized_wishart` are `tau` (the weight on the
graph covariance kernel) and `sigma` (a log-scale identity component that
captures diagonal variation not represented by that kernel). `sigma` is not a
Gaussian smoothing scale or a direct estimate of genetic drift.
A `tau` estimate near zero indicates no detectable graph-kernel contribution
under this response and model.

You can compare models that use the **same response, sites, graph domain,
likelihood family, and pair subset** with `aic_table()`; Wishart comparisons
must also use the same `nu`. `leastsquares` and `mlpe` both model pairwise
distances with a Gaussian likelihood, so they can be compared when those
conditions hold. `generalized_wishart`
evaluates a Wishart likelihood on the same distance matrix, which has a
different functional form and different normalizing constants; its AIC values
are not directly comparable to those of `leastsquares` or `mlpe`.  Use
`aic_table()` to rank models within a family (e.g., all `mlpe` fits differing
in their conductance formula, or all `generalized_wishart` fits differing in
their conductance formula).

```{r compare-measurement-models}
# Valid: both use a Gaussian likelihood on pairwise distances.
aic_table(
  list(fit_ls, fit_mlpe),
  mod_names = c("leastsquares", "mlpe")
)
```

---

## Extract and interpret coefficients

```{r coef}
coef(fit_mlpe)
```

The log-linear conductance model means:

- θ~forestcover~ > 0: higher forest cover is associated with higher fitted
  relative conductance, conditional on altitude and the model.
- θ~altitude~ < 0: higher altitude is associated with lower fitted relative
  conductance, conditional on forest cover and the model.

You can convert back from the log scale to a relative conductance ratio. For
example, a one standard-deviation increase in forest cover multiplies
conductance by exp(θ~forestcover~):

```{r interpret}
theta <- coef(fit_mlpe)
cat("One SD increase in forest cover multiplies conductance by",
    round(exp(theta["forestcover"]), 2), "\n")
cat("One SD increase in altitude multiplies conductance by",
    round(exp(theta["altitude"]), 2), "\n")
```

---

## Visualize the fit

### Observed vs. fitted genetic distances

The `plot()` S3 method with `type = "fit"` produces an observed-vs-fitted
scatter plot with a least-squares regression line:

```{r plot-fit, fig.cap = "***Observed vs. resistance-distance-fitted genetic distances.***"}
plot(fit_mlpe, type = "fit")
```

### Fitted conductance surface

`type = "surface"` renders the point-estimate conductance surface alongside its
lower and upper confidence interval bounds in a single three-panel figure. All
panels share the same color scale so relative differences are easy to read:

```{r conductance-surface, fig.cap = "***Estimated conductance surface with 95% confidence interval bounds.***"}
plot(fit_mlpe, type = "surface", data = surface)
```

Brighter cells have higher fitted relative conductance. Panels where the lower
and upper bounds are close together indicate narrower conditional uncertainty
under the fitted model; they do not directly measure movement or permeability.

### Marginal associations with predicted genetic distance (default)

The default plot (`type = "marginal_response"`) shows how each raster
covariate is associated with the **predicted genetic distance** (the response
variable itself) rather than the intermediate conductance scale. Because
genetic distance is the observed response that the model fits, this is usually
the most directly interpretable view.

```{r marginal-effects, fig.cap = "***Marginal association of each covariate with predicted genetic distance, with predictive bands.***"}
plot(fit_mlpe, data = surface)
```

**How to read these plots**

Each panel sweeps one covariate across its observed range while all other
covariates are kept at their *observed* values at every grid cell (not at a
single mean value). The curve is therefore a weighted average over the actual
landscape, which matters most when the associations of different covariates
interact.

The **central curve** is the expected genetic distance at the given covariate
value, marginalized over the rest of the landscape.

The **shaded predictive band** combines three sources of uncertainty:

1. *Theta (conductance parameter) uncertainty*: imprecision in the estimated
   conductance parameters θ, propagated through the resistance-distance
   calculation via the delta method.
2. *Phi (measurement-model parameter) uncertainty*: imprecision in the
   nuisance parameters (intercept α, IBR slope β, log precision τ, and the
   unconstrained correlation parameter ρ), again via the delta method. The
   corresponding shared-site correlation is `plogis(rho) / 2`.
3. *Residual variation*: the irreducible scatter in pairwise genetic
   distances around the fitted mean, with variance `exp(-tau)`.

A **rising curve** means that higher covariate values are associated with
higher predicted genetic distance under the fitted model. A **declining curve**
means the conditional predicted association is negative. These plots do not
establish a causal permeability or connectivity mechanism.

> **Note:** Quadratic terms (`I(x^2)`) and interactions (`x * z`) in the
> formula are handled automatically. For an interaction, each panel holds the
> other covariate at its observed values across all grid cells, so the curve
> represents the average fitted association of the focal covariate over the joint
> distribution of the landscape.

### Marginal associations on the conductance scale

If you prefer to see the fitted association on the underlying conductance surface (without
residual variation in the band, and on the conductance rather than genetic
distance scale), use `type = "marginal"`:

```{r marginal-conductance, fig.cap = "***Marginal association of each covariate with conductance, with a 95% CI.***"}
plot(fit_mlpe, type = "marginal", data = surface)
```

Here the band is a **confidence interval** (delta-method, 95% pointwise) on
the mean conductance, not a predictive interval; it reflects only parameter
uncertainty in θ, not residual variation. A rising curve means higher covariate
values are associated with higher fitted conductance; a declining curve means
the fitted conditional association is negative.

### Support-constrained prediction and plotting

If a covariate has sparse focal support at one tail, marginal curves and
conductance maps can be dominated by extrapolation. `terradish` now supports
focal-support clamping directly in both `plot()` and `conductance()`:

```{r support-clamp, eval = FALSE}
# Clamp selected covariates to focal-site quantile support before plotting.
plot(
  fit_mlpe,
  type = "marginal",
  data = surface,
  support = "focal",
  support_probs = c(0.01, 0.99),
  clamp_covariates = c("altitude", "forestcover")
)

# Apply the same support rule to raster conductance prediction.
cond_focal <- conductance(
  surface,
  fit_mlpe,
  support = "focal",
  support_probs = c(0.01, 0.99),
  clamp_covariates = c("altitude", "forestcover")
)
```

Use `support = "none"` (default) for full-domain evaluation, and
`support = "focal"` when you want inference constrained to focal sampling
support.

---

## Quick-reference: the full workflow in one block

```{r full-workflow, eval = FALSE}
library(terradish)
library(terra)

# 1. Load data
data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

# 2. Scale covariates and build surface
covariates <- c(melip.forestcover, melip.altitude)
names(covariates) <- c("forestcover", "altitude")
covariates <- scale_covariates(covariates)

surface <- conductance_surface(covariates, melip.coords,
                               directions = 8, saveStack = TRUE)

# 2b. (Optional) Profile graph and identify fastest settings
#     Most useful for large graphs (> 10 000 vertices) or spline models.
assessment <- terradish_assess_settings(
  melip.Fst ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  probe_maxit       = 2
)
cat(assessment$comparison$summary, "\n")  # one-line verdict
rec <- assessment$recommended             # reuse for all fits below

# 3. Fit model - choose a measurement model:
#   leastsquares      : fast, independent errors (distance matrix)
#   mlpe              : shared-site correlation correction (distance matrix)
#   generalized_wishart: Wishart model for an admissible squared-distance response;
#                        requires nu
#   wishart_covariance : Wishart model, requires nu (covariance matrix)
#   For Wishart IBE or site-level environmental covariance kernels, use
#   wishart_covariates(); pair_subset_measurement_model() is for pairwise
#   regression rows and does not apply to the full-matrix Wishart likelihood.
fit <- terradish(melip.Fst ~ forestcover + altitude,
                 data              = surface,
                 conductance_model = loglinear_conductance,
                 measurement_model = mlpe,
                 optimizer         = rec$optimizer,
                 control           = rec$control,
                 solver            = rec$solver,
                 solver_control    = rec$solver_control,
                 approximation     = rec$approximation,
                 approximation_control = rec$approximation_control)

# For generalized_wishart, derive and check a squared-distance response:
# S_cov <- cov_from_genetic_data(genotype_matrix, groups = population)
# S_dist <- dist_from_cov(S_cov)
# check_distance_response(S_dist)
# fit_gw <- terradish(S_dist ~ forestcover + altitude, data = surface,
#                     conductance_model = loglinear_conductance,
#                     measurement_model = generalized_wishart,
#                     nu = effective_marker_df)

# 4. Summarize and visualize
summary(fit)

plot(fit, type = "fit")                                # observed vs. fitted scatter
plot(fit, type = "surface", data = surface)            # conductance map + 95% CI panels
plot(fit, data = surface)                              # marginal associations on genetic distance (default)
plot(fit, type = "marginal", data = surface)           # marginal associations on conductance scale

# 5. Optional: focal-support clamping for robust tails
plot(fit, type = "marginal", data = surface,
     support = "focal", support_probs = c(0.01, 0.99),
     clamp_covariates = c("forestcover", "altitude"))

cond <- conductance(surface, fit,
                    support = "focal", support_probs = c(0.01, 0.99),
                    clamp_covariates = c("forestcover", "altitude"))
```

## Summary of key functions

| Function | Purpose |
|----------|---------|
| `scale_covariates()` | Center and scale raster layers to mean 0, sd 1 before fitting |
| `scale_to_0_1()` | Rescale layers to the unit interval (needed by `linear_conductance()`) |
| `crop_to_focal_buffer()` | Trim the raster extent to a buffer around the focal sites |
| `conductance_surface()` | Build the weighted graph linking raster cells and focal sites |
| `terradish_assess_settings()` | Profile the graph and recommend solver and optimizer settings |
| `terradish()` | Fit the model by maximum likelihood |
| `loglinear_conductance()` | Default conductance model, `c = exp(X theta)` |
| `leastsquares()`, `mlpe()` | Measurement models for a genetic distance matrix |
| `check_distance_response()` | Diagnose squared-Euclidean admissibility before a generalized-Wishart fit |
| `generalized_wishart()`, `wishart_covariance()` | Wishart models for squared-distance and covariance responses, respectively; both need `nu` |
| `summary()`, `coef()`, `confint()` | Estimates, standard errors, and intervals on the log-conductance scale |
| `plot()` | Observed vs. fitted, conductance surface, and marginal associations |
| `conductance()` | Extract the fitted conductance surface as a `SpatRaster` |

## See also

Each companion vignette assumes the workflow above and takes it further.

- **`vignette("model-comparison", package = "terradish")`**, testing and
  ranking competing models with likelihood ratio tests, AIC, AICc, BIC, and
  cross-validation.
- **`vignette("ibe-ibr-workflow", package = "terradish")`**, fitting
  isolation by environment and isolation by resistance together in one model.
- **`vignette("wishart-covariance", package = "terradish")`**, building a
  genetic covariance matrix from raw genotypes and fitting it with
  `wishart_covariance()`, including how to choose `nu`.
- **`vignette("spline-conductance", package = "terradish")`**, non-linear
  conductance responses through spline terms such as `s(altitude, df = 4)`.
- **`vignette("gaussian-scale-optimization", package = "terradish")`**,
  estimating a Gaussian raster-smoothing scale inside the model rather than by
  pre-smoothing.
- **`vignette("hierarchical-conductance", package = "terradish")`**, adding a
  diagnostic residual log-conductance field and evaluating spatial confounding.
- **`vignette("directional-conductance", package = "terradish")`**,
  fitting antisymmetric edge-rate bias through symmetric commute time.
- **`vignette("large-landscapes", package = "terradish")`**, solvers,
  curvature, and exact graph reduction for rasters with hundreds of thousands
  of cells.
- **`vignette("simulation-design", package = "terradish")`**, simulating
  covariance data to choose sample sizes and marker counts before fieldwork.
