---
title: "Getting Started with terradish"
author: "Bill Peterman"
output: 
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting Started with terradish}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4.5,
  warning  = FALSE,
  message  = FALSE
)
```

## What is terradish for?

`terradish` estimates how landscape features (altitude, forest cover, roads,
water bodies, etc.) shape genetic differentiation among populations.

The core idea is **isolation by resistance (IBR)**: instead of assuming that
genetic distance simply tracks straight-line geographic distance (isolation by
distance, IBD), IBR recognizes that the landscape is heterogeneous. Some cells
are easy to move through; others are barriers. The ease of movement through a
cell is its **conductance** (the inverse of **resistance**).

If we can map conductance across the landscape as a function of measurable
covariates (rasters), we can:

1. Build a **conductance surface**: a weighted graph over the raster grid.
2. Compute **resistance distances** among sampling locations. These are
   effective distances that account for the landscape structure.
3. Fit a **statistical model** relating those resistance distances to observed
   genetic distances.
4. Optimize the conductance parameters by maximum likelihood.

`terradish` automates all four steps and provides gradients, standard errors,
model-comparison tools, and visualization helpers.

---

## A conceptual road map

Before writing any code, it helps to see how the pieces fit together:

```
Raster covariates          Sampling locations
(altitude, forestcover)    (GPS coordinates)
        |                          |
        +-----> conductance_surface() <----+
                        |
                        |   (builds the weighted graph)
                        v
              terradish(formula, surface,
                        conductance_model,  <-- how covariates -> conductance
                        measurement_model)  <-- how resistance dist -> genetic dist
                        |
                        v
                   fitted model
                   coef(), summary(), fitted(),
                   conductance(), anova()
```

The **conductance model** maps raster values at each grid cell to a conductance
value. `loglinear_conductance` is the default: conductance =
exp(╬ĖŌéü┬Ęaltitude + ╬ĖŌéé┬Ęforestcover + ...). The parameters ╬Ė are what we
estimate.

The **measurement model** describes the statistical relationship between
pairwise resistance distances and pairwise genetic distances. Two choices cover
most use-cases:

- `leastsquares`: ordinary regression; fast, assumes independent errors.
- `mlpe`: "maximum likelihood population effects"; accounts for the
  non-independence among all pairwise measurements that share a sampling site.
  This is generally the preferred choice.

---

## Installation

```{r, eval = FALSE}
# Install from local source (recommended while the package is in development)
devtools::install()
# or
remotes::install_local(".")

# corMLPE is also required (not on CRAN)
devtools::install_github("nspope/corMLPE")
```

---

## Load the package and example data

```{r load}
library(terradish)
library(terra)
```

`terradish` ships with example data from a study of *Melipona* bees in
Mexico. Load everything with one call:

```{r data}
data(melip)

# The rasters are stored as "packed" objects to comply with CRAN policy;
# unwrap() converts them back to usable SpatRaster / SpatVector objects.
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)
```

What do we have?

- **`melip.altitude`**: a raster of elevation (meters) across the study area.
- **`melip.forestcover`**: a raster of forest cover proportion (0ŌĆō1).
- **`melip.coords`**: 34 sampling locations as a `SpatVector`.
- **`melip.Fst`**: a 34 ├Ś 34 matrix of pairwise F~ST~ (a standard measure
  of genetic differentiation; larger values mean more differentiation).

```{r inspect-data}
# How many sampling sites?
nrow(melip.coords)

# Quick look at the genetic distance matrix
# (only the first 5 x 5 block for readability)
round(melip.Fst[1:5, 1:5], 3)
```

F~ST~ is symmetric (the differentiation between site A and site B is the same
as between B and A), so only the lower triangle carries new information.

---

## Visualize the spatial data

```{r plot-covariates, fig.cap = "***Altitude raster with sampling sites.***"}
plot(melip.altitude, main = "Altitude (m)")
points(melip.coords, pch = 19, cex = 0.8, col = 'red')
```

```{r plot-forestcover, fig.cap = "***Forest cover raster (0 = no forest, 1 = full cover).***"}
plot(melip.forestcover, main = "Forest cover")
points(melip.coords, pch = 19, cex = 0.8, col = 'red')
```

---

## Scale your rasters: an important step

Conductance in the loglinear model is exp(╬Ė┬Ęx). If `x` (e.g. altitude in
meters) has a very large range, small changes in ╬Ė produce enormous changes in
conductance, making optimization numerically unstable. **Always scale continuous
covariates** before fitting.

```{r scale}
covariates <- c(melip.altitude, melip.forestcover)
names(covariates) <- c("altitude", "forestcover")
covariates <- scale_covariates(covariates)
```

By default, `scale_covariates()` subtracts each layer's mean and divides by its
standard deviation, giving every continuous covariate a mean of 0 and a
standard deviation of 1. It also stores the original-scale center and scale for
each layer, which lets `plot(fit, type = "marginal", data = surface)` label
x-axes in the original units automatically.

---

## Build the conductance surface

```{r surface}
surface <- conductance_surface(covariates, melip.coords, directions = 8)
```

`conductance_surface()` does three things:

1. Reads the raster values at every non-NA grid cell.
2. Identifies which cells correspond to the sampling locations.
3. Builds a graph where adjacent cells (here, using all 8 neighbors including
   diagonals) are connected by edges whose weights will be determined by the
   conductance model.

`directions = 8` is typical for landscape genetics; it allows diagonal movement
and is less sensitive to grid orientation than `directions = 4`.

```{r surface-check}
# How many grid cells are in the graph?
nrow(surface$x)

# The covariate data for the first few cells
head(surface$x)
```

### Optional: reduce graph size for large rasters

The example raster is small enough that we can fit directly. For larger
landscapes, graph size becomes the main driver of runtime. If your sampling
locations occupy only part of the raster, you can crop the raster to a buffered
sampling extent before the graph is constructed:

```{r crop-buffer-example, eval=FALSE}
surface_cropped <- conductance_surface(
  covariates,
  melip.coords,
  directions = 8,
  saveStack = TRUE,
  crop_buffer = 0.05
)
```

The buffer is measured in the raster's map units. In a projected raster this
might be meters; in a longitude/latitude raster it is degrees. Cropping is a
safe speed tool only when the cropped area still contains the landscape context
needed to estimate resistance distances. A good habit is to try a few buffer
widths and confirm that the fitted coefficients are stable.

Another speed option is a coarse-raster warm start. This fits the same model on
aggregated rasters first, then refines the estimate on the full-resolution
surface:

```{r coarse-raster-example, eval=FALSE}
surface_for_coarse <- conductance_surface(
  covariates,
  melip.coords,
  directions = 8,
  saveStack = TRUE
)

fit_coarse <- terradish(
  melip.Fst ~ forestcover + altitude,
  data = surface_for_coarse,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  approximation = "coarse_raster",
  approximation_control = list(
    factor = c(4, 2),
    exact_refine = TRUE
  )
)
```

With `exact_refine = TRUE`, the final reported model is still evaluated on the
original graph. The coarse fits are only used to find better starting values.

---

## Checking computational settings

For the small *Melipona* example the default settings run in seconds and no
tuning is needed. For larger landscapes ŌĆö say, a high-resolution raster with
hundreds of thousands of graph vertices, or a spline conductance model that
expands to many parameters ŌĆö the right optimizer and solver can meaningfully
reduce wall-clock time.

`terradish_assess_settings()` is an **advisory tool** for this. It:

1. **Profiles the graph** ŌĆö counts vertices, edges, focal sites, and
   conductance parameters, and reports those back without running any
   optimization.
2. **Benchmarks the direct solver** ŌĆö times several combinations of Cholesky
   factorization style and solve backend to find the fastest configuration for
   your specific graph.
3. **Runs short optimizer probes** ŌĆö executes two Newton steps for each of
   three optimizer/line-search combinations (Newton + Hager-Zhang, BFGS +
   Hager-Zhang, BFGS + Armijo) and reports which was fastest.
4. **Optionally probes coarse-raster warm-starts** ŌĆö checks whether fitting on
   an aggregated raster first produces an acceptable speedup.
5. **Returns `$recommended`** ŌĆö a named list of the best settings that can be
   passed directly to `terradish()`.

The function does **not** change any package defaults and does **not** affect
parameter estimation; only the computational path differs.

```{r assess-settings-demo, eval = FALSE}
# saveStack = TRUE is required for the coarse-raster probe
surface_for_assess <- conductance_surface(
  covariates, melip.coords,
  directions = 8, saveStack = TRUE
)

assessment <- terradish_assess_settings(
  melip.Fst ~ forestcover + altitude,
  data              = surface_for_assess,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  probe_maxit       = 2,      # keep small: probes compare early-iteration cost
  coarse_probe      = TRUE,   # also time the coarse-raster warm-start path
  verbose           = TRUE    # print progress as probes run
)

# Print a full summary: graph profile, probe results, recommended settings
assessment
```

The printed output has three main parts:

```
terradish setting assessment
Graph: 1234 vertices, 4802 edges, 34 focal sites
Conductance parameters: 2

Recommended settings:
  optimizer:   newton  (fastest acceptable probe: newton_hager_zhang)
  line search: HagerZhangControl
  solver:      direct  (fastest direct solver probe: supernodal_ll + cholmod_cpp_cached)
  approximation: none  (exact full-resolution fit)

Defaults comparison: Assessment matched terradish defaults.
```

- **Graph profile** ŌĆö tells you how large your problem is. A vertex count above
  ~50 000 is where solver choice starts to matter; above ~500 000 it is often
  the dominant cost.
- **Recommended settings** ŌĆö the fastest combination found by the probes.
  `"(fastest acceptable probe: ...)"` shows which probe won and why.
- **Defaults comparison** ŌĆö a one-line summary of whether the assessment
  recommends anything different from what `terradish()` would use by default.
  If it says `"Assessment matched terradish defaults"`, just call `terradish()`
  normally.

### Using `$recommended` in the final fit

The `$recommended` list maps directly onto `terradish()` arguments:

```{r assess-use, eval = FALSE}
fit_tuned <- terradish(
  melip.Fst ~ forestcover + altitude,
  data              = surface_for_assess,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = assessment$recommended$optimizer,
  control           = assessment$recommended$control,
  solver            = assessment$recommended$solver,
  solver_control    = assessment$recommended$solver_control,
  approximation     = assessment$recommended$approximation,
  approximation_control = assessment$recommended$approximation_control
)
```

You can also inspect and edit the list before using it. For example, to override
the recommended `maxit` without losing the solver recommendation:

```{r assess-edit, eval = FALSE}
my_control <- assessment$recommended$control
my_control$maxit <- 50          # allow more iterations than the probe default

fit_tuned2 <- terradish(
  melip.Fst ~ forestcover + altitude,
  data    = surface_for_assess,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  optimizer      = assessment$recommended$optimizer,
  control        = my_control,
  solver         = assessment$recommended$solver,
  solver_control = assessment$recommended$solver_control
)
```

### When assessment is most useful

| Situation | Assessment benefit |
|---|---|
| Graph < 10 000 vertices | Minimal ŌĆö defaults usually converge in seconds |
| Graph 10 000 ŌĆō 100 000 vertices | Solver benchmark often helps |
| Graph > 100 000 vertices | All three probes are worthwhile |
| Spline conductance (`df Ōēź 4`) | Parameter expansion can shift optimizer recommendation from Newton to BFGS |
| Fitting many candidate models | Run assessment once, reuse `$recommended` for all fits |
| Coarse-raster approximation desired | Set `coarse_probe = TRUE` to check whether warm-start pays off |

### Troubleshooting direct-solver failures

If a fit stops with a `CHOLMOD` warning or a
`non-positive-definite reduced Laplacian` message, the direct sparse Cholesky
solver could not factorize the graph at the current conductance values. This is
usually a numerical or data-preparation problem rather than a genetic-distance
model result.

Immediate checks:

- Confirm the raster layers used by the conductance model are finite on the
  retained graph cells. Missing, infinite, or extremely large values can produce
  unusable conductances after the exponential transform.
- Scale continuous covariates with `scale_covariates()` and start conductance
  parameters near zero. Very large positive or negative starts can make some
  graph edges effectively infinite or zero.
- Re-run `terradish_assess_settings(..., verbose = TRUE)` to inspect graph size,
  direct-solver settings, and the optimizer path before launching a long fit.
- While diagnosing the surface, try `solver = "auto"` or `solver = "amg"`.
  These paths do not fix invalid conductance values, but they can help separate
  a direct Cholesky factorization issue from a broader model setup problem.

---

## Fit your first model

We will start with the simplest case: do both altitude and forest cover predict
genetic differentiation?

```{r fit-leastsquares}
fit_ls <- terradish(
  melip.Fst ~ forestcover + altitude,  # formula: response ~ covariates
  data              = surface,          # the conductance surface
  conductance_model = loglinear_conductance,
  measurement_model = leastsquares      # ordinary regression
)
```

The formula works like a standard R formula:

- The **left-hand side** is the matrix of observed genetic distances.
- The **right-hand side** lists which raster layers drive conductance.

`leastsquares` is the quickest measurement model. It treats the pairwise
genetic distances as independent observations regressed on resistance distances.
That is not strictly correct (measurements sharing a site are correlated), but
it gives a useful quick answer.

```{r summary-ls}
summary(fit_ls)
```

**How to read the output:**

- **theta**: the conductance parameters. A positive ╬Ė for a covariate means
  that higher values of that covariate increase conductance (i.e., make movement
  easier), which *decreases* resistance distance. A negative ╬Ė means the
  covariate acts as a barrier.
- **phi**: nuisance parameters of the measurement model (intercept, slope,
  residual precision). These are profiled out automatically and you generally
  do not need to interpret them directly.
- The standard errors and z-statistics help gauge parameter uncertainty.

---

## Upgrade to the MLPE measurement model

Pairwise genetic distances are not independent: the distance between sites AŌĆōB
and AŌĆōC both involve site A. `mlpe` ("maximum likelihood population effects")
accounts for this correlation structure and is the recommended choice for
inference.

```{r fit-mlpe}
fit_mlpe <- terradish(
  melip.Fst ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe
)

summary(fit_mlpe)
```

Compare the point estimates and standard errors between `fit_ls` and
`fit_mlpe`. They are often similar, but MLPE standard errors better reflect the
actual precision of your estimates.

---

## Choosing a measurement model

`terradish` provides four measurement models. The right choice depends on
the form of your genetic data and how you want to handle the non-independence
of pairwise observations.

| Model | Input `S` | Requires `nu` | Correlation structure |
|---|---|---|---|
| `leastsquares` | distance matrix | no | none (independent errors) |
| `mlpe` | distance matrix | no | MLPE shared-site correction |
| `generalized_wishart` | distance matrix | yes | Wishart (McCullagh 2009) |
| `wishart_covariance` | **covariance** matrix | yes | Wishart |

`leastsquares` is the fastest and simplest. It is a reasonable first pass but
underestimates standard errors because it ignores the shared-site correlation.

`mlpe` corrects for the non-independence that arises because every pair (AŌĆōB)
and (AŌĆōC) both involve site A. It is the default recommendation for
distance-matrix data when the number of genetic markers is not recorded.

`generalized_wishart` models the observed distance matrix as arising from a
generalized Wishart distribution (McCullagh 2009, Peterson et al. 2019). It
requires `nu`, the number of genetic markers used to compute `S`, and is the
principled choice when that count is available. Like `mlpe` it accepts a
pairwise distance matrix (e.g. F~ST~).

`wishart_covariance` is the Wishart analogue for data supplied as a
**covariance** matrix (e.g. from `cov_from_biallelic()`) rather than a
distance matrix. It also requires `nu`.

### Fitting with `generalized_wishart`

```{r fit-wishart}
# nu is the number of genetic markers used to calculate melip.Fst.
# 8 in the original data set
fit_gw <- terradish(
  melip.Fst ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = generalized_wishart,
  nu                = 8
)

summary(fit_gw)
```

The nuisance parameters for `generalized_wishart` are `tau` (the scaling of
the resistance-distance covariance) and `sigma` (a log-scale identity
component that absorbs sampling noise not explained by landscape resistance).
A `tau` estimate near zero indicates no detectable resistance-distance signal.

You can compare models fitted with different measurement models using
`aic_table()`. Note that AIC is only meaningful across measurement models when
the response data `S` is identical and the likelihoods are on the same scale;
all three models here use the same `melip.Fst` distance matrix, so the
comparison is valid.

```{r compare-measurement-models}
aic_table(
  list(fit_ls, fit_mlpe, fit_gw),
  mod_names = c("leastsquares", "mlpe", "generalized_wishart")
)
```

---

## Extract and interpret coefficients

```{r coef}
coef(fit_mlpe)
```

The log-linear conductance model means:

- ╬Ė~forestcover~ > 0: areas with more forest are *easier* to move through
  (higher conductance, lower resistance).
- ╬Ė~altitude~ < 0: higher altitude areas are *barriers* to movement.

You can convert back from the log scale to a relative conductance ratio. For
example, a one standard-deviation increase in forest cover multiplies
conductance by exp(╬Ė~forestcover~):

```{r interpret}
theta <- coef(fit_mlpe)
cat("One SD increase in forest cover multiplies conductance by",
    round(exp(theta["forestcover"]), 2), "\n")
cat("One SD increase in altitude multiplies conductance by",
    round(exp(theta["altitude"]), 2), "\n")
```

---

## Visualize the fit

### Observed vs. fitted genetic distances

The `plot()` S3 method with `type = "fit"` produces an observed-vs-fitted
scatter plot with a least-squares regression line:

```{r plot-fit, fig.cap = "***Observed vs. resistance-distance-fitted genetic distances.***"}
plot(fit_mlpe, type = "fit")
```

### Fitted conductance surface

`type = "surface"` renders the point-estimate conductance surface alongside its
lower and upper confidence interval bounds in a single three-panel figure. All
panels share the same color scale so relative differences are easy to read:

```{r conductance-surface, fig.cap = "***Estimated conductance surface with 95% confidence interval bounds.***"}
plot(fit_mlpe, type = "surface", data = surface)
```

Brighter cells are more conductive (easier to move through). Panels where the
lower and upper bounds are close together indicate regions where the available
genetic data tightly constrain conductance, typically near sampling sites.

### Marginal effects of each covariate on predicted genetic distance (default)

The default plot (`type = "marginal_response"`) shows how each raster
covariate influences the **predicted genetic distance** (the response
variable itself) rather than the intermediate conductance scale. Because
genetic distance is what you measured and what the model ultimately explains,
this is usually the most directly interpretable view.

```{r marginal-effects, fig.cap = "***Marginal effect of each covariate on predicted genetic distance with predictive bands.***"}
plot(fit_mlpe, data = surface)
```

**How to read these plots**

Each panel sweeps one covariate across its observed range while all other
covariates are kept at their *observed* values at every grid cell (not at a
single mean value). The curve is therefore a weighted average over the actual
landscape, which matters most when the effects of different covariates
interact.

The **central curve** is the expected genetic distance at the given covariate
value, marginalized over the rest of the landscape.

The **shaded predictive band** combines three sources of uncertainty:

1. *Theta (conductance parameter) uncertainty*: imprecision in the estimated
   conductance parameters ╬Ė, propagated through the resistance-distance
   calculation via the delta method.
2. *Phi (measurement-model parameter) uncertainty*: imprecision in the
   nuisance parameters (intercept ╬▒, IBR slope ╬▓, residual precision Žä, and
   MLPE correlation Žü), again via the delta method.
3. *Residual variation*: the irreducible scatter in pairwise genetic
   distances around the fitted mean, estimated from Žä.

A **rising curve** (genetic distance increases with the covariate) means that
higher covariate values make the landscape *less permeable*, increasing
resistance distance and hence genetic differentiation. A **declining curve**
means the covariate promotes connectivity.

> **Note:** Quadratic terms (`I(x^2)`) and interactions (`x * z`) in the
> formula are handled automatically. For an interaction, each panel holds the
> other covariate at its observed values across all grid cells, so the curve
> represents the average effect of the focal covariate over the joint
> distribution of the landscape.

### Marginal effects on the conductance scale

If you prefer to see the effect on the underlying conductance surface (without
residual variation in the band, and on the conductance rather than genetic
distance scale), use `type = "marginal"`:

```{r marginal-conductance, fig.cap = "***Marginal effect of each covariate on conductance with 95% CI.***"}
plot(fit_mlpe, type = "marginal", data = surface)
```

Here the band is a **confidence interval** (delta-method, 95% pointwise) on
the mean conductance, not a predictive interval; it reflects only parameter
uncertainty in ╬Ė, not residual variation. A rising curve means higher
covariate values increase conductance (easier movement); a declining curve
indicates a barrier effect.

### Support-constrained prediction and plotting

If a covariate has sparse focal support at one tail, marginal curves and
conductance maps can be dominated by extrapolation. `terradish` now supports
focal-support clamping directly in both `plot()` and `conductance()`:

```{r support-clamp, eval = FALSE}
# Clamp selected covariates to focal-site quantile support before plotting.
plot(
  fit_mlpe,
  type = "marginal",
  data = surface,
  support = "focal",
  support_probs = c(0.01, 0.99),
  clamp_covariates = c("altitude", "forestcover")
)

# Apply the same support rule to raster conductance prediction.
cond_focal <- conductance(
  surface,
  fit_mlpe,
  support = "focal",
  support_probs = c(0.01, 0.99),
  clamp_covariates = c("altitude", "forestcover")
)
```

Use `support = "none"` (default) for full-domain evaluation, and
`support = "focal"` when you want inference constrained to focal sampling
support.

---

## What comes next?

This vignette covered the core workflow. Three companion vignettes go further:

- **`vignette("model-comparison", package = "terradish")`** ŌĆö testing and
  comparing models (likelihood ratio tests, AIC, cross-validation, likelihood
  surface visualization).
- **`vignette("ibe-ibr-workflow", package = "terradish")`** ŌĆö fitting
  isolation-by-environment (IBE) and IBR simultaneously in the same model.
- **`vignette("wishart-covariance", package = "terradish")`** ŌĆö preparing a
  genetic covariance matrix from raw genotype data with `cov_from_genetic_data()`
  and fitting with the `wishart_covariance` measurement model.

---

## Quick-reference: the full workflow in one block

```{r full-workflow, eval = FALSE}
library(terradish)
library(terra)

# 1. Load data
data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

# 2. Scale covariates and build surface
covariates <- c(melip.forestcover, melip.altitude)
names(covariates) <- c("forestcover", "altitude")
covariates <- scale_covariates(covariates)

surface <- conductance_surface(covariates, melip.coords,
                               directions = 8, saveStack = TRUE)

# 2b. (Optional) Profile graph and identify fastest settings
#     Most useful for large graphs (> 10 000 vertices) or spline models.
assessment <- terradish_assess_settings(
  melip.Fst ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  probe_maxit       = 2
)
cat(assessment$comparison$summary, "\n")  # one-line verdict
rec <- assessment$recommended             # reuse for all fits below

# 3. Fit model ŌĆö choose a measurement model:
#   leastsquares      : fast, independent errors (distance matrix)
#   mlpe              : shared-site correlation correction (distance matrix)
#   generalized_wishart: Wishart model, requires nu (distance matrix)
#   wishart_covariance : Wishart model, requires nu (covariance matrix)
fit <- terradish(melip.Fst ~ forestcover + altitude,
                 data              = surface,
                 conductance_model = loglinear_conductance,
                 measurement_model = mlpe,
                 optimizer         = rec$optimizer,
                 control           = rec$control,
                 solver            = rec$solver,
                 solver_control    = rec$solver_control,
                 approximation     = rec$approximation,
                 approximation_control = rec$approximation_control)

# generalized_wishart example (set nu to your actual marker count):
fit_gw <- terradish(melip.Fst ~ forestcover + altitude,
                    data              = surface,
                    conductance_model = loglinear_conductance,
                    measurement_model = generalized_wishart,
                    nu                = 1000,
                    optimizer         = rec$optimizer,
                    control           = rec$control,
                    solver            = rec$solver,
                    solver_control    = rec$solver_control,
                    approximation     = rec$approximation,
                    approximation_control = rec$approximation_control)

# 4. Summarize and visualize
summary(fit)

plot(fit, type = "fit")                                # observed vs. fitted scatter
plot(fit, type = "surface", data = surface)            # conductance map + 95% CI panels
plot(fit, data = surface)                              # marginal effects on genetic distance (default)
plot(fit, type = "marginal", data = surface)           # marginal effects on conductance scale

# 5. Optional: focal-support clamping for robust tails
plot(fit, type = "marginal", data = surface,
     support = "focal", support_probs = c(0.01, 0.99),
     clamp_covariates = c("forestcover", "altitude"))

cond <- conductance(surface, fit,
                    support = "focal", support_probs = c(0.01, 0.99),
                    clamp_covariates = c("forestcover", "altitude"))
```
