---
title: "Getting Started with terradish"
author: "Bill Peterman"
output: 
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting Started with terradish}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4.5,
  warning  = FALSE,
  message  = FALSE
)
```

## What is terradish for?

`terradish` estimates how landscape features — altitude, forest cover, roads,
water bodies, etc. — shape genetic differentiation among populations.

The core idea is **isolation by resistance (IBR)**: instead of assuming that
genetic distance simply tracks straight-line geographic distance (isolation by
distance, IBD), IBR recognises that the landscape is heterogeneous. Some cells
are easy to move through; others are barriers. The ease of movement through a
cell is its **conductance** (the inverse of **resistance**).

If we can map conductance across the landscape as a function of measurable
covariates (rasters), we can:

1. Build a **conductance surface** — a weighted graph over the raster grid.
2. Compute **resistance distances** among sampling locations — effective
   distances that account for the landscape structure.
3. Fit a **statistical model** relating those resistance distances to observed
   genetic distances.
4. Optimise the conductance parameters by maximum likelihood.

`terradish` automates all four steps and provides gradients, standard errors,
model-comparison tools, and visualisation helpers.

---

## A conceptual road map

Before writing any code, it helps to see how the pieces fit together:

```
Raster covariates          Sampling locations
(altitude, forestcover)    (GPS coordinates)
        |                          |
        +-----> conductance_surface() <----+
                        |
                        |   (builds the weighted graph)
                        v
              terradish(formula, surface,
                        conductance_model,  <-- how covariates -> conductance
                        measurement_model)  <-- how resistance dist -> genetic dist
                        |
                        v
                   fitted model
                   coef(), summary(), fitted(),
                   conductance(), anova()
```

The **conductance model** maps raster values at each grid cell to a conductance
value. `loglinear_conductance` is the default: conductance =
exp(θ₁·altitude + θ₂·forestcover + ...). The parameters θ are what we
estimate.

The **measurement model** describes the statistical relationship between
pairwise resistance distances and pairwise genetic distances. Two choices cover
most use-cases:

- `leastsquares` — ordinary regression; fast, assumes independent errors.
- `mlpe` — "maximum likelihood population effects"; accounts for the
  non-independence among all pairwise measurements that share a sampling site.
  This is generally the preferred choice.

---

## Installation

```{r, eval = FALSE}
# Install from local source (recommended while the package is in development)
devtools::install()
# or
remotes::install_local(".")

# corMLPE is also required (not on CRAN)
devtools::install_github("nspope/corMLPE")
```

---

## Load the package and example data

```{r load}
library(terradish)
library(terra)
```

`terradish` ships with example data from a study of *Melipona* bees in
Mexico. Load everything with one call:

```{r data}
data(melip)

# The rasters are stored as "packed" objects to comply with CRAN policy;
# unwrap() converts them back to usable SpatRaster / SpatVector objects.
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)
```

What do we have?

- **`melip.altitude`** — a raster of elevation (metres) across the study area.
- **`melip.forestcover`** — a raster of forest cover proportion (0–1).
- **`melip.coords`** — 34 sampling locations as a `SpatVector`.
- **`melip.Fst`** — a 34 × 34 matrix of pairwise F~ST~ (a standard measure
  of genetic differentiation; larger values mean more differentiation).

```{r inspect-data}
# How many sampling sites?
nrow(melip.coords)

# Quick look at the genetic distance matrix
# (only the first 5 x 5 block for readability)
round(melip.Fst[1:5, 1:5], 3)
```

F~ST~ is symmetric (the differentiation between site A and site B is the same
as between B and A), so only the lower triangle carries new information.

---

## Visualise the spatial data

```{r plot-covariates, fig.cap = "Altitude raster with sampling sites."}
plot(melip.altitude, main = "Altitude (m)")
points(melip.coords, pch = 19, cex = 0.8)
```

```{r plot-forestcover, fig.cap = "Forest cover raster (0 = no forest, 1 = full cover)."}
plot(melip.forestcover, main = "Forest cover")
points(melip.coords, pch = 19, cex = 0.8)
```

---

## Scale your rasters — an important step

Conductance in the loglinear model is exp(θ·x). If `x` (e.g. altitude in
metres) has a very large range, small changes in θ produce enormous changes in
conductance, making optimisation numerically unstable. **Always scale continuous
covariates** before fitting.

```{r scale}
covariates <- c(terra::scale(melip.altitude),
                terra::scale(melip.forestcover))
names(covariates) <- c("altitude", "forestcover")
```

`terra::scale()` subtracts the mean and divides by the standard deviation,
giving each layer a mean of 0 and a standard deviation of 1.

---

## Build the conductance surface

```{r surface}
surface <- conductance_surface(covariates, melip.coords, directions = 8)
```

`conductance_surface()` does three things:

1. Reads the raster values at every non-NA grid cell.
2. Identifies which cells correspond to the sampling locations.
3. Builds a graph where adjacent cells (here, using all 8 neighbours including
   diagonals) are connected by edges whose weights will be determined by the
   conductance model.

`directions = 8` is typical for landscape genetics; it allows diagonal movement
and is less sensitive to grid orientation than `directions = 4`.

```{r surface-check}
# How many grid cells are in the graph?
nrow(surface$x)

# The covariate data for the first few cells
head(surface$x)
```

---

## Fit your first model

We will start with the simplest case: do both altitude and forest cover predict
genetic differentiation?

```{r fit-leastsquares}
fit_ls <- terradish(
  melip.Fst ~ forestcover + altitude,  # formula: response ~ covariates
  data              = surface,          # the conductance surface
  conductance_model = loglinear_conductance,
  measurement_model = leastsquares      # ordinary regression
)
```

The formula works like a standard R formula:

- The **left-hand side** is the matrix of observed genetic distances.
- The **right-hand side** lists which raster layers drive conductance.

`leastsquares` is the quickest measurement model. It treats the pairwise
genetic distances as independent observations regressed on resistance distances.
That is not strictly correct (measurements sharing a site are correlated), but
it gives a useful quick answer.

```{r summary-ls}
summary(fit_ls)
```

**How to read the output:**

- **theta**: the conductance parameters. A positive θ for a covariate means
  that higher values of that covariate increase conductance (i.e., make movement
  easier), which *decreases* resistance distance. A negative θ means the
  covariate acts as a barrier.
- **phi**: nuisance parameters of the measurement model (intercept, slope,
  residual precision). These are profiled out automatically — you generally do
  not need to interpret them directly.
- The standard errors and z-statistics help gauge parameter uncertainty.

---

## Upgrade to the MLPE measurement model

Pairwise genetic distances are not independent: the distance between sites A–B
and A–C both involve site A. `mlpe` ("maximum likelihood population effects")
accounts for this correlation structure and is the recommended choice for
inference.

```{r fit-mlpe}
fit_mlpe <- terradish(
  melip.Fst ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe
)

summary(fit_mlpe)
```

Compare the point estimates and standard errors between `fit_ls` and
`fit_mlpe`. They are often similar, but MLPE standard errors better reflect the
actual precision of your estimates.

---

## Extract and interpret coefficients

```{r coef}
coef(fit_mlpe)
```

The log-linear conductance model means:

- θ~forestcover~ > 0: areas with more forest are *easier* to move through
  (higher conductance, lower resistance).
- θ~altitude~ < 0: higher altitude areas are *barriers* to movement.

You can convert back from the log scale to a relative conductance ratio. For
example, a one standard-deviation increase in forest cover multiplies
conductance by exp(θ~forestcover~):

```{r interpret}
theta <- coef(fit_mlpe)
cat("One SD increase in forest cover multiplies conductance by",
    round(exp(theta["forestcover"]), 2), "\n")
cat("One SD increase in altitude multiplies conductance by",
    round(exp(theta["altitude"]), 2), "\n")
```

---

## Visualise the fit

### Observed vs. fitted genetic distances

```{r plot-fit, fig.cap = "Observed vs. resistance-distance-fitted genetic distances."}
plot(
  fitted(fit_mlpe, "distance"),
  melip.Fst,
  pch  = 19,
  cex  = 0.5,
  xlab = "Optimised resistance distance",
  ylab = expression(F[ST]),
  main = "Observed vs. fitted"
)
abline(lm(c(melip.Fst) ~ c(fitted(fit_mlpe, "distance"))),
       col = "steelblue", lwd = 2)
```

### Fitted conductance surface

```{r conductance-surface, fig.cap = "Estimated conductance surface with 95% confidence interval bounds."}
fitted_cond <- conductance(surface, fit_mlpe, quantile = 0.95)

# Estimated conductance
plot(fitted_cond[["est"]],
     main = "Fitted conductance (MLPE)")
points(melip.coords, pch = 19, cex = 0.6)

# Lower 95% confidence interval
plot(fitted_cond[["lower95"]],
     main = "Lower 95% CI — conductance")

# Upper 95% confidence interval
plot(fitted_cond[["upper95"]],
     main = "Upper 95% CI — conductance")
```

Darker cells are more conductive (easier to move through). The confidence
interval layers show where estimates are more or less certain — typically,
areas with many sampling sites nearby have narrower intervals.

---

## What comes next?

This vignette covered the core workflow. Two companion vignettes go further:

- **`vignette("model-comparison", package = "terradish")`** — testing and
  comparing models (likelihood ratio tests, AIC, cross-validation, likelihood
  surface visualisation).
- **`vignette("ibe-ibr-workflow", package = "terradish")`** — fitting
  isolation-by-environment (IBE) and IBR simultaneously in the same model.

---

## Quick-reference: the full workflow in one block

```{r full-workflow, eval = FALSE}
library(terradish)
library(terra)

# 1. Load data
data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

# 2. Scale covariates and build surface
covariates <- c(terra::scale(melip.altitude),
                terra::scale(melip.forestcover))
names(covariates) <- c("altitude", "forestcover")
surface <- conductance_surface(covariates, melip.coords, directions = 8)

# 3. Fit model
fit <- terradish(melip.Fst ~ forestcover + altitude,
                 data              = surface,
                 conductance_model = loglinear_conductance,
                 measurement_model = mlpe)

# 4. Summarise and visualise
summary(fit)
plot(fitted(fit, "distance"), melip.Fst, pch = 19,
     xlab = "Resistance distance", ylab = expression(F[ST]))
plot(conductance(surface, fit, quantile = 0.95)[["est"]])
```
