---
title: "Hierarchical conductance: covariates plus a smooth field"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Hierarchical conductance: covariates plus a smooth field}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4.5,
  warning  = FALSE,
  message  = FALSE
)
# Results are pre-computed (model fitting is too slow for a CRAN vignette build).
# The chunk below loads a saved object; the fitting code is shown with eval=FALSE.
res <- tryCatch(
  readRDS("vignette-hierarchical.rds"),
  error = function(e) NULL
)
have_res <- !is.null(res)
```

## Why add a field to a covariate model?

A covariate conductance model (`terradish` with `loglinear_conductance`) is
mechanistic and predictive, but it is blind to any process its covariates do not
encode. If an important driver of gene flow is unmapped, the optimizer does not
return a clean null; it bends the *mapped* covariates to soak up the missing
structure, biasing their estimated effects. This is the omitted-variable problem
documented for correlative resistance estimation (Graves et al. 2013; Daniel et
al. 2024).

`terradish_hierarchical()` addresses this by writing log-conductance as a
mechanistic covariate term **plus a spatially smooth Gaussian random field**:

$$\log c = X\theta + u, \qquad u \sim \mathrm{Normal}\!\big(0,\ \tau^2 (L_{coarse} + \epsilon I)^{-1}\big).$$

The GMRF prior on $u$ is exactly the FEEMS spatial-smoothness penalty written as
a proper distribution: it pulls neighbouring field values together, with $\tau^2$
controlling how much structure the field may absorb. This single model nests both
parents:

- set the field to zero ($\tau^2 \to 0$) and you recover **`terradish`**;
- drop the covariates and let $u$ carry the whole surface and you recover a
  **FEEMS-style** free, smoothed surface;
- keep both and you get a model that is mechanistic where covariates explain the
  data and flexible where they do not.

Two payoffs follow. First, the covariate effects $\theta$ are **protected from
omitted-variable bias**, because unmodelled spatial structure is absorbed by $u$
rather than distorting $\theta$. Second, the fitted field $u$ is **a deliverable
in its own right**: a map of the conductance variation your covariates failed to
explain, and therefore a residual diagnostic and a guide to the next covariate
worth collecting.

## How it is fit

The penalized objective is

$$J(\theta, u) = -\ell(\theta, u \mid \phi) + \tfrac{1}{2\tau^2}\, u^\top (L_{coarse}+\epsilon I)\, u,$$

where $\ell$ is the measurement-model log-likelihood (the nuisance parameters
$\phi$ profiled out) and $L_{coarse}$ is the Laplacian of the coarse-field
adjacency graph. The field lives on a coarse `field_resolution` $\times$
`field_resolution` grid over the landscape — deliberately coarser than the
covariate grid, which is the main safeguard against $\theta$ and $u$ competing to
explain the same signal.

Optimization is by L-BFGS over $(\theta, u)$ using only the gradient, which flows
through the same single sparse-Cholesky solve `terradish` already uses; the dense
Hessian over the field is never formed. The field variance $\tau^2$ is chosen by
a Laplace (empirical-Bayes) marginal likelihood that integrates the field out.

## A worked example

We use the bundled *Melipona* landscape and **simulate** a covariance response
whose true conductance depends on altitude and forest cover **plus an unmapped
smooth feature** (a localized corridor/barrier the analyst did not measure).

```{r build-surface, eval = FALSE}
library(terradish)
library(terra)

data(melip)
covs <- c(terra::scale(terra::unwrap(melip.altitude)),
          terra::scale(terra::unwrap(melip.forestcover)))
names(covs) <- c("altitude", "forestcover")
surface <- conductance_surface(covs, terra::unwrap(melip.coords), directions = 8)
```

```{r simulate, eval = FALSE}
# True log-conductance = 0.5*altitude - 0.6*forestcover + an UNMAPPED blob.
Vc <- surface$vertex_coordinates
xr <- range(Vc[, 1]); yr <- range(Vc[, 2])
blob <- 1.6 * exp(-(((Vc[, 1] - (xr[1] + 0.35 * diff(xr)))^2 +
                     (Vc[, 2] - (yr[1] + 0.6 * diff(yr)))^2) /
                    (2 * (0.18 * sqrt(diff(xr)^2 + diff(yr)^2))^2)))
blob <- blob - mean(blob)
# (helper builds a conductance model from a raw design matrix; see ?terradish)
S <- ... # cov_from_genetic_data() on real genotypes; here a simulated Wishart draw
```

Fit the hierarchical model, letting $\tau^2$ be chosen by REML, and a plain
`terradish` model for comparison:

```{r fit, eval = FALSE}
fit_h <- terradish_hierarchical(
  S ~ altitude + forestcover,
  data              = surface,
  measurement_model = wishart_covariance,
  nu                = 1500,
  field_resolution  = 6,
  tau2              = "reml"
)

fit_plain <- terradish(
  S ~ altitude + forestcover, data = surface,
  conductance_model = loglinear_conductance,
  measurement_model = wishart_covariance, nu = 1500
)
```

### The field absorbs the unmapped feature, protecting `theta`

```{r show-coef, echo = FALSE, eval = have_res}
knitr::kable(res$coef_table, digits = 3,
             caption = "Conductance coefficients (true: altitude 0.50, forestcover -0.60)")
```

```{r coef-text, eval = FALSE}
# With the field: altitude ~0.54, forestcover ~-0.63 (close to truth 0.50, -0.60)
coef(fit_h)
# Without the field: forestcover ~-0.43 -- biased ~30% toward zero by the
# unmodelled blob.
coef(fit_plain)
```

```{r recovery-text, echo = FALSE, results = "asis", eval = have_res}
cat(sprintf(
  "The fitted field recovers the unmapped feature with correlation **%.2f** to the true blob, and the selected field variance is **tau^2 = %.3g**.\n",
  res$cor_field_blob, res$tau2))
```

The plain model, missing the blob, biases the forest-cover effect toward zero;
the hierarchical model recovers it because the field — not `theta` — absorbs the
unmodelled structure.

### Mapping the unexplained conductance

`conductance_field()` returns the field as a raster (when the graph stored its
stack), so you can see *where* the covariates fell short:

```{r field-map, eval = FALSE}
field_raster <- conductance_field(fit_h, surface, type = "field")
plot(field_raster)            # the smooth residual conductance surface u

# Or the full fitted surface, exp(X theta + u):
plot(conductance_field(fit_h, surface, type = "conductance"))
```

```{r field-plot, echo = FALSE, eval = have_res, fig.cap = "Fitted smooth field u: the conductance variation the covariates did not explain. The peak coincides with the unmapped feature."}
if (!is.null(res$field_raster)) {
  terra::plot(terra::unwrap(res$field_raster), main = "Fitted field (u)")
}
```

## Selecting `tau2`

With `tau2 = "reml"` (the default), `terradish_hierarchical()` evaluates a
Laplace marginal likelihood across `tau2_grid` and reports the profile:

```{r tau2-table, echo = FALSE, eval = have_res}
knitr::kable(res$tau2_selection, digits = 3,
             caption = "Laplace marginal log-likelihood across the tau^2 grid (interior maximum).")
```

A clear interior maximum (rather than a value pinned at the smallest or largest
grid point) indicates the data support a field of intermediate smoothness. If the
maximum sits at a grid boundary, widen `tau2_grid`. You can also pass a fixed
numeric `tau2` if you want to control the amount of flexibility directly, or set
it from external information.

## Practical guidance

- **Keep the field coarser than your covariates.** `field_resolution` controls
  this; a field as fine as the covariate grid can mimic covariate gradients and
  erode identifiability. Start coarse (the default `6`) and increase only if a
  clear, stable feature demands it.
- **Read the field as a diagnostic, not a mechanism.** A strong field feature
  says "something here is not in your covariates" — a hypothesis generator for
  the next layer to collect, not an explanation by itself.
- **Genetic data constrain conductance weakly.** Added flexibility is only
  helpful if paid for by an informative prior; that is exactly what the GMRF on
  the (coarse) field provides. Sanity-check that `theta` and predictive accuracy
  are stable across a couple of `field_resolution` values.
- **It nests your simpler models.** Fitting `terradish_hierarchical()` and
  comparing the field magnitude (and `theta`) against a plain `terradish` fit is
  a direct check on whether omitted spatial structure is distorting your
  covariate effects.

## Quick-reference

```{r quickref, eval = FALSE}
library(terradish)

# 1. Build the conductance surface (as for any terradish workflow)
surface <- conductance_surface(covariates, coords, directions = 8)

# 2. Fit the hierarchical model (covariates + smooth field), REML tau^2
fit <- terradish_hierarchical(
  S ~ forestcover + altitude,
  data              = surface,
  measurement_model = wishart_covariance,   # or generalized_wishart / mlpe
  nu                = nu,
  field_resolution  = 6,
  tau2              = "reml"
)

# 3. Inspect mechanism + the unexplained-conductance map
summary(fit)                                  # theta (protected) + tau^2 profile
plot(conductance_field(fit, surface))         # the field u
```
