---
title: "Hierarchical conductance: covariates plus a smooth field"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Hierarchical conductance: covariates plus a smooth field}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4.5,
  warning  = FALSE,
  message  = FALSE
)
# Results are pre-computed (model fitting is too slow for a CRAN vignette build).
# The chunk below loads a saved object; the fitting code is shown with eval=FALSE.
res <- tryCatch(
  readRDS("vignette-hierarchical.rds"),
  error = function(e) NULL
)
have_res <- !is.null(res)
```

## Why add a field to a covariate model?

A covariate conductance model (`terradish` with `loglinear_conductance`) can leave
spatial structure that the supplied covariates and transformations do not
represent. That residual pattern can reflect an omitted covariate, an incorrect
functional form, spatial confounding, sampling structure, historical signal, or
measurement-model mismatch.

`terradish_hierarchical()` represents log-conductance as a covariate term **plus
a spatially smooth Gaussian random field**:

$$\log c = X\theta + u, \qquad u \sim \mathrm{Normal}\!\big(0,\ \tau^2 (L_{coarse} + \epsilon I)^{-1}\big).$$

The proper GMRF prior pulls neighboring field values together, with $\tau^2$
controlling how much residual structure the field may absorb. This model includes
two limiting cases:

- set the field to zero ($\tau^2 \to 0$) and you recover **`terradish`**;
- drop the covariates and let $u$ carry the whole surface to obtain a free,
  smoothed field;
- keep both to estimate covariate coefficients conditional on a flexible
  residual spatial term.

The fitted field $u$ is primarily a **diagnostic**: it maps residual
log-conductance structure after the supplied covariates. The field may reduce
some omitted spatial pattern in a controlled setting, but it can also compete
with spatially structured covariates. It does not guarantee unbiased $\theta$,
identify a missing mechanism, or establish causality. Interpret both $\theta$
and $u$ conditionally and examine sensitivity to field resolution and $\tau^2$.

## How it is fit

The penalized objective is

$$J(\theta, u) = -\ell(\theta, u \mid \phi) + \tfrac{1}{2\tau^2}\, u^\top (L_{coarse}+\epsilon I)\, u,$$

where $\ell$ is the measurement-model log-likelihood (the nuisance parameters
$\phi$ profiled out) and $L_{coarse}$ is the Laplacian of the coarse-field
adjacency graph. The field lives on a coarse `field_resolution` $\times$
`field_resolution` grid over the landscape, deliberately coarser than the
covariate grid. This reduces, but does not eliminate, competition between
$\theta$ and $u$ for the same spatial pattern.

Optimization is by L-BFGS over $(\theta, u)$ using only the gradient, which flows
through the same single sparse-Cholesky solve `terradish` already uses; the dense
Hessian over the field is never formed. The field variance $\tau^2$ is chosen by
a Laplace (empirical-Bayes) marginal likelihood that integrates the field out.

## A worked example

We use the bundled *Melipona* landscape and **simulate** a covariance response
whose generating conductance depends on altitude and forest cover **plus an
omitted smooth feature**. This deliberately favorable simulation illustrates
the method under a known model; it is not a general validation of field or
coefficient recovery.

```{r build-surface, eval = FALSE}
library(terradish)
library(terra)

data(melip)
covs <- c(terra::scale(terra::unwrap(melip.altitude)),
          terra::scale(terra::unwrap(melip.forestcover)))
names(covs) <- c("altitude", "forestcover")
surface <- conductance_surface(covs, terra::unwrap(melip.coords), directions = 8)
```

```{r simulate, eval = FALSE}
set.seed(20)

n  <- length(surface$demes)                       # 37 focal sites
Vc <- surface$vertex_coordinates                  # cell centers of the graph
X  <- as.matrix(surface$x[, c("altitude", "forestcover")])

# The UNMAPPED feature: a smooth Gaussian bump in log-conductance that no
# raster in the model encodes. Centering it keeps it out of the intercept.
xr <- range(Vc[, 1]); yr <- range(Vc[, 2])
blob <- 1.6 * exp(-(((Vc[, 1] - (xr[1] + 0.35 * diff(xr)))^2 +
                     (Vc[, 2] - (yr[1] + 0.6 * diff(yr)))^2) /
                    (2 * (0.18 * sqrt(diff(xr)^2 + diff(yr)^2))^2)))
blob <- blob - mean(blob)

# True log-conductance = 0.5*altitude - 0.6*forestcover + 1.0*blob.
# .design_loglinear_model() builds a conductance model straight from a design
# matrix, which lets us put the blob in the truth without giving it to the fit.
E <- as.matrix(terradish_algorithm(
  terradish:::.design_loglinear_model(cbind(X, blob)),
  leastsquares, surface, S = diag(n), theta = c(0.5, -0.6, 1.0),
  objective = FALSE, gradient = FALSE, hessian = FALSE,
  partial = FALSE)$covariance)

# Draw a genetic covariance matrix from that truth. With real data this is
# where cov_from_genetic_data() or cov_from_biallelic() would supply S, and nu
# would be your effective marker count.
nu <- 1500
S  <- rWishart(1, df = nu, Sigma = (E + 0.25 * diag(n)) / nu)[, , 1]
```

Fit the hierarchical model, letting $\tau^2$ be chosen by REML, and a plain
`terradish` model for comparison:

```{r fit, eval = FALSE}
fit_h <- terradish_hierarchical(
  S ~ altitude + forestcover,
  data              = surface,
  measurement_model = wishart_covariance,
  nu                = 1500,
  field_resolution  = 6,
  tau2              = "reml"
)

fit_plain <- terradish(
  S ~ altitude + forestcover, data = surface,
  conductance_model = loglinear_conductance,
  measurement_model = wishart_covariance, nu = 1500
)
```

### Behavior in this controlled simulation

```{r show-coef, echo = FALSE, eval = have_res}
knitr::kable(res$coef_table, digits = 3,
             caption = "Conductance coefficients (true: altitude 0.50, forestcover -0.60)")
```

```{r coef-text, eval = FALSE}
# With the field: altitude ~0.54, forestcover ~-0.63 (close to truth 0.50, -0.60)
coef(fit_h)
# Without the field: forestcover ~-0.43 in this particular simulation.
coef(fit_plain)
```

```{r recovery-text, echo = FALSE, results = "asis", eval = have_res}
cat(sprintf(
  "In this controlled simulation, the fitted field correlates **%.2f** with the generating blob, and the selected field variance is **tau^2 = %.3g**.\n",
  res$cor_field_blob, res$tau2))
```

In this one simulation, the plain model's forest-cover coefficient is closer to
zero and the hierarchical coefficient is closer to the generating value. That
contrast demonstrates possible behavior under a deliberately matched setting.
It does not show that a field generally removes omitted-variable bias.

### Mapping fitted residual log-conductance

`conductance_field()` returns the field as a raster (when the graph stored its
stack), so you can inspect where residual structure remains under the fitted
model:

```{r field-map, eval = FALSE}
field_raster <- conductance_field(fit_h, surface, type = "field")
plot(field_raster)            # the smooth residual conductance surface u

# Or the full fitted surface, exp(X theta + u):
plot(conductance_field(fit_h, surface, type = "conductance"))
```

```{r field-plot, echo = FALSE, eval = have_res, fig.cap = "Fitted smooth residual field u in the controlled simulation. The peak overlaps the omitted generating feature."}
if (!is.null(res$field_raster)) {
  terra::plot(terra::unwrap(res$field_raster), main = "Fitted field (u)")
}
```

## Selecting `tau2`

With `tau2 = "reml"` (the default), `terradish_hierarchical()` evaluates a
Laplace marginal likelihood across `tau2_grid` and reports the profile:

```{r tau2-table, echo = FALSE, eval = have_res}
knitr::kable(res$tau2_selection, digits = 3,
             caption = "Laplace marginal log-likelihood across the tau^2 grid (interior maximum).")
```

A clear interior maximum (rather than a value pinned at the smallest or largest
grid point) indicates that an intermediate field variance maximizes this
empirical-Bayes profile. It does not identify a unique biological spatial
scale. If the maximum sits at a grid boundary, widen `tau2_grid`. You can also
pass a fixed numeric `tau2` if you want to control the amount of flexibility
directly, or set it from external information.

## Practical guidance

- **Keep the field coarser than your covariates.** `field_resolution` controls
  this; a field as fine as the covariate grid can mimic covariate gradients and
  erode identifiability. Start coarse (the default `6`) and increase only if a
  clear, stable feature demands it.
- **Read the field as a diagnostic, not a mechanism.** A strong field feature
  indicates residual structure under the fitted model. It is a hypothesis
  generator, not an explanation by itself.
- **Genetic data constrain conductance weakly.** Added flexibility is only
  helpful if paid for by an informative prior; that is exactly what the GMRF on
  the (coarse) field provides. Sanity-check that `theta` and predictive accuracy
  are stable across a couple of `field_resolution` values.
- **Compare sensitivity, not just one fit.** Compare `theta`, the field, and
  predictive behavior across a plain fit, several `field_resolution` values,
  and a defensible `tau2_grid`. Stability is more informative than any single
  residual map.

## Quick-reference

```{r quickref, eval = FALSE}
library(terradish)

# 1. Build the conductance surface (as for any terradish workflow)
surface <- conductance_surface(covariates, coords, directions = 8)

# 2. Fit the hierarchical model (covariates + smooth field), REML tau^2
fit <- terradish_hierarchical(
  S ~ forestcover + altitude,
  data              = surface,
  measurement_model = wishart_covariance,   # or generalized_wishart / mlpe
  nu                = nu,
  field_resolution  = 6,
  tau2              = "reml"
)

# 3. Inspect conditional coefficients and the residual-field diagnostic
summary(fit)                                  # theta + tau^2 profile
plot(conductance_field(fit, surface))         # the field u
```

## Summary of key functions

| Function | Purpose |
|----------|---------|
| `conductance_surface()` | Build the weighted graph over the raster grid |
| `terradish_hierarchical()` | Fit `log c = X theta + u`, selecting `tau^2` by REML |
| `summary()` | Covariate table, field sd and edf, and the `tau^2` profile |
| `conductance_field()` | Return the field `u` or the full surface as a `SpatRaster` |
| `plot(type = "field")` | Map the fitted residual log-conductance field |
| `AIC()` | Conditional-likelihood summary that charges the field's edf; use cautiously with sensitivity and predictive checks |

## See also

- **`vignette("getting-started", package = "terradish")`**, the plain
  covariate model this one extends.
- **`vignette("spline-conductance", package = "terradish")`**, added
  flexibility in the covariate response rather than in space.
- **`vignette("wishart-covariance", package = "terradish")`**, the covariance
  measurement model used here and how to choose `nu`.
- **`vignette("model-comparison", package = "terradish")`**, deciding whether
  the field earns the effective degrees of freedom it costs.
