---
title: "Concurrent IBE and IBR in terradish"
author: "Bill Peterman"
output: 
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Concurrent IBE and IBR in terradish}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4.5,
  warning  = FALSE,
  message  = FALSE
)
```

## What are IBE and IBR, and why fit them together?

Landscape geneticists commonly ask: why are some pairs of populations more
genetically different than others? Two broad classes of explanation exist:

- **Isolation by resistance (IBR):** movement is easiest through certain
  landscape types and hardest through others. Genetic differentiation
  accumulates between populations separated by high-resistance terrain,
  even when straight-line distance is short.
- **Isolation by environment (IBE):** populations inhabiting similar
  environments tend to exchange genes more readily (or experience parallel
  selection), so environmentally dissimilar sites show greater genetic
  differentiation regardless of the path connecting them.

Both processes can operate at the same time. If you only test IBR, you may
attribute environmental mismatch to landscape barriers, and vice versa. Fitting
IBE and IBR **concurrently** lets each explain the variation that truly belongs
to it, rather than competing for credit in a single predictor.

### Where each process lives in the model

`terradish` separates the two processes cleanly across its two model layers:

```
                           CONDUCTANCE SIDE
  Raster covariates   -->  loglinear_conductance  -->  resistance distances
  (altitude, forest)         theta parameters               R_ij

                           MEASUREMENT SIDE
  R_ij + endpoint          mlpe_covariates()      -->  genetic distances
  differences (Z_ij)         beta, gamma, rho           S_ij
```

The mean structure of the joint model is:

$$S_{ij} = α + β · R_{ij} + Z_{ij} · γ + e_{ij}$$

where:

- $R_{ij}$ is the pairwise resistance distance (determined by the conductance
  surface and the θ parameters).
- $Z_{ij}$ is a pairwise endpoint-difference covariate (some measure of how
  environmentally dissimilar sites *i* and *j* are).
- $β$ captures the IBR signal: does resistance distance predict genetic
  differentiation after accounting for environment?
- **γ** captures the IBE signal: does environmental mismatch predict genetic
  differentiation after accounting for landscape resistance?
- $e_{ij}$ is a residual following the MLPE correlation structure (see
  `vignette("getting-started", package = "terradish")`).

---

## Prerequisites

This vignette assumes familiarity with the basic `terradish` workflow. If you
have not yet read
`vignette("getting-started", package = "terradish")`, start there.

---

## Set up: data and scaled covariates

```{r load}
library(terradish)
library(terra)

data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

# Scale continuous covariates before building the conductance surface.
# This keeps the conductance parameters on a comparable scale and avoids
# numerical overflow in exp(theta * x).
covariates <- c(melip.altitude, melip.forestcover)
names(covariates) <- c("altitude", "forestcover")
covariates <- scale_covariates(covariates)

surface <- conductance_surface(covariates, melip.coords, directions = 8)
```

---

## Step 1 — Construct IBE covariates

IBE operates at the level of pairs of sites. For each pair (i, j) we need a
number that quantifies how environmentally different those two sites are. The
helper `pairwise_endpoint_covariates()` computes this from site-level
environmental values.

```{r ibe-covariates}
z_ibe <- pairwise_endpoint_covariates(
  melip.altitude,          # site-level environmental data (raster or matrix)
  coords    = melip.coords,# focal-point locations (used to extract from raster)
  transform = "absdiff",  # how to convert site values into pairwise values
  scale     = TRUE         # standardize site values first
)

# Each row is one pair of sites; each column is one pairwise covariate.
dim(z_ibe)
head(z_ibe)
```

### Choosing a pairwise transform

Four transforms are available:

| Transform | Formula | When to use |
|-----------|---------|-------------|
| `"absdiff"` | \|x\_i − x\_j\| | One covariate; symmetric, intuitive |
| `"sqdiff"` | (x\_i − x\_j)² | Penalizes large differences more; useful with noisy data |
| `"euclidean"` | √Σ(x\_ik − x\_jk)² | Multivariate environment; collapses to a single distance |
| `"manhattan"` | Σ\|x\_ik − x\_jk\| | Multivariate environment; less sensitive to outlier dimensions |

Use `"absdiff"` (the default) for a single continuous covariate. For a
multivariate environment (e.g. altitude + precipitation + temperature combined),
`"euclidean"` or `"manhattan"` produce a single distance column that is easier
to interpret and avoid inflating the model with many correlated predictors.

For two or more covariates treated separately, `"absdiff"` or `"sqdiff"` return
one column per covariate:

```{r ibe-covariates-multi}
# Example: separate absolute differences for altitude and forest cover
z_multi <- pairwise_endpoint_covariates(
  covariates,            # the two-layer scaled raster stack
  coords    = melip.coords,
  transform = "absdiff",
  scale     = FALSE      # already scaled in the raster
)
dim(z_multi)   # one column per layer
colnames(z_multi)
```

For this vignette we use altitude only (`z_ibe`) to keep the IBE side simple
and interpretable.

---

## Step 2 — Build the joint measurement model

`mlpe_covariates()` wraps the pairwise covariates into an MLPE measurement
model. The resulting object (`g_ibe`) can be passed anywhere that `mlpe` can:
`terradish()`, `terradish_grid()`, and `terradish_cv()`.

```{r build-model}
g_ibe <- mlpe_covariates(z_ibe)
```

`g_ibe` is a function, not a fitted model; it defines the likelihood structure.
The actual parameters (β, γ, and the MLPE correlation) are estimated when you
call `terradish()`.

---

## Step 3 — Fit the four candidate models

To understand the unique contributions of IBR and IBE, fit all four
combinations and compare them.

```{r fit-models}
# NewtonRaphsonControl limits the number of optimizer iterations.
# A small maxit is fine here for illustration; use the default (100) in a
# real analysis.
ctrl <- NewtonRaphsonControl(maxit = 20, verbose = FALSE)

# IBD: null model, no landscape or environment predictors
fit_ibd <- terradish(
  melip.Fst ~ 1,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  leverage          = FALSE,
  control           = ctrl
)

# IBE only: environmental mismatch, but no landscape resistance
fit_ibe <- terradish(
  melip.Fst ~ 1,           # no raster covariates on the conductance side
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = g_ibe, # IBE is on the measurement side
  leverage          = FALSE,
  control           = ctrl
)

# IBR only: landscape resistance, no environmental mismatch
fit_ibr <- terradish(
  melip.Fst ~ altitude + forestcover,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  leverage          = FALSE,
  control           = ctrl
)

# Joint IBE + IBR: both processes in a single model
fit_joint <- terradish(
  melip.Fst ~ altitude + forestcover,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = g_ibe,
  leverage          = FALSE,
  control           = ctrl
)
```

> **Note on `leverage = FALSE`:** By default, `terradish()` computes
> influence/leverage diagnostics for each observation, which requires
> additional linear-algebra work. Setting `leverage = FALSE` skips this step
> to reduce run time. For a final analysis where you want diagnostics, omit
> this argument or set it to `TRUE`.

---

## Step 4 — Interpret the joint model

### Conductance parameters (theta)

`coef()` returns the conductance parameters (the θ values that map raster
covariates to landscape permeability):

```{r coef-ibr}
coef(fit_ibr)   # IBR-only model
coef(fit_joint) # joint model (theta side only)
```

Interpretation is the same as in any IBR model: a positive θ means higher
covariate values increase conductance (easier movement), a negative θ means
the covariate acts as a barrier. See
`vignette("getting-started", package = "terradish")` for how to convert these
to conductance ratios.

Notice that the θ estimates may shift between `fit_ibr` and `fit_joint`. That
is expected: when the joint model can attribute some of the pairwise variation
to environmental mismatch (γ), it no longer needs the IBR coefficients to carry
that variation.

### IBE coefficients (gamma, inside phi)

Because IBE lives in the measurement model rather than the conductance model,
its coefficient (γ) is reported inside `phi`, the vector of nuisance
parameters:

```{r summary-joint}
summary(fit_joint)
```

Look for the row named `absdiff_<layer>` in the phi table. A positive γ means
that sites with greater altitude differences are more genetically different,
after accounting for the resistance distance between them. A negative γ (rarely
biologically expected for IBE) would mean the opposite.

The other phi parameters (`alpha`, `beta`, `tau`, `rho`) are the MLPE
regression nuisance parameters and are the same as in a plain `mlpe` fit:
intercept, IBR slope, residual precision, and MLPE correlation.

### Quick summary of which parameter lives where

| Parameter | Symbol | Where it appears | What it measures |
|-----------|--------|-----------------|-----------------|
| Conductance | θ | `coef(fit)` | How raster covariates affect landscape permeability |
| IBR slope | β (`beta`) | `summary(fit)$phi` | How resistance distance predicts genetic distance |
| IBE slope | γ (`absdiff_*`) | `summary(fit)$phi` | How environmental mismatch predicts genetic distance |
| Intercept | α (`alpha`) | `summary(fit)$phi` | Baseline genetic distance at zero resistance |
| Residual precision | exp(τ) (`tau`) | `summary(fit)$phi` | Inverse residual variance |
| MLPE correlation | plogis(ρ)/2 (`rho`) | `summary(fit)$phi` | Shared-site correlation in pairwise residuals |

---

## Step 5 — Compare the model set

### Information criteria

`aic_table()` ranks all four models:

```{r aic-table}
aic_table(
  list(fit_ibd, fit_ibe, fit_ibr, fit_joint),
  mod_names = c("IBD", "IBE only", "IBR only", "IBE + IBR (joint)")
)
```

A lower AIC indicates a better balance of fit and parsimony. The delta column
shows how far each model is from the best. Models within 2 AIC units of the
best are generally considered to have comparable support.

Use AICc when the number of sampling sites is modest relative to the number of
parameters:

```{r aicc-table}
aic_table(
  list(fit_ibd, fit_ibe, fit_ibr, fit_joint),
  mod_names = c("IBD", "IBE only", "IBR only", "IBE + IBR (joint)"),
  AICc = TRUE
)
```

### Likelihood ratio tests for nested comparisons

Within the IBR family (models with a conductance surface), nested models can be
tested with `anova()`:

```{r lrt}
# Does the joint model improve on IBR-only?
anova(fit_joint, fit_ibr)

# Does IBR-only improve on IBD?
anova(fit_ibr, fit_ibd)
```

Note that IBE-only (`fit_ibe`) and IBR-only (`fit_ibr`) are **not** nested
in each other; they differ on which side of the model (conductance vs.
measurement) carries the predictors. Use `aic_table()`, not `anova()`, to
compare them.

### Cross-validation for predictive comparison

`cv_model_selection()` combines held-out log-likelihoods with an AIC table so
you can compare models on both in-sample fit and out-of-sample prediction.

`terradish_cv()` requires at least one conductance parameter to cross-validate
on, so it can only be called for models with raster covariates on the
conductance side (IBR-only and joint). The IBD and IBE-only models use `~ 1`,
meaning all cells have equal conductance and there are no θ values to
estimate from the training split. For those models we pass the in-sample
log-likelihood directly as the held-out value, a conservative choice that
does not inflate their apparent predictive performance.

```{r cv-model-selection}
cv_ctrl <- NewtonRaphsonControl(maxit = 15, verbose = FALSE)

# IBR-only: has conductance parameters, so true CV is possible
cv_ibr <- terradish_cv(
  pts = melip.coords, covariates = covariates,
  fmla = melip.Fst ~ altitude + forestcover, model = mlpe,
  prop_train = 2 / 3, seed = 1, fit_full = TRUE,
  control = cv_ctrl
)

# Joint IBE + IBR: has conductance parameters, so true CV is possible
cv_joint <- terradish_cv(
  pts = melip.coords, covariates = covariates,
  fmla = melip.Fst ~ altitude + forestcover, model = g_ibe,
  prop_train = 2 / 3, seed = 1, fit_full = TRUE,
  control = cv_ctrl
)

# IBD and IBE-only: no conductance parameters; use in-sample loglik as proxy
comparison <- cv_model_selection(
  list(
    list(train_mod = fit_ibd,
         cv_loglik  = fit_ibd$loglik,
         full_mod   = fit_ibd),
    list(train_mod = fit_ibe,
         cv_loglik  = fit_ibe$loglik,
         full_mod   = fit_ibe),
    list(train_mod = cv_ibr$train_mod,
         cv_loglik  = cv_ibr$cv_loglik,
         full_mod   = cv_ibr$full_mod),
    list(train_mod = cv_joint$train_mod,
         cv_loglik  = cv_joint$cv_loglik,
         full_mod   = cv_joint$full_mod)
  ),
  cv_names = c("IBD", "IBE only", "IBR only", "IBE + IBR (joint)"),
  aic = TRUE
)

# Held-out log-likelihood (higher is better predictive performance)
comparison$loglik_tab

# AIC table from the full-data fits
comparison$AIC_tab
```

> **Why cross-validate?** AIC measures in-sample fit penalized for complexity.
> Cross-validation directly measures whether the model generalizes to new sites.
> For IBE + IBR models, AIC and CV often agree, but they can diverge when the
> extra IBE parameter fits noise in the training data that does not generalize.

For a more stable estimate of predictive performance, repeat across multiple
random splits using `terradish_cv_replicates()`:

```{r cv-replicates}
cv_joint_reps <- terradish_cv_replicates(
  pts        = melip.coords,
  covariates = covariates,
  fmla       = melip.Fst ~ altitude + forestcover,
  model      = g_ibe,
  seeds      = 1:5,
  fit_full   = FALSE,
  keep_fits  = FALSE,
  control    = cv_ctrl
)

summary(cv_joint_reps)
```

---

## Step 6 — Visualize the joint model

### Observed vs. fitted genetic distances

`plot(fit, type = "fit")` shows the pairwise genetic distances against the
resistance distances implied by the fitted model, with a least-squares
regression line:

```{r obs-vs-fitted, fig.cap = "Observed vs. fitted pairwise genetic distances from the joint model."}
plot(fit_joint, type = "fit")
```

### Fitted conductance surface

The conductance surface reflects only the θ side of the model; the IBE term
does not change how permeable the landscape is estimated to be.
`plot(fit, type = "surface", data = surface)` shows the point estimate and
95% confidence interval bounds in a single three-panel figure:

```{r conductance-surface, fig.cap = "Estimated conductance surface from the joint IBE + IBR model."}
plot(fit_joint, type = "surface", data = surface)
```

### Marginal effects of landscape covariates

`plot(fit, type = "marginal", data = surface)` illustrates how each raster
covariate influences conductance after accounting for the other terms in the
model. Because we used `scale_covariates()`, x-axes are back-transformed to
the original units automatically:

```{r marginal-joint, fig.cap = "Marginal conductance effects for the joint IBE + IBR model."}
plot(fit_joint, type = "marginal", data = surface)
```

Compare these curves to those from the IBR-only model to see whether adding
the IBE term changes the estimated landscape permeability:

```{r marginal-ibr, fig.cap = "Marginal conductance effects for the IBR-only model."}
plot(fit_ibr, type = "marginal", data = surface)
```

---

## Step 7 — Saved-results utilities

Some workflows fit `terradish` models once, save them to disk, and later
extract summaries. Two helpers support this pattern.

`terradish_results()` inspects a terradish-style results directory and recovers
the fitted objects and metadata stored there:

```{r saved-results}
# Create a temporary directory to illustrate the pattern.
tmp_dir <- tempfile("terradish-results-")
dir.create(tmp_dir)

# Save the joint fit and a minimal AllResults object.
saveRDS(fit_joint, file.path(tmp_dir, "fit--mlpe.rds"))
saveRDS(
  list(
    effect_size = coef(fit_joint),
    pts         = melip.coords,
    covariates  = covariates
  ),
  file.path(tmp_dir, "AllResults_list.rds")
)

results_info <- terradish_results(tmp_dir)
names(results_info)
```

`terradish_parameters()` extracts a compact coefficient table from a saved
fitted model. This is useful in simulation studies where you want to compare
known effect sizes with estimated values across many saved runs:

```{r saved-parameters}
parameter_table <- terradish_parameters(
  tmp_dir,
  model      = "mlpe",
  save_table = FALSE
)

parameter_table
```

---

## Putting it all together: a decision guide

| Goal | Recommended approach |
|------|---------------------|
| Does landscape resistance matter? | `anova(fit_ibr, fit_ibd)` |
| Does environment matter after accounting for resistance? | `anova(fit_joint, fit_ibr)`, or inspect γ in `summary(fit_joint)` |
| Which single model has the best in-sample fit? | `aic_table(list(...))` |
| Which model predicts new sites best? | `cv_model_selection(list(...))` |
| Is my predictive result stable across splits? | `terradish_cv_replicates()` |
| What is the fitted landscape surface? | `conductance(surface, fit_joint)` |

The joint model is most useful when you suspect both IBE and IBR are operating.
If the γ estimate is near zero with a wide standard error, the IBE term is not
adding explanatory power and the simpler IBR model is preferable.

---

## What comes next?

- **`vignette("getting-started", package = "terradish")`** — the core
  single-process workflow (load data, scale covariates, fit, visualize).
- **`vignette("model-comparison", package = "terradish")`** — likelihood
  ratio tests, AIC/AICc/BIC, likelihood surface visualization, and
  cross-validation in depth.

---

## Quick reference

```{r quick-reference, eval = FALSE}
library(terradish)
library(terra)

# 1. Load and scale data
data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

covariates <- c(melip.altitude, melip.forestcover)
names(covariates) <- c("altitude", "forestcover")
covariates <- scale_covariates(covariates)
surface <- conductance_surface(covariates, melip.coords, directions = 8)

# 2. Build IBE covariates and measurement model
z_ibe <- pairwise_endpoint_covariates(
  melip.altitude, coords = melip.coords,
  transform = "absdiff", scale = TRUE
)
g_ibe <- mlpe_covariates(z_ibe)

# 3. Fit IBD, IBE, IBR, and joint models
fit_ibd   <- terradish(melip.Fst ~ 1,
                       surface, loglinear_conductance, mlpe)
fit_ibe   <- terradish(melip.Fst ~ 1,
                       surface, loglinear_conductance, g_ibe)
fit_ibr   <- terradish(melip.Fst ~ altitude + forestcover,
                       surface, loglinear_conductance, mlpe)
fit_joint <- terradish(melip.Fst ~ altitude + forestcover,
                       surface, loglinear_conductance, g_ibe)

# 4. Compare
aic_table(list(fit_ibd, fit_ibe, fit_ibr, fit_joint),
          mod_names = c("IBD", "IBE", "IBR", "IBE + IBR"))
anova(fit_joint, fit_ibr)   # does IBE add to IBR?
anova(fit_ibr,   fit_ibd)   # does IBR improve on IBD?

# 5. Interpret theta (IBR) and gamma (IBE inside phi)
coef(fit_joint)
summary(fit_joint)$phi

# 6. Visualize
plot(fit_joint, type = "fit")                       # observed vs. fitted
plot(fit_joint, type = "surface", data = surface)   # conductance map + CI panels
plot(fit_joint, type = "marginal", data = surface)  # marginal covariate effects
```
