---
title: "Concurrent IBE and IBR in terradish"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Concurrent IBE and IBR in terradish}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Overview

`terradish` is built around two model layers:

- A conductance model that turns raster covariates into a resistance surface.
- A measurement model that relates pairwise resistance distances to pairwise genetic distances.

That split makes it natural to fit isolation by environment (IBE) and isolation
by resistance (IBR) concurrently:

- Keep IBR on the conductance side, where raster covariates determine movement
  through space.
- Add IBE on the measurement-model side as fixed endpoint-difference covariates
  between sampling locations.

This vignette walks through a small comparison set:

- IBD: no environmental or resistance predictors.
- IBE only: endpoint-difference covariates, but no resistance-surface predictors.
- IBR only: raster predictors in the conductance model.
- Joint IBE + IBR: both together.

## Example Data

```{r}
library(terradish)
library(terra)

data(melip)
melip.altitude <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords <- terra::unwrap(melip.coords)

# keep the vignette compact and quick to fit
keep <- 1:12
gd_mat <- melip.Fst[keep, keep, drop = FALSE]
coords <- melip.coords[keep]

# scale raster covariates before fitting the conductance model
covariates <- c(terra::scale(melip.altitude),
                terra::scale(melip.forestcover))
names(covariates) <- c("altitude", "forestcover")

surface <- conductance_surface(covariates, coords, directions = 8)
```

## Construct IBE Covariates

The new helper `pairwise_endpoint_covariates()` converts site-level
environmental values into one or more pairwise endpoint-difference covariates.

```{r}
names(melip.altitude) <- "altitude_site"

z_ibe <- pairwise_endpoint_covariates(
  melip.altitude,
  coords = coords,
  transform = "absdiff",
  scale = TRUE
)

head(z_ibe)
dim(z_ibe)
```

Here we use absolute pairwise altitude differences, but the helper also
supports squared differences and multivariate Euclidean or Manhattan
environmental distances.

## Build the Joint MLPE Measurement Model

`mlpe_covariates()` wraps the pairwise endpoint-difference covariates into an
MLPE measurement model with mean structure

\[
S_{ij} = \alpha + \beta R_{ij} + Z_{ij}\gamma + e_{ij}.
\]

```{r}
g_ibe <- mlpe_covariates(z_ibe)
```

## Fit IBD, IBE-only, IBR-only, and Joint Models

```{r}
ctrl <- NewtonRaphsonControl(maxit = 8, verbose = FALSE)

fit_ibd <- radish(
  gd_mat ~ 1,
  data = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  leverage = FALSE,
  control = ctrl
)

fit_ibe <- radish(
  gd_mat ~ 1,
  data = surface,
  conductance_model = loglinear_conductance,
  measurement_model = g_ibe,
  leverage = FALSE,
  control = ctrl
)

fit_ibr <- radish(
  gd_mat ~ altitude + forestcover,
  data = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  leverage = FALSE,
  control = ctrl
)

fit_joint <- radish(
  gd_mat ~ altitude + forestcover,
  data = surface,
  conductance_model = loglinear_conductance,
  measurement_model = g_ibe,
  leverage = FALSE,
  control = ctrl
)
```

The conductance coefficients remain on the `theta` side:

```{r}
coef(fit_ibr)
coef(fit_joint)
```

The IBE coefficient lives in the nuisance-parameter vector `phi` because it is
part of the measurement model rather than the conductance model:

```{r}
summary(fit_joint)$phi
```

## Compare the Model Set

For a quick comparison on the same focal set, `cv_model_selection()` can rank
the four models by likelihood and AIC.

```{r}
comparison <- cv_model_selection(
  list(
    list(train_mod = fit_ibd,   cv_loglik = fit_ibd$loglik,   full_mod = fit_ibd),
    list(train_mod = fit_ibe,   cv_loglik = fit_ibe$loglik,   full_mod = fit_ibe),
    list(train_mod = fit_ibr,   cv_loglik = fit_ibr$loglik,   full_mod = fit_ibr),
    list(train_mod = fit_joint, cv_loglik = fit_joint$loglik, full_mod = fit_joint)
  ),
  cv_names = c("IBD", "IBE", "IBR", "IBE + IBR"),
  aic = TRUE
)

comparison$loglik_tab
comparison$AIC_tab
```

## Cross-validation

`radish_cv()` now knows how to rebuild `mlpe_covariates()` models on train/test
subsets, as long as the measurement model was created from
`pairwise_endpoint_covariates()` or from site-level covariates that can be
converted by that helper.

```{r}
cv_joint <- radish_cv(
  pts = coords,
  covariates = covariates,
  fmla = gd_mat ~ altitude + forestcover,
  model = g_ibe,
  prop_train = 2 / 3,
  seed = 1,
  fit_full = FALSE,
  control = ctrl
)

cv_joint$cv_loglik
```

## Interpretation

A useful way to read the joint model is:

- `theta` asks whether landscape permeability affects effective movement among
  sites.
- `gamma` asks whether endpoint environmental mismatch explains additional
  genetic differentiation after accounting for resistance distance.

In other words, the joint model estimates the unique contribution of IBR and
IBE rather than forcing them to compete in a single predictor class.
