---
title: "Concurrent IBE and IBR in terradish"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Concurrent IBE and IBR in terradish}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Overview

`terradish` is built around two model layers:

- A conductance model that turns raster covariates into a resistance surface.
- A measurement model that relates pairwise resistance distances to pairwise genetic distances.

That split makes it natural to fit isolation by environment (IBE) and isolation
by resistance (IBR) concurrently:

- Keep IBR on the conductance side, where raster covariates determine movement
  through space.
- Add IBE on the measurement-model side as fixed endpoint-difference covariates
  between sampling locations.

This vignette walks through a small comparison set:

- IBD: no environmental or resistance predictors.
- IBE only: endpoint-difference covariates, but no resistance-surface predictors.
- IBR only: raster predictors in the conductance model.
- Joint IBE + IBR: both together.

## Example Data

```{r}
library(terradish)
library(terra)

data(melip)
melip.altitude <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords <- terra::unwrap(melip.coords)

# keep the vignette compact and quick to fit
keep <- 1:12
gd_mat <- melip.Fst[keep, keep, drop = FALSE]
coords <- melip.coords[keep]

# scale raster covariates before fitting the conductance model
covariates <- c(terra::scale(melip.altitude),
                terra::scale(melip.forestcover))
names(covariates) <- c("altitude", "forestcover")

surface <- conductance_surface(covariates, coords, directions = 8)
```

## Construct IBE Covariates

The new helper `pairwise_endpoint_covariates()` converts site-level
environmental values into one or more pairwise endpoint-difference covariates.

```{r}
names(melip.altitude) <- "altitude_site"

z_ibe <- pairwise_endpoint_covariates(
  melip.altitude,
  coords = coords,
  transform = "absdiff",
  scale = TRUE
)

head(z_ibe)
dim(z_ibe)
```

Here we use absolute pairwise altitude differences, but the helper also
supports squared differences and multivariate Euclidean or Manhattan
environmental distances.

## Build the Joint MLPE Measurement Model

`mlpe_covariates()` wraps the pairwise endpoint-difference covariates into an
MLPE measurement model with mean structure

\[
S_{ij} = \alpha + \beta R_{ij} + Z_{ij}\gamma + e_{ij}.
\]

```{r}
g_ibe <- mlpe_covariates(z_ibe)
```

## Fit IBD, IBE-only, IBR-only, and Joint Models

```{r}
ctrl <- NewtonRaphsonControl(maxit = 8, verbose = FALSE)

fit_ibd <- terradish(
  gd_mat ~ 1,
  data = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  leverage = FALSE,
  control = ctrl
)

fit_ibe <- terradish(
  gd_mat ~ 1,
  data = surface,
  conductance_model = loglinear_conductance,
  measurement_model = g_ibe,
  leverage = FALSE,
  control = ctrl
)

fit_ibr <- terradish(
  gd_mat ~ altitude + forestcover,
  data = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  leverage = FALSE,
  control = ctrl
)

fit_joint <- terradish(
  gd_mat ~ altitude + forestcover,
  data = surface,
  conductance_model = loglinear_conductance,
  measurement_model = g_ibe,
  leverage = FALSE,
  control = ctrl
)
```

The conductance coefficients remain on the `theta` side:

```{r}
coef(fit_ibr)
coef(fit_joint)
```

The IBE coefficient lives in the nuisance-parameter vector `phi` because it is
part of the measurement model rather than the conductance model:

```{r}
summary(fit_joint)$phi
```

## Compare the Model Set

For a quick comparison on the same focal set, `cv_model_selection()` can rank
the four models by likelihood and AIC.

```{r}
comparison <- cv_model_selection(
  list(
    list(train_mod = fit_ibd,   cv_loglik = fit_ibd$loglik,   full_mod = fit_ibd),
    list(train_mod = fit_ibe,   cv_loglik = fit_ibe$loglik,   full_mod = fit_ibe),
    list(train_mod = fit_ibr,   cv_loglik = fit_ibr$loglik,   full_mod = fit_ibr),
    list(train_mod = fit_joint, cv_loglik = fit_joint$loglik, full_mod = fit_joint)
  ),
  cv_names = c("IBD", "IBE", "IBR", "IBE + IBR"),
  aic = TRUE
)

comparison$loglik_tab
comparison$AIC_tab
```

`cv_model_selection()` is convenient when you already have cross-validation
objects or want to keep held-out log-likelihood and information-criterion
ranking together in one place.

If you already have fitted `terradish` models and only want an
information-criterion table, `aic_table()` is a simpler helper:

```{r}
# Use the fitted models directly when you only want model ranking.
model_table_aic <- aic_table(
  list(fit_ibd, fit_ibe, fit_ibr, fit_joint),
  mod_names = c("IBD", "IBE", "IBR", "IBE + IBR")
)

# AICc can be useful when the number of focal populations is modest
# relative to the number of estimated parameters.
model_table_aicc <- aic_table(
  list(fit_ibd, fit_ibe, fit_ibr, fit_joint),
  mod_names = c("IBD", "IBE", "IBR", "IBE + IBR"),
  AICc = TRUE
)

# BIC applies a stronger complexity penalty than AIC.
model_table_bic <- aic_table(
  list(fit_ibd, fit_ibe, fit_ibr, fit_joint),
  mod_names = c("IBD", "IBE", "IBR", "IBE + IBR"),
  BIC = TRUE
)

model_table_aic
```

For an entry-level workflow, a good rule of thumb is:

- Start with held-out log-likelihood if prediction is your main goal.
- Use AIC or AICc when you want a quick likelihood-based model ranking on the
  same data set.
- Use BIC when you want a stricter penalty on model complexity.

## Cross-validation

`terradish_cv()` now knows how to rebuild `mlpe_covariates()` models on train/test
subsets, as long as the measurement model was created from
`pairwise_endpoint_covariates()` or from site-level covariates that can be
converted by that helper.

```{r}
cv_joint <- terradish_cv(
  pts = coords,
  covariates = covariates,
  fmla = gd_mat ~ altitude + forestcover,
  model = g_ibe,
  prop_train = 2 / 3,
  seed = 1,
  fit_full = FALSE,
  control = ctrl
)

cv_joint$cv_loglik
```

That result is based on one random split of the focal populations into a
training set and a test set. A single split is helpful for a quick check, but
it can be noisy if the result depends strongly on which sites happened to land
in the training set.

`terradish_cv_replicates()` repeats that same process across multiple random
splits and summarizes the held-out log-likelihood:

```{r}
# Repeat the same train/test workflow across several random splits.
cv_joint_reps <- terradish_cv_replicates(
  pts = coords,
  covariates = covariates,
  fmla = gd_mat ~ altitude + forestcover,
  model = g_ibe,
  seeds = c(1, 2, 3),
  fit_full = FALSE,
  keep_fits = FALSE,
  control = ctrl
)

# The print method gives a quick overview.
cv_joint_reps

# summary() adds the replicate count and the range of held-out loglikelihood.
summary(cv_joint_reps)
```

In practice:

- Use `terradish_cv()` when you want one quick train/test split.
- Use `terradish_cv_replicates()` when you want a more stable sense of average
  predictive performance.

## Saved Results Utilities

Some workflows fit `terradish` models once, save them to disk, and later come
back to extract parameter summaries. The saved-results helpers are designed for
that downstream step rather than for model fitting itself.

`terradish_results()` lets you inspect what a terradish-style results directory
contains:

```{r}
# Create a temporary results directory for demonstration.
tmp_results <- tempfile("terradish-results-")
dir.create(tmp_results)

# Save one fitted model and a simple AllResults object.
saveRDS(fit_joint, file.path(tmp_results, "fit--mlpe.rds"))
saveRDS(
  list(
    effect_size = coef(fit_joint),
    pts = coords,
    covariates = covariates
  ),
  file.path(tmp_results, "AllResults_list.rds")
)

# Inspect what terradish can recover from the directory.
results_info <- terradish_results(tmp_results)
names(results_info)
```

`terradish_parameters()` then extracts a compact coefficient table from the
saved fitted model:

```{r}
# Extract fitted conductance coefficients and their standard errors.
parameter_table <- terradish_parameters(
  tmp_results,
  model = "mlpe",
  save_table = FALSE
)

parameter_table
```

This is especially useful in simulation studies, where you may want to compare
known effect sizes against the fitted values from many saved runs.

## Interpretation

A useful way to read the joint model is:

- `theta` asks whether landscape permeability affects effective movement among
  sites.
- `gamma` asks whether endpoint environmental mismatch explains additional
  genetic differentiation after accounting for resistance distance.

In other words, the joint model estimates the unique contribution of IBR and
IBE rather than forcing them to compete in a single predictor class.
