---
title: "Large Landscapes"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Large Landscapes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4.25,
  warning  = FALSE,
  message  = FALSE
)
```

This vignette covers fitting `terradish` models on large rasters: hundreds of
thousands to millions of grid cells. If you have not read
`vignette("getting-started", package = "terradish")`, start there. It explains
conductance surfaces, the model formula, and how to read a fitted model. Here we
assume you already know how to fit a small model and focus on what changes when
the landscape gets big.

The runnable code uses the bundled `melip` data to show the workflow.
Time and memory depend on graph geometry, conductance contrast, the number of
focal sites, and hardware. Measure those costs on your own design before
committing to a large analysis.

```{r load}
library(terradish)

data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

covariates <- c(melip.altitude, melip.forestcover)
names(covariates) <- c("altitude", "forestcover")
covariates <- scale_covariates(covariates)
surface    <- conductance_surface(covariates, melip.coords, directions = 8)
```

---

## When a landscape is "large"

`terradish` builds a weighted graph over the raster grid and solves a linear
system on that graph to get resistance distances among focal sites. The cost of
that solve grows with the number of grid cells, `N`. Two things change as `N`
grows into the hundreds of thousands and beyond:

1. The linear **solver** that computes resistance distances can run out of
   memory. Picking the right one (Section 3) is what keeps a fit feasible.
2. The **inference** machinery (standard errors and confidence intervals) is
   unchanged in form, but larger analyses can include more focal sites. Wishart
   precision also depends on the supplied effective degrees of freedom, `nu`
   (Section 4).

"Large" here means that graph factorization or repeated derivative solves
dominate the analysis. There is no universal cell-count limit: missing cells,
graph connectivity, and conductance contrast can change the practical limit.

---

## Choosing a solver

Both `terradish()` (the fitting wrapper) and `terradish_algorithm()` (one
likelihood and gradient evaluation) take a `solver` argument. The solver only
changes **how** the resistance-distance linear system is solved; it does not
change the statistical model. Iterative solvers should agree with the direct
solver within their numerical tolerances, but estimates need not be bitwise
identical.

```{r solver-calls, eval = FALSE}
# direct (the default): exact sparse Cholesky factorization
fit <- terradish(melip.Fst ~ forestcover + altitude, surface,
                 conductance_model = loglinear_conductance,
                 solver = "direct")

# amg: algebraic multigrid, the large-N path
fit <- terradish(melip.Fst ~ forestcover + altitude, surface,
                 conductance_model = loglinear_conductance,
                 solver = "amg")
```

The choices are:

- **`direct`** (default) uses sparse Cholesky factorization. It is a useful
  reference for checking iterative results. The factor can contain many more
  nonzero entries than the graph, so its memory requirements can grow rapidly.
- **`amg`** uses algebraic multigrid as an iterative solver. It can reduce memory
  demands on large grids, but convergence depends on conductance contrast and
  graph structure. Strong barriers can require more iterations or tighter
  solver settings. Check agreement with a direct solve on a manageable case.
- **`auto`** uses `direct` below 1,500,000 graph vertices and `amg` at or above
  that threshold with the default controls. The number of focal sites does not
  override the large-graph switch. `solver_control$auto_amg_min_vertices`
  changes the threshold; the legacy `auto_direct_max_rhs` is ignored.

Use `terradish_assess_settings()` for a local speed and accuracy probe. Its
recommendation concerns the tested settings and graph, not the scientific
adequacy of the measurement model. Record the resolved solver and compare
coefficient estimates and likelihoods when changing solver tolerances.

This applies to every symmetric resistance model: any `conductance_model`
(`loglinear_conductance`, `smooth_loglinear_conductance`, the Gaussian
scale-aware model) paired with any measurement model (`mlpe`,
`leastsquares`, `generalized_wishart`, and `wishart_covariance`). The solver sits underneath
all of them.

---

## Inference at scale

Scaling the solver lets you fit larger graph domains and sampling designs. More
informative designs can tighten inference, but cell count alone does not.
Several computational choices matter here.

## Reusing Windows workers

Worker reuse is **experimental**. On Windows, `cores > 1` requests PSOCK
workers for supported Hessian and partial-derivative calculations. It does
not parallelize every step of a fit. The pool is reused during optimization
and stopped before return; fitted objects do not retain live workers.
AMG and the cached CHOLMOD backend ignore this setting. On small graphs,
worker startup and data transfer can make multiple cores slower than one.
Benchmark `cores = 1` first, then compare the elapsed time and numerical
results with the requested worker count.

Use one level of parallelism at a time. In particular, fixed-domain
cross-validation evaluates folds serially and passes `cores` to each fit. This
avoids creating an outer fold cluster whose workers each create another inner
cluster.

## Controlling the nuisance profile

Each conductance-parameter trial profiles the nuisance parameters of the
measurement model. Smooth conductance models paired with
`generalized_wishart` can make that two-level optimization costly because each
outer step may require many inner Newton steps. The inner optimizer already
uses the previous nuisance solution as a warm start. Use
`measurement_control` to inspect or tune it separately from `control`:

```{r measurement-control, eval = FALSE}
fit_smooth_gw <- terradish(
  admissible_squared_distance ~ s(forestcover, df = 4) +
    s(altitude, df = 4),
  surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = generalized_wishart,
  nu = effective_marker_df,
  optimizer = "bfgs",
  control = NewtonRaphsonControl(maxit = 80, verbose = TRUE),
  measurement_control = NewtonRaphsonControl(
    maxit = 40,
    ctol = 1e-8,
    ftol = 1e-10,
    verbose = FALSE
  )
)

fit_smooth_gw$fit$subproblem
```

The returned `convergence` code and `iters` count describe the final nuisance
profile. A value of 0 indicates convergence; 1 means the profile reached
`maxit`; 2 indicates a line-search failure. Do not loosen tolerances simply to
remove a warning. First inspect coefficient stability, gradient size, the
likelihood, and sensitivity to tighter settings. The control separates a
computational decision from the scientific model, but overly loose profiling
can still change the fitted likelihood surface.

## Curvature for standard errors

`terradish()` and `terradish_algorithm()` take `curvature = c("exact",
"gauss_newton")`. The objective and gradient are the same; the options differ
in the Hessian approximation used for optimization and standard errors. Fits
should be compared for numerical agreement rather than assumed identical.

```{r curvature-calls, eval = FALSE}
# Generalized Wishart requires an admissible squared-distance response that is
# coherently related to a centered positive-semidefinite covariance matrix.
# exact (default): full observed Hessian
fit <- terradish(admissible_squared_distance ~ forestcover + altitude, surface,
                 measurement_model = generalized_wishart,
                 nu = effective_marker_df,
                 curvature = "exact")

# gauss_newton: information-based curvature
fit <- terradish(admissible_squared_distance ~ forestcover + altitude, surface,
                 measurement_model = generalized_wishart,
                 nu = effective_marker_df,
                 curvature = "gauss_newton")
```

- **`exact`** uses the full observed Hessian.
- **`gauss_newton`** drops the two residual-weighted second-derivative terms,
  leaving the **Gauss-Newton information** curvature. It needs only the
  first derivatives of conductance and is positive semidefinite when the
  retained measurement-model curvature is positive semidefinite. It flows
  unchanged into `vcov()`, `summary()`, and `confint()`, so the standard errors
  become asymptotic, information-based errors. It approaches the exact Hessian
  when the omitted residual-weighted terms are negligible; equality is not
  guaranteed merely because a fit has converged.

**How to choose.** Both are validated for `leastsquares` and the
`generalized_wishart` measurement model. Use the `exact` default for routine
work. Reach for `gauss_newton` when the optimizer struggles with an indefinite
Hessian far from the optimum, or when you want strictly information-based
standard errors. At a well-behaved fit they may agree closely. Compare them as
a sensitivity check when curvature choice materially changes uncertainty or
convergence.

The focused benchmark in `inst/benchmarks/benchmark-curvature-large-landscape.R`
makes this comparison reproducible. It builds a synthetic heterogeneous raster,
simulates a `generalized_wishart` squared-distance response from known conductance
parameters, then fits the same model with `curvature = "exact"` and
`curvature = "gauss_newton"`.

```{r curvature-benchmark-run, eval = FALSE}
source(system.file("benchmarks", "benchmark-curvature-large-landscape.R",
                   package = "terradish"))

bench <- run_curvature_large_landscape_benchmark(CONFIG)
bench$comparison
```

The default case is deliberately small enough to run during review. To repeat
the same comparison in the large-N solver regime, copy `CONFIG`, increase
`side`, and set `solver = "amg"` or `solver = "auto"` before calling
`run_curvature_large_landscape_benchmark()`.

**How to read the benchmark.** Compare coefficients and likelihoods first,
then standard errors, Hessian eigenvalues, and elapsed time. Agreement in
point estimates does not guarantee agreement in curvature. A large difference
in uncertainty warrants investigation of residuals, conditioning, and model
misspecification before choosing the faster option.

## Information and effective degrees of freedom

For Wishart fits, `nu` is a positive effective degrees-of-freedom value supplied
by the analyst. It is not estimated from the response and is not automatically
the marker count. Linkage and shared population history can reduce the
information represented by many markers. More graph cells alone do not supply
more independent genetic information.

Holding the response and model fixed, changing `nu` leaves the likelihood
optimum unchanged in theory and changes interior standard errors in proportion
to `1 / sqrt(nu)`. Use `terradish_rescale_nu()` to inspect that sensitivity
without refitting. Boundary inference needs separate care. Choose conductance
complexity using held-out prediction and report inference across plausible
`nu` values; see `vignette("model-comparison", package = "terradish")`.

## Cropping and coarse warm starts

Cropping removes routes outside the retained domain and can increase resistance
between sites even when every site remains inside the crop. The supplied
release audit found increases of 15–23% with a two-cell buffer and 1–2% with a
ten-cell buffer. Those are results for that landscape, not universal buffer
recommendations. Compare progressively wider buffers while retaining the same
response, sites, formula, and covariate scaling.

```{r crop-sensitivity, eval = FALSE}
# Scale on the original domain once, before cropping.
buffers <- c(2, 5, 10) * max(terra::res(covariates))
fits_buffer <- lapply(buffers, function(buffer) {
  cropped_surface <- conductance_surface(
    covariates, melip.coords, directions = 8, crop_buffer = buffer
  )
  terradish(melip.Fst ~ forestcover + altitude, cropped_surface,
            measurement_model = mlpe)
})
lapply(fits_buffer, coef)
lapply(fits_buffer, confint)
```

**How to read the output.** Compare estimates and intervals across buffers,
and inspect fitted resistance distances with `terradish_distance()`. Continue
expanding the domain if the substantive conclusion changes. The small-buffer
message is a diagnostic prompt, not proof that a larger buffer is adequate.

Coarse rasters can provide starting values, but the final fit must be refined
on the original graph. Coarsening changes both routes and covariate summaries.
The supported coarse-raster workflow therefore requires exact refinement:

```{r coarse-warm-start, eval = FALSE}
fit <- terradish(
  melip.Fst ~ forestcover + altitude, surface,
  measurement_model = mlpe,
  approximation = "coarse_raster",
  approximation_control = list(factor = c(4, 2), exact_refine = TRUE)
)
summary(fit)
```

Inspect convergence after the final refinement. A converged coarse fit does
not establish convergence on the original graph. `terradish_grid()` accepts
only `approximation = "none"`; coarse-only likelihoods are not supported
for ranking models.

---

## Storing large fitted objects

Fitted conductance and measurement-model closures can retain their construction
environments, including large graph objects. That is useful for prediction but
can make `saveRDS()` much larger than the visible coefficient table suggests.
Use `slim_terradish()` when the durable artifact is intended for inference
summaries rather than later prediction:

```{r slim-fit, eval = FALSE}
fit_slim <- slim_terradish(fit)
fit_slim$storage

saveRDS(fit_slim, "terradish-fit-slim.rds")

# Equivalent shorthand at fit time
fit_slim <- terradish(
  admissible_squared_distance ~ cov1 + cov2,
  surface,
  measurement_model = generalized_wishart,
  nu = effective_marker_df,
  slim = TRUE
)
```

The default removes `submodels` and `leverage`. Coefficients, likelihood,
Hessian-based summaries, fitted values, diagnostics, and permutation
simulation remain available. Conductance prediction and most plot types do not,
because they must reevaluate the removed closures. Slimming is irreversible.
Keep the original formula, graph, response, and an unslimmed fit whenever
prediction or model reconstruction may be needed. Add `"response"` to `drop`
only when the response itself must also be excluded from the archive; doing so
also removes the strongest stored check for later model comparison.

---

## Quick reference

```{r quick-ref, eval = FALSE}
# 1. Build the surface as usual
surface <- conductance_surface(covariates, coords, directions = 8)

# 2. Fit at scale: use amg once the raster is too large for the direct solver
# Generalized Wishart requires an admissible squared-distance response.
fit <- terradish(admissible_squared_distance ~ cov1 + cov2, surface,
                 conductance_model = loglinear_conductance,
                 measurement_model = generalized_wishart,
                 nu = effective_marker_df,
                 solver    = "amg",            # large-N linear solver
                 curvature = "gauss_newton")   # information-based standard errors

summary(fit)                                   # conditional estimates and SEs

# Optional archival copy when prediction from the saved object is unnecessary
fit_archive <- slim_terradish(fit)
saveRDS(fit_archive, "fit-archive.rds")

```

## Summary of key functions

| Function | Purpose |
|----------|---------|
| `conductance_surface()` | Build the weighted graph over the raster grid |
| `terradish()` | Fit the model; takes `solver` and `curvature` |
| `terradish_algorithm()` | One likelihood and gradient evaluation; same `solver` and `curvature` |
| `crop_to_focal_buffer()` | Crop with an explicit map-unit buffer; check domain sensitivity |
| `terradish_assess_settings()` | Compare computational settings on the current graph |
| `slim_terradish()` | Remove prediction closures and other optional large components before storage |

## See also

- **`vignette("getting-started", package = "terradish")`** the core workflow:
  data in, model fit, interpretation out.
- **`vignette("wishart-covariance", package = "terradish")`** the distance and
  covariance Wishart measurement models and the role of `nu`.
- **`vignette("model-comparison", package = "terradish")`** ranking competing
  models with AIC and likelihood ratio tests.
