---
title: "Large Landscapes"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Large Landscapes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4.25,
  warning  = FALSE,
  message  = FALSE
)
```

This vignette covers fitting `terradish` models on large rasters: hundreds of
thousands to millions of grid cells. If you have not read
`vignette("getting-started", package = "terradish")`, start there. It explains
conductance surfaces, the model formula, and how to read a fitted model. Here we
assume you already know how to fit a small model and focus on what changes when
the landscape gets big.

A note on the numbers below. A vignette cannot run a million-cell fit, so the
runnable code uses the small bundled `melip` data only to show the calls. The
scaling results are values measured on rasters far larger than `melip`, on a
64 GB workstation. Absolute times are hardware dependent; the pattern is the
point.

```{r load}
library(terradish)

data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

covariates <- c(melip.altitude, melip.forestcover)
names(covariates) <- c("altitude", "forestcover")
covariates <- scale_covariates(covariates)
surface    <- conductance_surface(covariates, melip.coords, directions = 8)
```

---

## When a landscape is "large"

`terradish` builds a weighted graph over the raster grid and solves a linear
system on that graph to get resistance distances among focal sites. The cost of
that solve grows with the number of grid cells, `N`. Three things change as `N`
grows into the hundreds of thousands and beyond:

1. The linear **solver** that computes resistance distances can run out of
   memory. Picking the right one (Section 3) is what keeps a fit feasible.
2. The **inference** machinery (standard errors and confidence intervals) is
   unchanged in form, but larger analyses can include more focal sites. Wishart
   precision also depends on the supplied effective degrees of freedom, `nu`
   (Section 4).
3. An exact, **memory-bounded reduction** onto the focal sites becomes useful as
   a diagnostic and as an out-of-core building block (Section 5).

"Large" here means the regime where the default solver's memory use becomes the
binding constraint. On a 64 GB workstation that boundary falls between roughly
1.3 and 2.1 million cells; on a smaller machine it falls sooner.

---

## Choosing a solver

Both `terradish()` (the fitting wrapper) and `terradish_algorithm()` (one
likelihood and gradient evaluation) take a `solver` argument. The solver only
changes **how** the resistance-distance linear system is solved; it does not
change the statistical model. Iterative solvers should agree with the direct
solver within their numerical tolerances, but estimates need not be bitwise
identical.

```{r solver-calls, eval = FALSE}
# direct (the default): exact sparse Cholesky factorization
fit <- terradish(melip.Fst ~ forestcover + altitude, surface,
                 conductance_model = loglinear_conductance,
                 solver = "direct")

# amg: algebraic multigrid, the large-N path
fit <- terradish(melip.Fst ~ forestcover + altitude, surface,
                 conductance_model = loglinear_conductance,
                 solver = "amg")
```

The choices:

- **`direct`** (default): a sparse Cholesky factorization. It is the fastest
  option through several hundred thousand cells. Its factor "fill" (the nonzeros
  the factorization adds) grows faster than the graph itself, so past some size
  the factor no longer fits in memory. That is the **memory wall**.
- **`amg`**: smoothed-aggregation **algebraic multigrid**, an iterative solver.
  Each solve costs roughly linear time in `N`, and its iteration count stays
  nearly constant as `N` grows (this is called mesh-independent convergence), so
  total cost scales near-linearly. This is the intended path past the wall.
- **`block_cg`**, **`pcg`**, **`pcg_jacobi`**: conjugate-gradient variants for
  small or well-conditioned graphs. `block_cg` shares one Krylov subspace across
  all focal sites. These are fallbacks, not the large-N path.
- **`auto`**: lets `terradish` pick.

Measured behavior on a heterogeneous grid (one likelihood and gradient
evaluation, 50 focal sites):

| cells     | `direct` time | `amg` time | `amg` iterations |
|-----------|---------------|------------|------------------|
| 65,536    | 1.1 s         | 2.9 s      | 10               |
| 262,144   | 3.9 s         | 7.0 s      | 13               |
| 524,176   | 6.9 s         | 10.0 s     | 11               |
| 1,048,576 | 14.4 s        | 21.2 s     | 12               |
| 1,327,104 | 22.4 s        | 30.1 s     | 14               |
| 2,096,704 | memory wall   | 39.2 s     | 12               |
| 4,194,304 | memory wall   | 74.2 s     | 13               |

**How to read this table.** `direct` is faster wherever it fits, but its time
grows faster than the cell count and it eventually runs out of memory (here
between 1.3 and 2.1 million cells). `amg` is slower at small sizes because each
solve carries setup overhead, but its **iteration count barely moves** (10 to 14
across a 64-fold increase in cells), so doubling the cells roughly doubles the
time rather than more than doubling it, and it keeps running well past the wall.
The decision rule: use `direct` while it fits in memory, and switch to `amg`
once the raster is large enough that `direct` fails or thrashes.

This applies to every symmetric resistance model: any `conductance_model`
(`loglinear_conductance`, `smooth_loglinear_conductance`, the Gaussian
scale-aware and hierarchical models) paired with any measurement model (`mlpe`,
`leastsquares`, `generalized_wishart`, and `wishart_covariance`). The solver sits underneath
all of them. The one exception is the directional model
(`terradish_directed()`), which runs a different algorithm on an asymmetric
generator; see `vignette("directional-conductance", package = "terradish")`.

---

## Inference at scale

Scaling the solver lets you fit larger graph domains and sampling designs. More
informative designs can tighten inference, but cell count alone does not. Two
settings matter here.

## Curvature for standard errors

`terradish()` and `terradish_algorithm()` take `curvature = c("exact",
"gauss_newton")`. The objective and gradient are the same; the options differ
in the Hessian approximation used for optimization and standard errors. Fits
should be compared for numerical agreement rather than assumed identical.

```{r curvature-calls, eval = FALSE}
# Generalized Wishart requires an admissible squared-distance response that is
# coherently related to a centered positive-semidefinite covariance matrix.
# exact (default): full observed Hessian
fit <- terradish(admissible_squared_distance ~ forestcover + altitude, surface,
                 measurement_model = generalized_wishart,
                 nu = effective_marker_df,
                 curvature = "exact")

# gauss_newton: information-based curvature
fit <- terradish(admissible_squared_distance ~ forestcover + altitude, surface,
                 measurement_model = generalized_wishart,
                 nu = effective_marker_df,
                 curvature = "gauss_newton")
```

- **`exact`** uses the full observed Hessian.
- **`gauss_newton`** drops the two residual-weighted second-derivative terms,
  leaving the **Gauss-Newton information** curvature. It needs only the
  first derivatives of conductance and is positive semidefinite when the
  retained measurement-model curvature is positive semidefinite. It flows
  unchanged into `vcov()`, `summary()`, and `confint()`, so the standard errors
  become asymptotic, information-based errors. It approaches the exact Hessian
  when the omitted residual-weighted terms are negligible; equality is not
  guaranteed merely because a fit has converged.

**How to choose.** Both are validated for `leastsquares` and the
`generalized_wishart` measurement model. Use the `exact` default for routine
work. Reach for `gauss_newton` when the optimizer struggles with an indefinite
Hessian far from the optimum, or when you want strictly information-based
standard errors. At a well-behaved fit they may agree closely. Compare them as
a sensitivity check when curvature choice materially changes uncertainty or
convergence.

The focused benchmark in `inst/benchmarks/benchmark-curvature-large-landscape.R`
makes this comparison reproducible. It builds a synthetic heterogeneous raster,
simulates a `generalized_wishart` squared-distance response from known conductance
parameters, then fits the same model with `curvature = "exact"` and
`curvature = "gauss_newton"`.

```{r curvature-benchmark-run, eval = FALSE}
source(system.file("benchmarks", "benchmark-curvature-large-landscape.R",
                   package = "terradish"))

bench <- run_curvature_large_landscape_benchmark(CONFIG)
bench$comparison
```

The default case is deliberately small enough to run during review. To repeat
the same comparison in the large-N solver regime, copy `CONFIG`, increase
`side`, and set `solver = "amg"` or `solver = "auto"` before calling
`run_curvature_large_landscape_benchmark()`.

On the development workstation, the default 96 x 96 synthetic case gave:

| Comparison | Result |
|------------|--------|
| Maximum absolute coefficient difference | 0 |
| Absolute log-likelihood difference | 0 |
| `ridge` SE ratio, `gauss_newton` / `exact` | 0.997 |
| `basin` SE ratio, `gauss_newton` / `exact` | 0.994 |
| Elapsed-time ratio, `gauss_newton` / `exact` | 0.807 |
| Minimum exact-Hessian eigenvalue | 2363.5 |
| Minimum Gauss-Newton eigenvalue | 2375.9 |

**How to read this benchmark.** The fitted surface did not move: the
coefficients and log likelihood were identical. The standard errors were within
1% because the simulated data were generated from the fitted model and the
residual-weighted second-derivative terms were small. In this setting,
`gauss_newton` was slightly faster and returned a positive-definite information
matrix. If your real data show much larger differences in standard errors, run
both curvature settings as a sensitivity check; that gap is evidence that model
residuals or misspecification are affecting the observed curvature.

## Markers and sites buy precision

With the `generalized_wishart` measurement model, `nu` is the supplied effective
Wishart degrees of freedom, not an estimated parameter or automatically the raw
marker count. Holding the response and model fixed while varying `nu` shows the
theoretical scaling pattern (focal sites held at 80):

| `nu`  | mean standard error |
|-------|---------------------|
| 500   | 0.0044              |
| 1,000 | 0.0031              |
| 2,000 | 0.0022              |
| 4,000 | 0.0015              |

**How to read this.** The point estimates do not change with `nu`; only their
precision does. The standard error falls as `1 / sqrt(nu)`: increasing the
effective degrees of freedom eightfold (500 to 4,000) tightens the standard
error about 2.9-fold,
which is close to `sqrt(8) = 2.83`. Standard errors also shrink as you add focal
sites in many designs, but the amount depends on their placement, covariate
support, and the graph. Report a justified primary `nu` and sensitivity values;
for microsatellites, use locus count as primary and $\sum_l(K_l-1)$ only as a
larger sensitivity value.

---

## An exact, memory-bounded reduction

`terradish_kron_reduce_tiled()` reduces the full graph Laplacian onto the focal
sites exactly, returning the focal effective-resistance structure while keeping
peak memory in check. It is the same result as `terradish_kron_reduce()` (the
single-shot reduction), computed so that it does not have to factorize the whole
interior at once.

```{r reduce}
# a conductance vector to reduce (evaluate the model at some theta)
model <- loglinear_conductance(~ forestcover + altitude, surface$x)
cond  <- model(c(0.3, -0.2))$conductance

red <- terradish_kron_reduce_tiled(surface, cond)
red$method                 # which strategy "auto" chose
red$peak$interior          # largest single factorization (vertices)
red$n_interior             # interior vertices in total

# forcing nested dissection bounds the largest factorization far below the
# whole interior (this is the path that keeps memory in check at scale)
nested <- terradish_kron_reduce_tiled(surface, cond, method = "nested")
nested$peak$interior

# both are identical to the single-shot reduction
single <- terradish_kron_reduce(surface, cond)
P <- match(single$boundary, red$boundary)
max(abs(as.matrix(red$laplacian[P, P]) -
        as.matrix(single$laplacian))) / max(abs(as.matrix(single$laplacian)))
```

**How to read the output.** `method` reports the strategy actually used; on a
graph this small `"auto"` chooses `"direct"`, so `peak$interior` equals
`n_interior` (the single-shot path factorizes the whole interior at once). When
you force `"nested"`, `peak$interior` drops far below `n_interior`: that gap is
the memory the bounded path saves, and it is what lets the reduction run on
rasters where the single-shot factor would not fit. The final number is the
relative difference from the single-shot reduction; it is at the level of
floating-point round-off, confirming every path is exact.

The strategy is controlled by `method` and `mem_budget`:

| Argument     | Default  | What it does |
|--------------|----------|--------------|
| `method`     | `"auto"` | `"auto"` uses the fast single-shot factorization while it is estimated to fit `mem_budget`, and switches to nested dissection only when it would not. `"direct"` forces single-shot; `"nested"` forces nested dissection. |
| `mem_budget` | `4e9`    | The memory budget in bytes for the `"auto"` decision. The estimate (returned as `est_factor_bytes`) tracks the single-shot factor size. |
| `cores`      | `1`      | Worker processes for nested dissection. With `cores > 1` its independent recursive halves run in parallel by forking on Unix. The result is identical to the sequential run. |
| `n_tiles`    | `2000`   | The nested-dissection leaf size (vertices per subdomain). |

The two strategies trade speed against memory:

- **single-shot** (`"direct"`) factorizes the whole interior in one pass. It is
  fast (this is the CHOLMOD path), but the factor can exhaust memory at large
  scale, the same wall the `direct` solver hits.
- **nested dissection** (`"nested"`) recursively bisects the interior with thin
  separators, so no single factorization exceeds a separator width. Its memory
  stays bounded at scales where the single-shot factor would not fit. It is
  markedly slower per element, which is exactly why `"auto"` reserves it for the
  regime where the single-shot factor would run out of memory.

**Scope.** This reduction is a standalone, exact primitive: a diagnostic and an
out-of-core building block that returns the focal effective-resistance
structure. It is **not** wired into model fitting; to fit a model on a large
raster, use `solver = "amg"` (Section 3), not this function. It also applies to
the symmetric resistance models only, not the directional model.

---

## Quick reference

```{r quick-ref, eval = FALSE}
# 1. Build the surface as usual
surface <- conductance_surface(covariates, coords, directions = 8)

# 2. Fit at scale: use amg once the raster is too large for the direct solver
# Generalized Wishart requires an admissible squared-distance response.
fit <- terradish(admissible_squared_distance ~ cov1 + cov2, surface,
                 conductance_model = loglinear_conductance,
                 measurement_model = generalized_wishart,
                 nu = effective_marker_df,
                 solver    = "amg",            # large-N linear solver
                 curvature = "gauss_newton")   # information-based standard errors

summary(fit)                                   # conditional estimates and SEs

# 3. (Advanced) exact memory-bounded reduction onto the focal sites
cond_map <- conductance(surface, fit)[["est"]]
cond <- terra::values(cond_map, mat = FALSE)
cond <- cond[is.finite(cond)]                   # active-cell conductance vector
red  <- terradish_kron_reduce_tiled(surface, cond)  # auto: fast unless it would not fit
red$method                                     # "direct" or "nested"
```

## Summary of key functions

| Function | Purpose |
|----------|---------|
| `conductance_surface()` | Build the weighted graph over the raster grid |
| `terradish()` | Fit the model; takes `solver` and `curvature` |
| `terradish_algorithm()` | One likelihood and gradient evaluation; same `solver` and `curvature` |
| `terradish_kron_reduce()` | Single-shot exact reduction onto the focal sites |
| `terradish_kron_reduce_tiled()` | Memory-aware exact reduction (`auto`/`direct`/`nested`) |

## See also

- **`vignette("getting-started", package = "terradish")`** the core workflow:
  data in, model fit, interpretation out.
- **`vignette("wishart-covariance", package = "terradish")`** the distance and
  covariance Wishart measurement models and the role of `nu`.
- **`vignette("model-comparison", package = "terradish")`** ranking competing
  models with AIC and likelihood ratio tests.
- **`vignette("directional-conductance", package = "terradish")`** the
  directional model, which uses a separate algorithm.
