---
title: "Model Comparison and Selection in terradish"
author: "Bill Peterman"
output: 
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Model Comparison and Selection in terradish}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4.5,
  warning  = FALSE,
  message  = FALSE
)

```

## Overview

Once you can fit a `terradish` model (see
`vignette("getting-started", package = "terradish")`), the natural next
questions are:

1. Does forest cover have conditional likelihood support after accounting for
   the other predictors, or could the apparent association be noise?
   (**hypothesis testing**)
2. Is a model with both altitude and forest cover better than one with only
   altitude? (**model selection**)
3. How well does the model predict genetic structure at sites not used for
   fitting? (**cross-validation**)

This vignette walks through all three.

---

## Set up: data and baseline model

```{r load-and-fit}
library(terradish)
library(terra)

data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

covariates <- c(melip.altitude, melip.forestcover)
names(covariates) <- c("altitude", "forestcover")
covariates <- scale_covariates(covariates)

surface <- conductance_surface(
  covariates,
  melip.coords,
  directions = 8,
  saveStack = TRUE
)

# Lightweight one-time assessment: reuse recommendations for all fits below.
assessment <- terradish_assess_settings(
  melip.Fst ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  probe_maxit       = 1,
  verbose           = FALSE
)
rec <- assessment$recommended

# Full model: both covariates
fit_full <- terradish(
  melip.Fst ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)
```

### A note on computational settings

When fitting many candidate models (as this vignette does), it pays to profile
the graph and identify the fastest optimizer/solver combination **once**, then
reuse those settings for all fits. `terradish_assess_settings()` does this in a
few short probe iterations:

```{r assess-for-comparison, eval = FALSE}
assessment <- terradish_assess_settings(
  melip.Fst ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  probe_maxit       = 2,
  verbose           = FALSE
)

# One-line summary: did the assessment differ from defaults?
cat(assessment$comparison$summary, "\n")

# Extract recommended settings and reuse for every model in this vignette
rec <- assessment$recommended
```

Then pass `rec$optimizer`, `rec$control`, `rec$solver`, and `rec$solver_control`
to each `terradish()` call below. For the small *Melipona* graph the defaults
are fine; the pattern becomes important for large landscapes where each function
evaluation is expensive. See
`vignette("getting-started", package = "terradish")` for a full explanation of
the output and all available probe options.

---

## Hypothesis testing with likelihood ratio tests

A **likelihood ratio test (LRT)** asks whether adding a covariate improves the
model fit enough to reject the reduced model under its assumptions. The test
statistic is:

> 2 × (log-likelihood of larger model − log-likelihood of smaller model)

Under the null hypothesis that the extra parameter does not help, this statistic
follows asymptotically a χ² distribution with degrees of freedom equal to the
number of extra parameters.  This approximation improves as sample size
(number of focal sites) increases.

`terradish` provides `anova()` for this:

```{r lrt}
# Reduced model: only altitude, no forest cover
fit_altitude_only <- terradish(
  melip.Fst ~ altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)

anova(fit_full, fit_altitude_only)
```

A small p-value means that removing forest cover worsens the likelihood more
than expected under the reduced model and its asymptotic reference
distribution. This result supports a conditional association between forest
cover and genetic structure through the fitted conductance surface. It does not
by itself establish a causal effect, movement mechanism, or absolute
permeability.

### Testing against the null model of isolation by distance (IBD)

The formula `~ 1` fits a null conductance surface where all cells have equal
conductance. Resistance distance then equals graph distance (weighted only by
geography, not landscape features). This is the IBD baseline:

```{r ibd-test}
fit_ibd <- terradish(
  melip.Fst ~ 1,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)

# Is the full model significantly better than pure IBD?
anova(fit_full, fit_ibd)
```

### Testing for an interaction

An interaction term asks whether the fitted association between altitude and
conductance depends on forest cover (or vice versa):

```{r interaction-test}
fit_interaction <- terradish(
  melip.Fst ~ forestcover * altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)

# Is the interaction term worth keeping?
anova(fit_full, fit_interaction)
```

> **Tip:** Always compare nested models (smaller vs. larger). The model with
> fewer parameters must be a special case of the model with more parameters.
> `anova()` rejects different measurement models and obvious formula
> non-nesting, but the analyst must still verify substantive nesting and whether
> boundary parameters invalidate the usual chi-squared reference distribution.

---

## Information criteria: AIC, AICc, and BIC

Likelihood ratio tests require nested models. When comparing non-nested models
(e.g. altitude-only vs. forest-cover-only), use an information criterion
instead.

```{r fit-all-candidates}
fit_fc_only <- terradish(
  melip.Fst ~ forestcover,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)
```

`aic_table()` ranks fitted `terradish` models only when they use the same
response, focal sites, graph domain, measurement-model family, and pair subset.
Wishart fits must also use the same supplied `nu`. Cross-family comparisons,
such as MLPE versus Wishart, do not have a common likelihood basis and are
rejected.

```{r aic-table}
aic_table(
  list(fit_ibd, fit_altitude_only, fit_fc_only, fit_full, fit_interaction),
  mod_names = c("IBD", "Altitude only", "Forestcover only",
                "Full (A + FC)", "Interaction (A × FC)")
)
```

**How to read the table:**

- **logLik**: larger is better (less negative = higher likelihood).
- **df**: number of estimated parameters; more complex models have higher df.
- **AIC**: balances fit and complexity. Smaller is better.
- **delta**: difference from the best model's AIC. Values below about 2 are a
  common heuristic for similar support within the declared candidate set, not
  proof that the models are equivalent.

### When to use AICc

AIC can over-fit when the effective sample size is small relative to the number
of parameters. `terradish` provides an AICc calculation that uses the number of
focal sampling sites as `n`:

```{r aicc-table}
aic_table(
  list(fit_ibd, fit_altitude_only, fit_fc_only, fit_full, fit_interaction),
  mod_names = c("IBD", "Altitude only", "Forestcover only",
                "Full", "Interaction"),
  AICc = TRUE
)
```

The focal-site convention recognizes that pairwise genetic observations are
dependent, but it is still an approximation. Report the convention and use the
result as one part of the evidence. A common rule of thumb is to prefer AICc
when `n / K < 40`, where `K` includes all estimated parameters. Predeclare and
report the candidate model set and any search over Gaussian smoothing scale;
an optimized scale is an estimated parameter.

### BIC for stricter parsimony

BIC penalizes complexity more heavily than AIC and tends to favor simpler
models. The implementation here uses the number of selected pair rows for a
pair-subset fit and otherwise uses the number of pairwise observations
(`n = n_sites * (n_sites - 1) / 2`) as the BIC sample size. AICc instead uses
the number of focal sites. Which count is the most appropriate `n` for
landscape-genetic data is debated; treat BIC values as one input among several
rather than a single definitive arbiter of model choice:

```{r bic-table}
aic_table(
  list(fit_ibd, fit_altitude_only, fit_fc_only, fit_full, fit_interaction),
  mod_names = c("IBD", "Altitude only", "Forestcover only",
                "Full", "Interaction"),
  BIC = TRUE
)
```

---

## Visualizing the likelihood surface

When a model has only one or two conductance parameters, it is informative to
map how the log-likelihood changes across parameter space. This reveals:

- How sharp (or flat) the optimum is.
- Whether there are multiple local optima.
- Whether the estimates are well-determined or poorly constrained.

`terradish_grid()` evaluates the model at every point on a parameter grid:

```{r likelihood-grid}
# Define a 21 x 21 grid of (forestcover, altitude) combinations
theta_grid <- as.matrix(
  expand.grid(
    forestcover = seq(-1, 1, length.out = 21),
    altitude    = seq(-1, 1, length.out = 21)
  )
)

grid_result <- terradish_grid(
  theta     = theta_grid,
  formula   = melip.Fst ~ forestcover + altitude,
  data      = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe
)
```

Plot the result as a heat map with contours:

```{r plot-likelihood-surface, fig.cap = "***Log-likelihood surface. Red dot marks the MLE.***"}
# Build a data frame for plotting
grid_df <- data.frame(
  forestcover = theta_grid[, "forestcover"],
  altitude    = theta_grid[, "altitude"],
  loglik      = grid_result$loglik
)

# Base-R image plot (no ggplot2 dependency)
loglik_mat <- matrix(
  grid_df$loglik,
  nrow = length(unique(grid_df$forestcover)),
  ncol = length(unique(grid_df$altitude))
)

image(
  x    = sort(unique(grid_df$forestcover)),
  y    = sort(unique(grid_df$altitude)),
  z    = loglik_mat,
  xlab = expression(theta[forestcover]),
  ylab = expression(theta[altitude]),
  main = "Log-likelihood surface",
  col  = hcl.colors(64, "Blues", rev = TRUE)
)
contour(
  x    = sort(unique(grid_df$forestcover)),
  y    = sort(unique(grid_df$altitude)),
  z    = loglik_mat,
  add  = TRUE
)
points(coef(fit_full)["forestcover"],
       coef(fit_full)["altitude"],
       pch = 19, col = "red", cex = 1.5)
legend("topright", legend = "MLE", pch = 19, col = "red", bty = "n")
```

A well-behaved likelihood surface has a single, sharp peak at the MLE. A flat
or elongated surface signals that the parameters are poorly identified, often
because the covariates are highly correlated or the signal-to-noise ratio
in the genetic data is low.

For large rasters, a full likelihood grid can become expensive. If the surface
was created with `saveStack = TRUE`, `terradish_grid()` can screen a grid on an
aggregated raster:

```{r likelihood-grid-coarse, eval=FALSE}
grid_coarse <- terradish_grid(
  theta = theta_grid,
  formula = melip.Fst ~ forestcover + altitude,
  data = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  approximation = "coarse_raster",
  approximation_control = list(factor = 2L)
)
```

This is a screening tool rather than an exact replacement for the full grid:
the log-likelihoods are evaluated on the aggregated graph. Use it to locate
interesting regions of parameter space, then rerun a smaller grid or a final
optimization on the full-resolution graph.

---

## Confidence intervals on the conductance surface

`plot(fit, type = "surface", data = surface)` renders the fitted conductance
alongside its asymptotic lower and upper confidence interval bounds in a
single three-panel figure. Uncertainty in θ is propagated through the
log-linear model via the delta method, and all panels share the same color
scale:

```{r ci-conductance, fig.cap = "***Fitted conductance with 95% confidence bounds.***"}
plot(fit_full, type = "surface", data = surface)
```

Wide intervals in a region mean that the available genetic data provide little
constraint on conductance there, either because few sampling sites are nearby
or because the genetic signal is weak.

## Marginal associations on conductance

`plot(fit, type = "marginal", data = surface)` sweeps each covariate across
its observed range while averaging over all other covariates. The shaded ribbon
is a 95% pointwise confidence band from the delta method. Because we used
`scale_covariates()` to prepare the rasters, x-axes are automatically
back-transformed to the original measurement units:

```{r marginal-conductance, fig.cap = "***Marginal association of each covariate with conductance.***"}
plot(fit_full, type = "marginal", data = surface)
```

A rising curve means that higher covariate values are associated with higher
fitted conductance, conditional on the other predictors and the model. A
declining curve indicates the opposite fitted association. These curves do not
directly measure permeability, movement, habitat suitability, or a causal
barrier effect. Flat or wide-interval curves suggest that the covariate adds
little constraint beyond the other predictors already in the model.

### Handling weakly supported tails

When focal sites sparsely cover a covariate tail, unconstrained marginals can
show unstable extremes that dominate interpretation. Use focal-support clamping
to keep model comparison focused on supported covariate ranges:

```{r support-clamp-marginal, eval = FALSE}
plot(
  fit_full,
  type = "marginal",
  data = surface,
  support = "focal",
  support_probs = c(0.01, 0.99),
  clamp_covariates = c("altitude", "forestcover")
)
```

The same options are available in `conductance()` when producing raster
surfaces for map comparison:

```{r support-clamp-conductance, eval = FALSE}
cond_full_focal <- conductance(
  surface,
  fit_full,
  support = "focal",
  support_probs = c(0.01, 0.99),
  clamp_covariates = c("altitude", "forestcover")
)
```

---

## Cross-validation

Information criteria measure in-sample fit. Cross-validation evaluates
**out-of-sample** prediction under a specified splitting design.

`terradish_cv()` randomly partitions the focal sampling sites into a training
set and a test set, fits the conductance coefficients on the training set, and
evaluates the held-out likelihood at those coefficients. Measurement-model
nuisance parameters are reprofiled on the test subset. Random site splits can
retain strong spatial similarity between training and test sites, so they may
overstate transfer to new regions. Use spatially structured folds when that is
the intended prediction target.

```{r cv-single}
cv_result <- terradish_cv(
  pts       = melip.coords,
  covariates = covariates,
  fmla      = melip.Fst ~ forestcover + altitude,
  model     = mlpe,
  prop_train = 2 / 3,   # use 2/3 of sites for training
  seed      = 42,
  fit_full  = FALSE     # skip re-fitting on the full data (saves time)
)

cat("Held-out log-likelihood:", round(cv_result$cv_loglik, 2), "\n")
```

A single split can be noisy; it may depend heavily on which sites happened to
land in the training set. `terradish_cv_replicates()` repeats the process
across many random splits:

```{r cv-replicates}
cv_reps <- terradish_cv_replicates(
  pts        = melip.coords,
  covariates = covariates,
  fmla       = melip.Fst ~ forestcover + altitude,
  model      = mlpe,
  seeds      = 1:5,    # five random splits
  fit_full   = FALSE,
  keep_fits  = FALSE
)

print(cv_reps)
summary(cv_reps)
```

### Cross-validation for model selection

To compare models using held-out log-likelihood, fit each candidate with
`terradish_cv()` or `terradish_cv_replicates()` and then pass the results to
`cv_model_selection()`. In this example, we compare a one-covariate model to
the two-covariate model. A pure intercept-only MLPE model (`~ 1`) can fit on
the full data but still collapse to a no-coefficient boundary on some training
splits, which makes it a fragile teaching example for cross-validation.

```{r cv-selection}
# For speed, use a reduced control with fewer Newton iterations
ctrl <- NewtonRaphsonControl(maxit = 10, verbose = FALSE)

# Subset the raster stack for the simpler model so the example stays quiet
# and only uses the covariate that appears in the formula.
forestcover_only <- covariates[["forestcover"]]

cv_forestcover <- terradish_cv(
  pts = melip.coords, covariates = forestcover_only,
  fmla = melip.Fst ~ forestcover, model = mlpe,
  prop_train = 2 / 3, seed = 42, fit_full = TRUE, control = ctrl
)

cv_full <- terradish_cv(
  pts = melip.coords, covariates = covariates,
  fmla = melip.Fst ~ forestcover + altitude, model = mlpe,
  prop_train = 2 / 3, seed = 42, fit_full = TRUE, control = ctrl
)

cv_comparison <- cv_model_selection(
  list(
    list(train_mod = cv_forestcover$train_mod,
         cv_loglik = cv_forestcover$cv_loglik,
         full_mod  = cv_forestcover$full_mod),
    list(train_mod = cv_full$train_mod,
         cv_loglik = cv_full$cv_loglik,
         full_mod  = cv_full$full_mod)
  ),
  cv_names = c("Forest cover only", "Full (altitude + forestcover)"),
  aic = TRUE
)

cv_comparison$loglik_tab
cv_comparison$AIC_tab
```

---

## Combining the evidence

A complete analysis typically uses all three approaches together:

| Tool | Answers | When to use |
|------|---------|-------------|
| `anova()` | Does a larger model have incremental likelihood support? | Nested models within one measurement family |
| `aic_table()` | Which declared candidate has the strongest relative support? | Models satisfying the full comparison contract |
| `terradish_cv()` / `cv_model_selection()` | Which candidate predicts the held-out subsets best? | Prediction under the chosen split design |
| `mlpe_response_change()` | How large is an MLPE pairwise-covariate association on the response scale? | IBE terms fitted with `mlpe_covariates()` |

Concordant LRT, information-criterion, and held-out results strengthen evidence
for a conditional association and predictive value under the fitted model and
split design. They still do not establish causality, observed movement,
absolute permeability, or unique attribution to a landscape mechanism.

---

## Quick reference: the full comparison workflow

```{r qref-mc, eval = FALSE}
library(terradish)

# 1. Build the surface once and reuse it for every candidate model
surface <- conductance_surface(covariates, coords, directions = 8, saveStack = TRUE)

# 2. Profile the graph once; reuse the recommended settings for every fit so
#    models differ only in their formula, never in their optimizer
rec <- terradish_assess_settings(melip.Fst ~ forestcover + altitude,
                                 data = surface)$recommended

fit_full <- terradish(melip.Fst ~ forestcover + altitude, data = surface,
                      conductance_model = loglinear_conductance,
                      measurement_model = mlpe,
                      optimizer = rec$optimizer, control = rec$control)
fit_alt  <- terradish(melip.Fst ~ altitude, data = surface,
                      conductance_model = loglinear_conductance,
                      measurement_model = mlpe,
                      optimizer = rec$optimizer, control = rec$control)

# 3. Nested hypothesis test: does forest cover add likelihood support?
anova(fit_alt, fit_full)          # interpret with nesting and boundary checks

# 4. Rank the whole set; AICc when sites are few relative to parameters
aic_table(
  list(fit_full, fit_alt),
  mod_names = c("Full", "Altitude"),
  AICc = TRUE
)                                 # small delta is a heuristic within this set

# 5. Predictive comparison on held-out sites. terradish_cv() refits from the
#    raw inputs, so it takes points and covariates rather than a fitted object.
cv <- terradish_cv(pts = coords, covariates = covariates,
                   fmla = melip.Fst ~ forestcover + altitude,
                   model = mlpe, prop_train = 2 / 3, seed = 42,
                   fit_full = FALSE)
cv$cv_loglik                      # higher held-out loglik predicts better

# 6. Map the surface with uncertainty
plot(fit_full, type = "surface", data = surface)
```

## Summary of key functions

| Function | Purpose |
|----------|---------|
| `anova(fit1, fit2)` | Likelihood ratio test between two nested models |
| `aic_table(list(...))` | Rank models by AIC, AICc, or BIC |
| `terradish_grid()` | Evaluate log-likelihood on a parameter grid |
| `conductance()` | Fitted conductance with confidence intervals |
| `mlpe_response_change()` | Response-scale change and confidence interval for MLPE pairwise covariates |
| `terradish_cv()` | Single train/test split cross-validation |
| `terradish_cv_replicates()` | Repeated cross-validation across random splits |
| `cv_model_selection()` | Combine CV log-likelihoods with an AIC table |

## See also

- **`vignette("getting-started", package = "terradish")`**, the core workflow
  and how to read a single fitted model.
- **`vignette("spline-conductance", package = "terradish")`**, adding
  non-linear covariate responses to the candidate set compared here.
- **`vignette("ibe-ibr-workflow", package = "terradish")`**, building an
  IBD / IBE / IBR / joint model set to rank with these tools.
- **`vignette("simulation-design", package = "terradish")`**, how design
  choices affect model discrimination under stated simulation scenarios.
