---
title: "Model Comparison and Selection in terradish"
author: "Bill Peterman"
output: 
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Model Comparison and Selection in terradish}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4.5,
  warning  = FALSE,
  message  = FALSE
)

```

## Overview

Once you can fit a `terradish` model (see
`vignette("getting-started", package = "terradish")`), the natural next
questions are:

1. Does forest cover *actually* drive genetic structure, or could that result be
   noise? (**hypothesis testing**)
2. Is a model with both altitude and forest cover better than one with only
   altitude? (**model selection**)
3. How well does the model predict genetic structure at sites not used for
   fitting? (**cross-validation**)

This vignette walks through all three.

---

## Set up: data and baseline model

```{r load-and-fit}
library(terradish)
library(terra)

data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

covariates <- c(melip.altitude, melip.forestcover)
names(covariates) <- c("altitude", "forestcover")
covariates <- scale_covariates(covariates)

surface <- conductance_surface(
  covariates,
  melip.coords,
  directions = 8,
  saveStack = TRUE
)

# Lightweight one-time assessment: reuse recommendations for all fits below.
assessment <- terradish_assess_settings(
  melip.Fst ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  probe_maxit       = 1,
  verbose           = FALSE
)
rec <- assessment$recommended

# Full model: both covariates
fit_full <- terradish(
  melip.Fst ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)
```

### A note on computational settings

When fitting many candidate models — as this vignette does — it pays to profile
the graph and identify the fastest optimizer/solver combination **once**, then
reuse those settings for all fits. `terradish_assess_settings()` does this in a
few short probe iterations:

```{r assess-for-comparison, eval = FALSE}
assessment <- terradish_assess_settings(
  melip.Fst ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  probe_maxit       = 2,
  verbose           = FALSE
)

# One-line summary: did the assessment differ from defaults?
cat(assessment$comparison$summary, "\n")

# Extract recommended settings and reuse for every model in this vignette
rec <- assessment$recommended
```

Then pass `rec$optimizer`, `rec$control`, `rec$solver`, and `rec$solver_control`
to each `terradish()` call below. For the small *Melipona* graph the defaults
are fine; the pattern becomes important for large landscapes where each function
evaluation is expensive. See
`vignette("getting-started", package = "terradish")` for a full explanation of
the output and all available probe options.

---

## Hypothesis testing with likelihood ratio tests

A **likelihood ratio test (LRT)** asks: does adding a covariate significantly
improve the model fit? The test statistic is:

> 2 × (log-likelihood of larger model − log-likelihood of smaller model)

Under the null hypothesis that the extra parameter does not help, this statistic
follows a χ² distribution with degrees of freedom equal to the number of extra
parameters.

`terradish` provides `anova()` for this:

```{r lrt}
# Reduced model: only altitude, no forest cover
fit_altitude_only <- terradish(
  melip.Fst ~ altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)

anova(fit_full, fit_altitude_only)
```

A small p-value (< 0.05) means that removing forest cover significantly
worsens the fit, which is evidence that forest cover is an important predictor of
landscape permeability.

### Testing against the null model of isolation by distance (IBD)

The formula `~ 1` fits a null conductance surface where all cells have equal
conductance. Resistance distance then equals graph distance (weighted only by
geography, not landscape features). This is the IBD baseline:

```{r ibd-test}
fit_ibd <- terradish(
  melip.Fst ~ 1,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)

# Is the full model significantly better than pure IBD?
anova(fit_full, fit_ibd)
```

### Testing for an interaction

An interaction term asks whether the effect of altitude on conductance depends
on forest cover (or vice versa):

```{r interaction-test}
fit_interaction <- terradish(
  melip.Fst ~ forestcover * altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)

# Is the interaction term worth keeping?
anova(fit_full, fit_interaction)
```

> **Tip:** Always compare nested models (smaller vs. larger). The model with
> fewer parameters must be a special case of the model with more parameters.
> `anova()` will warn you if this is not satisfied.

---

## Information criteria: AIC, AICc, and BIC

Likelihood ratio tests require nested models. When comparing non-nested models
(e.g. altitude-only vs. forest-cover-only), use an information criterion
instead.

```{r fit-all-candidates}
fit_fc_only <- terradish(
  melip.Fst ~ forestcover,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)
```

`aic_table()` ranks any set of fitted `terradish` models:

```{r aic-table}
aic_table(
  list(fit_ibd, fit_altitude_only, fit_fc_only, fit_full, fit_interaction),
  mod_names = c("IBD", "Altitude only", "Forestcover only",
                "Full (A + FC)", "Interaction (A × FC)")
)
```

**How to read the table:**

- **logLik**: larger is better (less negative = higher likelihood).
- **df**: number of estimated parameters; more complex models have higher df.
- **AIC**: balances fit and complexity. Smaller is better.
- **delta**: difference from the best model's AIC. Values < 2 indicate
  roughly equivalent support.

### When to use AICc

AIC can over-fit when the number of observations (here, the number of unique
pairwise genetic distances) is small relative to the number of parameters.
AICc adds a second-order correction:

```{r aicc-table}
aic_table(
  list(fit_ibd, fit_altitude_only, fit_fc_only, fit_full, fit_interaction),
  mod_names = c("IBD", "Altitude only", "Forestcover only",
                "Full", "Interaction"),
  AICc = TRUE
)
```

As a rule of thumb, prefer AICc over AIC when the effective sample size
(number of independent pairwise observations) is less than about 40 times the
number of parameters.

### BIC for stricter parsimony

BIC penalizes complexity more heavily than AIC and tends to favor simpler
models:

```{r bic-table}
aic_table(
  list(fit_ibd, fit_altitude_only, fit_fc_only, fit_full, fit_interaction),
  mod_names = c("IBD", "Altitude only", "Forestcover only",
                "Full", "Interaction"),
  BIC = TRUE
)
```

---

## Visualizing the likelihood surface

When a model has only one or two conductance parameters, it is informative to
map how the log-likelihood changes across parameter space. This reveals:

- How sharp (or flat) the optimum is.
- Whether there are multiple local optima.
- Whether the estimates are well-determined or poorly constrained.

`terradish_grid()` evaluates the model at every point on a parameter grid:

```{r likelihood-grid, cache = TRUE}
# Define a 21 x 21 grid of (forestcover, altitude) combinations
theta_grid <- as.matrix(
  expand.grid(
    forestcover = seq(-1, 1, length.out = 21),
    altitude    = seq(-1, 1, length.out = 21)
  )
)

grid_result <- terradish_grid(
  theta     = theta_grid,
  formula   = melip.Fst ~ forestcover + altitude,
  data      = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe
)
```

Plot the result as a heat map with contours:

```{r plot-likelihood-surface, fig.cap = "***Log-likelihood surface. Red dot marks the MLE.***"}
# Build a data frame for plotting
grid_df <- data.frame(
  forestcover = theta_grid[, "forestcover"],
  altitude    = theta_grid[, "altitude"],
  loglik      = grid_result$loglik
)

# Base-R image plot (no ggplot2 dependency)
loglik_mat <- matrix(
  grid_df$loglik,
  nrow = length(unique(grid_df$forestcover)),
  ncol = length(unique(grid_df$altitude))
)

image(
  x    = sort(unique(grid_df$forestcover)),
  y    = sort(unique(grid_df$altitude)),
  z    = loglik_mat,
  xlab = expression(theta[forestcover]),
  ylab = expression(theta[altitude]),
  main = "Log-likelihood surface",
  col  = hcl.colors(64, "Blues", rev = TRUE)
)
contour(
  x    = sort(unique(grid_df$forestcover)),
  y    = sort(unique(grid_df$altitude)),
  z    = loglik_mat,
  add  = TRUE
)
points(coef(fit_full)["forestcover"],
       coef(fit_full)["altitude"],
       pch = 19, col = "red", cex = 1.5)
legend("topright", legend = "MLE", pch = 19, col = "red", bty = "n")
```

A well-behaved likelihood surface has a single, sharp peak at the MLE. A flat
or elongated surface signals that the parameters are poorly identified, often
because the covariates are highly correlated or the signal-to-noise ratio
in the genetic data is low.

For large rasters, a full likelihood grid can become expensive. If the surface
was created with `saveStack = TRUE`, `terradish_grid()` can screen a grid on an
aggregated raster:

```{r likelihood-grid-coarse, eval=FALSE}
grid_coarse <- terradish_grid(
  theta = theta_grid,
  formula = melip.Fst ~ forestcover + altitude,
  data = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  approximation = "coarse_raster",
  approximation_control = list(factor = 2L)
)
```

This is a screening tool rather than an exact replacement for the full grid:
the log-likelihoods are evaluated on the aggregated graph. Use it to locate
interesting regions of parameter space, then rerun a smaller grid or a final
optimization on the full-resolution graph.

---

## Confidence intervals on the conductance surface

`plot(fit, type = "surface", data = surface)` renders the fitted conductance
alongside its asymptotic lower and upper confidence interval bounds in a
single three-panel figure. Uncertainty in θ is propagated through the
log-linear model via the delta method, and all panels share the same color
scale:

```{r ci-conductance, fig.cap = "***Fitted conductance with 95% confidence bounds.***"}
plot(fit_full, type = "surface", data = surface)
```

Wide intervals in a region mean that the available genetic data provide little
constraint on conductance there, either because few sampling sites are nearby
or because the genetic signal is weak.

## Marginal effects on conductance

`plot(fit, type = "marginal", data = surface)` sweeps each covariate across
its observed range while averaging over all other covariates. The shaded ribbon
is a 95% pointwise confidence band from the delta method. Because we used
`scale_covariates()` to prepare the rasters, x-axes are automatically
back-transformed to the original measurement units:

```{r marginal-conductance, fig.cap = "***Marginal effect of each covariate on conductance.***"}
plot(fit_full, type = "marginal", data = surface)
```

A rising curve means higher covariate values increase conductance (more
permeable landscape); a declining curve indicates a barrier effect. Flat or
wide-interval curves suggest that covariate adds little constraint beyond the
other predictors already in the model.

### Handling weakly supported tails

When focal sites sparsely cover a covariate tail, unconstrained marginals can
show unstable extremes that dominate interpretation. Use focal-support clamping
to keep model comparison focused on supported covariate ranges:

```{r support-clamp-marginal, eval = FALSE}
plot(
  fit_full,
  type = "marginal",
  data = surface,
  support = "focal",
  support_probs = c(0.01, 0.99),
  clamp_covariates = c("altitude", "forestcover")
)
```

The same options are available in `conductance()` when producing raster
surfaces for map comparison:

```{r support-clamp-conductance, eval = FALSE}
cond_full_focal <- conductance(
  surface,
  fit_full,
  support = "focal",
  support_probs = c(0.01, 0.99),
  clamp_covariates = c("altitude", "forestcover")
)
```

---

## Cross-validation

Information criteria measure in-sample fit. Cross-validation measures
**out-of-sample** predictive ability, which is often a more honest indicator of
whether a model will generalize to new data.

`terradish_cv()` randomly partitions the focal sampling sites into a training
set and a test set, fits the model on the training set, and evaluates the
held-out log-likelihood on the test set:

```{r cv-single}
cv_result <- terradish_cv(
  pts       = melip.coords,
  covariates = covariates,
  fmla      = melip.Fst ~ forestcover + altitude,
  model     = mlpe,
  prop_train = 2 / 3,   # use 2/3 of sites for training
  seed      = 42,
  fit_full  = FALSE     # skip re-fitting on the full data (saves time)
)

cat("Held-out log-likelihood:", round(cv_result$cv_loglik, 2), "\n")
```

A single split can be noisy; it may depend heavily on which sites happened to
land in the training set. `terradish_cv_replicates()` repeats the process
across many random splits:

```{r cv-replicates}
cv_reps <- terradish_cv_replicates(
  pts        = melip.coords,
  covariates = covariates,
  fmla       = melip.Fst ~ forestcover + altitude,
  model      = mlpe,
  seeds      = 1:5,    # five random splits
  fit_full   = FALSE,
  keep_fits  = FALSE
)

print(cv_reps)
summary(cv_reps)
```

### Cross-validation for model selection

To compare models using held-out log-likelihood, fit each candidate with
`terradish_cv()` or `terradish_cv_replicates()` and then pass the results to
`cv_model_selection()`. In this example, we compare a one-covariate model to
the two-covariate model. A pure intercept-only MLPE model (`~ 1`) can fit on
the full data but still collapse to a no-coefficient boundary on some training
splits, which makes it a fragile teaching example for cross-validation.

```{r cv-selection}
# For speed, use a reduced control with fewer Newton iterations
ctrl <- NewtonRaphsonControl(maxit = 10, verbose = FALSE)

# Subset the raster stack for the simpler model so the example stays quiet
# and only uses the covariate that appears in the formula.
forestcover_only <- covariates[["forestcover"]]

cv_forestcover <- terradish_cv(
  pts = melip.coords, covariates = forestcover_only,
  fmla = melip.Fst ~ forestcover, model = mlpe,
  prop_train = 2 / 3, seed = 42, fit_full = TRUE, control = ctrl
)

cv_full <- terradish_cv(
  pts = melip.coords, covariates = covariates,
  fmla = melip.Fst ~ forestcover + altitude, model = mlpe,
  prop_train = 2 / 3, seed = 42, fit_full = TRUE, control = ctrl
)

cv_comparison <- cv_model_selection(
  list(
    list(train_mod = cv_forestcover$train_mod,
         cv_loglik = cv_forestcover$cv_loglik,
         full_mod  = cv_forestcover$full_mod),
    list(train_mod = cv_full$train_mod,
         cv_loglik = cv_full$cv_loglik,
         full_mod  = cv_full$full_mod)
  ),
  cv_names = c("Forest cover only", "Full (altitude + forestcover)"),
  aic = TRUE
)

cv_comparison$loglik_tab
cv_comparison$AIC_tab
```

---

## Combining the evidence

A complete analysis typically uses all three approaches together:

| Tool | Answers | When to use |
|------|---------|-------------|
| `anova()` | Is covariate X significant? | Nested model comparisons |
| `aic_table()` | Which model is best overall? | Any set of models on the same surface |
| `terradish_cv()` / `cv_model_selection()` | Which model predicts new sites best? | Predictive rather than explanatory goals |
| `mlpe_response_change()` | How large is an MLPE pairwise-covariate effect? | IBE terms fitted with `mlpe_covariates()` |

When `anova()` gives a small p-value *and* AIC selects the same model *and*
cross-validation confirms it has better held-out log-likelihood, you have strong
evidence that the included covariates genuinely predict landscape permeability.

---

## Summary of key functions

| Function | Purpose |
|----------|---------|
| `anova(fit1, fit2)` | Likelihood ratio test between two nested models |
| `aic_table(list(...))` | Rank models by AIC, AICc, or BIC |
| `terradish_grid()` | Evaluate log-likelihood on a parameter grid |
| `conductance()` | Fitted conductance with confidence intervals |
| `mlpe_response_change()` | Response-scale change and confidence interval for MLPE pairwise covariates |
| `terradish_cv()` | Single train/test split cross-validation |
| `terradish_cv_replicates()` | Repeated cross-validation across random splits |
| `cv_model_selection()` | Combine CV log-likelihoods with an AIC table |
