---
title: "Simulating Covariance Data to Plan a Study"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Simulating Covariance Data to Plan a Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4.5,
  warning  = FALSE,
  message  = FALSE
)
```

## Why simulate before you sample?

Landscape-genetic studies are expensive: every additional sampling site means
field time and cost. Before
committing to a design it is worth asking, in advance:

- **How many sampling sites** do I need to detect the landscape effect I predict?
- **How many genetic markers** (the `nu` parameter) do I need?
- **Where should the sites go**: spread evenly across the landscape, or placed
  some other way?
- If I run the analysis, **how often will I recover the truth**, and how often
  will I be fooled by sampling noise?

`terradish` answers these questions by simulation. Because the Wishart
covariance model has a tractable generative form, you can draw synthetic
covariance matrices from a *known* conductance surface, refit the model, and
measure how reliably the truth is recovered, all without any forward-time
genetic simulation. This vignette shows how.

This is a **design tool**, not a replacement for individual-based or
coalescent simulation. It assumes the fitted model is correctly specified and
asks: given that model, how much data do I need? That is exactly the question a
power analysis should answer.

---

## The generative model

A covariance-based `terradish` fit (see
`vignette("wishart-covariance", package = "terradish")`) assumes the observed
genetic covariance matrix `S` is a noisy realization of a structured
covariance:

$$\Sigma = \tau \, E(\theta) + \sigma\, I, \qquad
  \nu\, S \sim \mathrm{Wishart}(\nu, \Sigma)$$

where:

- $E(\theta)$ is the **conductance-implied covariance**: the generalized
  inverse of the graph Laplacian at conductance parameters $\theta$. This is
  the part that carries the landscape signal.
- $\tau \ge 0$ scales that signal.
- $\sigma \ge 0$ is a **nugget** absorbing drift and other non-landscape
  variation.
- $\nu$ is the **effective number of genetic markers**. It controls the amount
  of sampling noise: large $\nu$ gives draws tightly clustered around $\Sigma$;
  small $\nu$ gives noisy draws.

`simulate_covariance_response()` draws from exactly this model, and
`covariance_response_power()` automates the simulate-refit-summarize loop across
many designs. Fitted `wishart_covariance()` models store the nugget internally
on the log scale, but both simulation helpers accept $\sigma$ as the
nonnegative variance added to the diagonal.

For a fixed landscape template, $\tau / \sigma$ is a convenient
**signal-to-nugget ratio**. Larger values place more weight on the
landscape-implied covariance relative to diagonal noise. The ratio should be
interpreted within a fixed template because rescaling $E(\theta)$ also changes
its meaning.

> **Keep `nu` front of mind.** As explained in
> `vignette("wishart-covariance", package = "terradish")`, `nu` acts as an
> *effective sample size*. It does not change the *point estimates* recovered
> from a given covariance matrix, but it controls their precision (standard
> errors scale as $1/\sqrt{\nu}$) and therefore the **statistical power** to
> detect an effect. Sample size (number of sites) and `nu` (number of markers)
> are the two main levers a study designer controls.

---

## Set up a landscape template

We need a landscape to define the conductance surface. Any
`conductance_surface()` works; here we reuse the bundled *Melipona* rasters as
a realistic template. Build the surface with `saveStack = TRUE` so the
simulation tools can subset focal sites efficiently.

```{r template}
library(terradish)
library(terra)

data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

covariates <- c(melip.altitude, melip.forestcover)
names(covariates) <- c("altitude", "forestcover")
covariates <- scale_covariates(covariates)

surface <- conductance_surface(covariates, melip.coords,
                               directions = 8, saveStack = TRUE)

cat("Candidate focal sites:", length(surface$demes), "\n")
```

We also fix the **truth** we want to recover: a moderate positive forest-cover
effect and a moderate negative altitude effect on conductance.
These are illustrative simulation settings, not estimates from the bundled
*Melipona* genetic response. The bundled rasters provide a realistic landscape
template, while the synthetic covariance response is generated from the values
below.

```{r truth}
theta_true <- c(forestcover = 0.6, altitude = -0.4)
tau_true   <- 2   # strength of the landscape signal
sigma_true <- 1   # nugget (non-landscape variance)
```

---

## A single simulated data set

`simulate_covariance_response()` returns one covariance matrix (plus the
ingredients used to build it). This is precisely the object you would otherwise
compute from real genotypes with `cov_from_genetic_data()`.

```{r single-sim}
sim <- simulate_covariance_response(
  theta             = theta_true,
  formula           = ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  tau               = tau_true,
  sigma             = sigma_true,
  nu                = 100,     # pretend we genotyped 100 independent markers
  seed              = 1
)

dim(sim$covariance)        # one row/column per focal site
round(sim$covariance[1:4, 1:4], 3)
```

Refitting recovers the parameters used to generate the data:

```{r single-fit}
fit <- terradish(
  sim$covariance ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = wishart_covariance,
  nu                = 100
)

rbind(true      = theta_true,
      estimated = round(coef(fit), 3))
```

A single replicate only tells you what happened *once*. To characterize a
design you must repeat the draw many times.

---

## The role of markers (`nu`) in statistical power

Holding the sampling design fixed, more markers means more precise estimates
and higher power. The amount of landscape signal also matters. To make both
effects easy to see, we temporarily isolate a modest altitude effect and vary
the signal-to-nugget ratio while holding the landscape template fixed.

```{r nu-power, eval = FALSE}
altitude_surface <- conductance_surface(
  covariates[["altitude"]], melip.coords,
  directions = 8, saveStack = TRUE
)

screen_theta  <- c(altitude = -0.25)
screen_sigma  <- 1
signal_ratios <- c(0.5, 1, 2, 4)

screen_power <- function(sample_size, nu, signal_ratio,
                         nsim = 10, seed = 100) {
  assessment <- covariance_response_power(
    theta        = screen_theta,
    formula      = ~ altitude,
    data         = altitude_surface,
    sample_sizes = sample_size,
    strategies   = "spacefill",
    tau          = signal_ratio * screen_sigma,
    sigma        = screen_sigma,
    nu           = nu,
    nsim         = nsim,
    seed         = seed,
    control      = NewtonRaphsonControl(maxit = 10, verbose = FALSE)
  )
  out <- assessment$parameter_summary
  data.frame(sample_size, nu, signal_ratio,
             fit_rate = out$fit_rate, power = out$power)
}

screen_grid <- function(sample_sizes, nu_values, seed) {
  grid <- expand.grid(sample_size = sample_sizes, nu = nu_values,
                      signal_ratio = signal_ratios)
  do.call(rbind, Map(screen_power,
                     sample_size = grid$sample_size,
                     nu = grid$nu,
                     signal_ratio = grid$signal_ratio,
                     seed = seed + seq_len(nrow(grid))))
}

marker_power <- screen_grid(sample_sizes = 20,
                            nu_values = c(20, 50, 100, 200),
                            seed = 100)
marker_power
```

```{r nu-power-load, echo = FALSE}
# Load pre-computed results (see vignettes/precompute.R to regenerate)
.art <- readRDS("vignette-nu-power.rds")
marker_power  <- .art$marker_power
site_power    <- .art$site_power
signal_ratios <- .art$signal_ratios
screen_nsim   <- .art$screen_nsim
marker_power
```

```{r nu-power-plot, fig.cap = "***Power responds to marker count and the tau / sigma signal ratio.*** Each point is the proportion of 10 simulated data sets in which the altitude coefficient had the correct sign and was significant at alpha = 0.05. The 20 sampling sites are held fixed. The small demonstration grid leaves visible Monte Carlo variation."}
cols <- setNames(c("#7570b3", "#1b9e77", "#d95f02", "#e7298a"),
                 signal_ratios)
plot(NA, xlim = range(marker_power$nu), ylim = c(0, 1), log = "x",
     xlab = "Number of markers (nu)", ylab = "Power to detect altitude",
     main = "Power vs. marker count")
for (signal_ratio in signal_ratios) {
  sub <- marker_power[marker_power$signal_ratio == signal_ratio, ]
  lines(sub$nu, sub$power, type = "b", pch = 19,
        col = cols[as.character(signal_ratio)])
}
abline(h = 0.8, lty = 2, col = "grey50")
legend("bottomright", legend = paste("tau / sigma =", signal_ratios),
       col = cols, lty = 1, pch = 19, bty = "n")
```

The curves make the cost-benefit tradeoff explicit. Extra markers help, but
they cannot fully rescue a weak landscape signal. Conversely, stronger
landscape signal reduces the marker count needed to reach useful power.

For covariance simulation, use `nu` values at least as large as the number of
sampled sites. `simulate_covariance_response()` uses `stats::rWishart()`, which
requires the Wishart degrees of freedom to be at least the covariance-matrix
dimension.

---

## The role of sample size in statistical power

Now hold marker count fixed at `nu = 100` and vary the number of sampling
sites. Each curve again represents a different $\tau / \sigma$ ratio.

```{r site-power, eval = FALSE}
sample_grid <- c(8, 12, 20, 34)
site_power <- screen_grid(sample_sizes = sample_grid,
                          nu_values = 100,
                          seed = 200)
site_power
```

```{r site-power-load, echo = FALSE}
site_power
```

```{r site-power-plot, fig.cap = "***Power responds to sampling sites and the tau / sigma signal ratio.*** Each point is the proportion of 10 simulated data sets in which the altitude coefficient had the correct sign and was significant at alpha = 0.05. Marker count is held fixed at nu = 100. The small demonstration grid leaves visible Monte Carlo variation."}
plot(NA, xlim = range(site_power$sample_size), ylim = c(0, 1),
     xlab = "Number of sampling sites", ylab = "Power to detect altitude",
     main = "Power vs. sample size")
for (signal_ratio in signal_ratios) {
  sub <- site_power[site_power$signal_ratio == signal_ratio, ]
  lines(sub$sample_size, sub$power, type = "b", pch = 19,
        col = cols[as.character(signal_ratio)])
}
abline(h = 0.8, lty = 2, col = "grey50")
legend("bottomright", legend = paste("tau / sigma =", signal_ratios),
       col = cols, lty = 1, pch = 19, bty = "n")
```

Across repeated assessments, more sites generally improve power because the
fitted covariance structure is informed by a larger sample from the landscape.
The ten-replicate teaching grid is intentionally small, so its curves are not
perfectly monotone. Site placement also matters. Real assessments should use
more replicates and compare plausible placement strategies rather than treat
sample count as the only design decision.

---

## Comparing sampling strategies and candidate models

`covariance_response_power()` automates the same loop across **sampling
designs**: it subsets the focal sites in the surface to a target sample size,
optionally using different placement strategies, simulates responses, refits
candidate models, and summarizes recovery. Because all designs reuse one graph,
it is fast.

Here we compare three sample sizes under two placement strategies, and we fit
two candidate models per replicate: the true two-covariate model and a reduced
altitude-only model. This lets us measure both **parameter recovery** and
**model-selection** performance.

```{r power-run, eval = FALSE}
power <- covariance_response_power(
  theta             = theta_true,
  formula           = ~ forestcover + altitude,
  data              = surface,
  sample_sizes      = c(10, 20, 34),
  strategies        = c("spacefill", "random"),
  conductance_model = loglinear_conductance,
  fit_models = list(
    full          = list(formula = ~ forestcover + altitude),
    altitude_only = list(formula = ~ altitude)
  ),
  tau        = tau_true,
  sigma      = sigma_true,
  nu         = 100,
  nsim       = 20,    # increase for smoother estimates in a real study
  n_designs  = 2,     # random designs evaluated per sample size
  seed       = 1,
  control    = NewtonRaphsonControl(maxit = 10, verbose = FALSE)
)

power
```

```{r power-run-load, echo = FALSE}
# Load pre-computed results (see vignettes/precompute.R to regenerate)
power <- readRDS("vignette-power.rds")
power
```

### Scenario-level summary

The `summary` table has one row per (strategy, sample size, model):

```{r power-summary}
power$summary[, c("strategy", "sample_size", "model",
                  "fit_rate", "conductance_power",
                  "mean_conductance_cor", "selected_AIC_rate")]
```

Key columns:

- **`fit_rate`**: fraction of replicates that converged. Values below 1 flag a
  design where the model is hard to fit (often too few sites).
- **`conductance_power`**: fraction of replicates whose fitted conductance
  surface correlates with the truth above `conductance_cor_threshold`
  (default 0.9). This is a holistic "did we recover the surface?" measure that
  works even for spline models where individual coefficients are not
  interpretable.
- **`mean_conductance_cor`**: average correlation between the fitted and true
  conductance surfaces.
- **`selected_AIC_rate`**: how often each model was the AIC winner. For the
  `full` model this is the rate at which the true model was correctly selected.

### Parameter-level summary

`parameter_summary` reports per-coefficient power, bias, RMSE, and confidence
interval coverage:

```{r power-params}
power$parameter_summary[
  power$parameter_summary$model == "full",
  c("strategy", "sample_size", "parameter",
    "power", "bias", "rmse", "coverage")
]
```

What to look for:

- **`power`**: probability of detecting a truly non-zero effect. Aim for
  $\ge 0.8$.
- **`bias`**: average estimate minus truth. Should be near zero; persistent bias
  signals model misspecification or a design too small for the asymptotics to
  hold.
- **`rmse`**: typical estimation error; should shrink as sample size grows.
- **`coverage`**: fraction of 95% confidence intervals that contain the truth.
  Well-calibrated intervals give ~0.95. Coverage well below 0.95 means the
  reported uncertainty is too optimistic for that design.

```{r power-plot, fig.cap = "***Power to recover each conductance coefficient as a function of sample size (spacefill design, true model).*** The dashed line marks 80% power.", fig.height = 4.5}
ps <- power$parameter_summary
ps <- ps[ps$model == "full" & ps$strategy == "spacefill", ]

cols <- c(forestcover = "#1b9e77", altitude = "#d95f02")
plot(NA, xlim = range(ps$sample_size), ylim = c(0, 1),
     xlab = "Number of sampling sites", ylab = "Power",
     main = "Parameter recovery vs. sample size")
for (p in names(cols)) {
  sub <- ps[ps$parameter == p, ]
  sub <- sub[order(sub$sample_size), ]
  lines(sub$sample_size, sub$power, type = "b", pch = 19, col = cols[p])
}
abline(h = 0.8, lty = 2, col = "grey50")
legend("bottomright", legend = names(cols), col = cols,
       lty = 1, pch = 19, bty = "n")
```

---

## Interpreting and acting on the results

A design assessment usually answers one of three questions:

| Question | Column(s) to read | Decision rule |
|----------|-------------------|---------------|
| Enough sites? | `parameter_summary$power`, `conductance_power` | Pick the smallest sample size with power $\ge 0.8$ for the effects you care about. |
| Enough markers? | power-vs-`nu` curves | Pick the smallest `nu` past the power elbow under a conservative $\tau / \sigma$ ratio. |
| Which site placement? | compare `strategy` rows | Prefer the strategy with higher power / surface correlation at equal cost. |
| Trustworthy intervals? | `parameter_summary$coverage` | Require coverage near 0.95 before trusting reported CIs. |

A few practical notes:

- **Increase `nsim`.** This vignette uses small `nsim` so it builds relatively quickly.
  For a real assessment use at least 100-500 simulations per design so that
  power and coverage estimates are stable.
- **Match the simulation to your expectations.** The answer depends on
  `theta`, `tau`, and `sigma`. If you are unsure, run the assessment across a
  grid of plausible effect sizes and $\tau / \sigma$ ratios, then report the
  range.
- **Use a conservative `nu`.** Markers in linkage disequilibrium are not
  independent, so the effective `nu` is smaller than the raw marker count.
  Planning with an optimistic `nu` overstates power.
- **Spline and Gaussian models.** For non-linear conductance models the
  individual coefficients are not interpretable; lean on
  `conductance_power` and `mean_conductance_cor` rather than per-parameter
  power. Pass the relevant factory via `conductance_model` and `fit_models`.

---

## Quick reference

```{r quick-reference, eval = FALSE}
library(terradish)
library(terra)

# 1. Build a landscape template (saveStack = TRUE enables site subsetting)
surface <- conductance_surface(covariates, coords,
                               directions = 8, saveStack = TRUE)

# 2. One simulated covariance matrix from a known truth
sim <- simulate_covariance_response(
  theta = c(x1 = 0.6, x2 = -0.4), formula = ~ x1 + x2,
  data = surface, tau = 2, sigma = 1, nu = 100, seed = 1
)

# 3. Refit to check recovery
fit <- terradish(sim$covariance ~ x1 + x2, data = surface,
                 conductance_model = loglinear_conductance,
                 measurement_model = wishart_covariance, nu = 100)
coef(fit)

# 4. Full design assessment across sample sizes / strategies / models
power <- covariance_response_power(
  theta = c(x1 = 0.6, x2 = -0.4), formula = ~ x1 + x2,
  data = surface, sample_sizes = c(10, 20, 30),
  strategies = c("spacefill", "random"),
  tau = 2, sigma = 1, nu = 100, nsim = 200, seed = 1
)
power$summary            # fit rate, surface recovery, model selection
power$parameter_summary  # power, bias, RMSE, coverage per coefficient
```

## See also

- `vignette("wishart-covariance", package = "terradish")`: the covariance
  workflow and a full explanation of `nu`.
- `vignette("getting-started", package = "terradish")`: the core modeling
  workflow.
- `vignette("model-comparison", package = "terradish")`: AIC, likelihood
  ratio tests, and cross-validation for real data.
