---
title: "Spline Conductance Models in terradish"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Spline Conductance Models in terradish}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 6,
  fig.height = 4.5,
  warning   = FALSE,
  message   = FALSE
)
```

## Why use splines for conductance?

The default `terradish` conductance model is log-linear:

$$C_i = \exp(\theta_1 x_{i1} + \theta_2 x_{i2} + \ldots)$$

This is often the right first model. It is stable, easy to interpret, and asks a
simple question: does conductance increase or decrease as a raster covariate
increases?

Ecological responses are not always that simple. Movement may be easiest at
intermediate elevation, or a predictor may have little effect until a threshold
is crossed. A straight line on the log-conductance scale cannot represent those
shapes.

`smooth_loglinear_conductance()` extends the log-linear model by expanding
selected raster covariates into spline basis columns before the conductance
surface is evaluated. The model is still a positive log-link conductance model:

$$C_i = \exp(B_i \theta)$$

where `B_i` is the row of the expanded design matrix for raster cell `i`. The
only change is that selected variables enter through basis columns rather than
one raw linear column.

This gives `terradish` a GAM-like conductance surface while keeping the usual
resistance-distance likelihood, MLPE measurement model, model comparison tools,
and plotting workflow.

---

## What is implemented?

Spline conductance models are created with `smooth_loglinear_conductance()`.
Smooth terms are written in the conductance formula with `s()`:

```{r smooth-syntax, eval = FALSE}
smooth_loglinear_conductance(
  ~ forestcover + s(altitude, df = 4),
  x = surface$x
)
```

The supported pieces are intentionally small and explicit:

- `s(variable, df = 4)` creates a spline basis for one numeric raster layer.
- `s(variable, k = 4)` is an alias for `df = 4`.
- `basis = "ns"` uses natural splines from `splines::ns()`.
- `basis = "bs"` uses B-splines from `splines::bs()`.
- Ordinary terms can be mixed with smooth terms, as in
  `~ forestcover + s(altitude, df = 4)`.

This first implementation does **not** estimate a smoothing penalty. The chosen
`df` or `k` controls the flexibility of the effect. For that reason, spline
models should usually be introduced in small steps and compared against simpler
linear models.

Two practical constraints follow from the implementation:

- Smooth terms currently support one raw numeric column name, such as
  `s(altitude)`, rather than transformations like `s(I(altitude^2))`.
- Derivatives are taken with respect to the expanded basis coefficients. The
  optimizer sees an ordinary log-linear conductance model after expansion.

---

## Worked example with the melip data

We will use the bundled *Melipona* bee example data. The goal is not to declare
a final ecological model from a tiny subset of the data. The goal is to show the
full spline workflow:

1. prepare raster covariates and sampling locations
2. fit a linear baseline model
3. fit one smooth effect
4. fit two smooth effects
5. compare the models
6. inspect fitted conductance and marginal effects

To keep the vignette fast, we use 12 sampling sites.

```{r load}
library(terradish)
library(terra)

data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

keep <- 1:12
melip.Fst <- melip.Fst[keep, keep, drop = FALSE]
coords <- melip.coords[keep]
```

### Prepare the raster covariates

Spline basis construction is sensitive to the scale and range of the covariate.
As in the standard `terradish` workflow, we center and scale continuous raster
layers before fitting:

```{r prepare-covariates}
covariates <- c(melip.altitude, melip.forestcover)
names(covariates) <- c("altitude", "forestcover")

covariates <- scale_covariates(covariates)

surface <- conductance_surface(
  covariates,
  coords,
  directions = 8,
  saveStack = TRUE
)
```

The `surface$x` data frame contains the covariate values for active graph
vertices. This is the object used internally to create the conductance design
matrix.

```{r inspect-surface}
head(surface$x)
```

---

## Step 1: fit a linear baseline

A spline model is easiest to interpret when it is compared to the linear model
it generalizes. We start with the usual log-linear conductance model:

```{r fit-linear}
control <- NewtonRaphsonControl(maxit = 8, verbose = FALSE)

fit_linear <- terradish(
  melip.Fst ~ altitude + forestcover,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  control           = control
)

coef(fit_linear)
```

In this model, each raster contributes one conductance parameter. A positive
coefficient means higher covariate values increase conductance and reduce
resistance distance; a negative coefficient means higher values act more like a
barrier.

---

## Step 2: fit one smooth effect

Now let altitude have a non-linear effect while keeping forest cover linear:

```{r fit-spline-altitude}
fit_spline_altitude <- terradish(
  melip.Fst ~ forestcover + s(altitude, df = 4),
  data              = surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  control           = control
)

coef(fit_spline_altitude)
```

The coefficient names show what changed. Instead of one `altitude` coefficient,
the fitted model has several `s(altitude).*` coefficients. These are not meant
to be interpreted one at a time. They define the shape of the altitude effect
together.

We can inspect the corresponding conductance model directly:

```{r inspect-model}
smooth_model <- smooth_loglinear_conductance(
  ~ forestcover + s(altitude, df = 4),
  surface$x
)

attr(smooth_model, "default")
attr(smooth_model, "smooth_loglinear_info")
```

The default parameter vector has one entry per expanded design column. When
`terradish()` fits a spline conductance model, these are the parameters it
optimizes.

---

## Step 3: fit two smooth effects

More flexibility is not automatically better. Still, it is useful to see the
syntax for a model where both raster covariates have smooth effects:

```{r fit-spline-both}
fit_spline_both <- terradish(
  melip.Fst ~ s(altitude, df = 4) + s(forestcover, df = 4),
  data              = surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  control           = control
)

coef(fit_spline_both)
```

This model estimates eight conductance basis coefficients before the MLPE
measurement parameters are considered. That is a lot of flexibility for 12
sampling sites, so model comparison and visual diagnostics are essential.

---

## Step 4: compare spline and linear models

Spline models can be compared with the same likelihood tools used elsewhere in
`terradish`. Because these candidate models have different numbers of
parameters, AIC or AICc is often more useful than a likelihood ratio test.

```{r compare-models}
aic_table(
  list(fit_linear, fit_spline_altitude, fit_spline_both),
  mod_names = c("Linear", "Altitude spline", "Altitude + forestcover splines"),
  AICc = TRUE
)
```

Read this as a first-pass diagnostic, not as a universal rule that splines
should always be retained or rejected. A spline model is most compelling when it
both improves the information criterion and produces a biologically plausible
effect curve.

---

## Step 5: plot the fitted conductance surface

The fitted conductance surface is still obtained with `conductance()`:

```{r conductance-raster, fig.cap = "***Fitted conductance surface from the altitude-spline model. Sampling sites are shown in red.***"}
spline_conductance <- conductance(surface, fit_spline_altitude)

plot(spline_conductance, main = "Altitude spline conductance")
points(coords, pch = 19, cex = 0.8, col = "red")
```

This map shows the combined effect of all terms in the conductance model. In the
altitude-spline model, that means the smooth altitude effect plus the linear
forest-cover effect.

---

## Step 6: plot marginal effects on the conductance scale

Spline coefficients are hard to interpret directly. The effect curve is usually
the right diagnostic:

```{r marginal-plot, fig.cap = "***Marginal conductance-scale effects from the altitude-spline model.***"}
plot(
  fit_spline_altitude,
  type = "marginal",
  data = surface,
  n = 50
)
```

For spline-expanded models, marginal plots are shown against the original
covariate, such as `altitude`, rather than against individual basis columns.
That is the easiest way to ask whether the fitted shape makes ecological sense.

You can also inspect the same marginal shape on the response scale:

```{r marginal-response-plot, fig.cap = "***Marginal response-scale effects from the altitude-spline model.***"}
plot(
  fit_spline_altitude,
  type = "marginal_response",
  data = surface,
  n = 50
)
```

Use these plots as shape diagnostics. With small datasets, apparently complex
curves can be unstable even when the optimizer converges.

---

## Natural splines and B-splines

By default, `smooth_loglinear_conductance()` uses natural splines. Natural
splines are often a good default because they are constrained to be linear
beyond their boundary knots, which reduces edge behavior.

B-splines are also available:

```{r bs-model}
fit_bs_altitude <- terradish(
  melip.Fst ~ forestcover + s(altitude, k = 4, basis = "bs", degree = 2),
  data              = surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  control           = control
)

coef(fit_bs_altitude)
```

Here `k = 4` requests four basis columns and `degree = 2` requests a quadratic
B-spline basis. The same caution applies: choose small basis dimensions unless
the data strongly support more flexibility.

---

## A small recoverability check

The package also ships a longer example script that simulates a covariance
response from a known non-linear conductance surface and then compares linear
and spline recovery:

```{r recovery-script-path}
system.file("examples", "spline-recovery-melip.R", package = "terradish")
```

That script is intended for interactive exploration from RStudio:

```{r recovery-script, eval = FALSE}
source(system.file("examples", "spline-recovery-melip.R", package = "terradish"))

recovery <- run_spline_recovery_example()
recovery$model_comparison

melip_spline <- fit_melip_spline_example()
melip_spline$model_comparison
```

The simulation part is deliberately not run inside this vignette because it is
more expensive than the basic worked example. It is useful when you want to
check whether a candidate spline specification can recover a known non-linear
conductance surface under controlled conditions.

---

## Practical workflow

For real analyses, use spline conductance models as a staged extension of the
linear workflow:

1. Fit a simple IBD or linear conductance baseline.
2. Add one smooth term to the covariate with the clearest biological rationale.
3. Keep `df` or `k` small for first-pass model comparison.
4. Compare against the simpler model with AICc or held-out prediction.
5. Inspect the marginal curve before interpreting the result.
6. Add additional smooth terms only if the one-smooth model is stable.

This keeps the spline model tied to an interpretable landscape-genetics
question: does the landscape effect look meaningfully non-linear, and is that
extra flexibility supported by the data?

