---
title: "Spline Conductance Models in terradish"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Spline Conductance Models in terradish}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  warning   = FALSE,
  message   = FALSE
)

```

## Why splines for conductance?

The default `terradish` conductance model is **log-linear**:

$$C_i = \exp(\theta_1 x_{i1} + \theta_2 x_{i2} + \cdots)$$

This model asks a simple binary question for each landscape covariate: does
conductance go up or down as the covariate increases? That is exactly the right
question for many problems. Forest cover, for example, often has a monotone
positive effect on connectivity for forest-dependent species: higher cover,
higher conductance, end of story.

But ecological responses to landscape features are often more complex. Consider:

- **Elevation**: Movement may be easiest at intermediate elevations; high mountain
  peaks and valley floors are both barriers, but for different reasons (snow vs.
  flooding vs. predation risk).
- **Slope**: Slight grades may be navigable; steep terrain may become a barrier
  only past some threshold.
- **Temperature**: Many ectotherms have a thermal optimum; conductance may be
  highest at intermediate temperatures and fall off steeply in both directions.
- **Canopy density**: Too little or too much canopy may suppress movement for a
  species that needs dappled light.

None of these shapes can be captured by a single linear term on the log scale. A
negative coefficient makes the covariate a monotone barrier; a positive
coefficient makes it a monotone highway. There is no room for "best in the
middle."

`smooth_loglinear_conductance()` extends the log-linear model by expanding
selected covariates into **spline basis functions**. The model structure is still:

$$C_i = \exp(B_i \boldsymbol{\theta})$$

where $B_i$ is a row of the **expanded design matrix** for raster cell $i$. For
covariates not wrapped in `s()`, $B_i$ contains the raw value. For covariates
wrapped in `s()`, $B_i$ contains several basis-function evaluations instead of
one raw value. The optimizer still sees an ordinary log-linear model; the
splines live entirely inside the design matrix, which is computed once and then
treated as fixed.

---

## What are spline basis functions?

A spline is a piecewise polynomial function that is constrained to be smooth
(continuous and differentiable) at the join points, called **knots**. Instead of
estimating a single slope for altitude, we estimate a weighted combination of
several basis functions, each capturing part of the altitude range.

The figure below shows what natural spline and B-spline basis functions look like
for a covariate that runs from −2 to 2 (as after `scale_covariates()`):

```{r basis-demo, fig.cap = "***Natural spline (ns) and B-spline (bs) basis functions for a standardized covariate.*** Each colored line is one basis column. The fitted conductance effect is a weighted combination of all basis functions in a panel.", fig.height = 5}
library(splines)
library(terradish)

x_seq <- seq(-2, 2, length.out = 300)

# Natural spline basis with df = 4
B_ns <- ns(x_seq, df = 4)

# B-spline basis with k = 4 columns, degree 3 (cubic)
B_bs <- bs(x_seq, df = 4, degree = 3)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

matplot(x_seq, B_ns, type = "l", lty = 1, lwd = 2,
        col = c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3"),
        xlab = "Standardized covariate", ylab = "Basis value",
        main = "Natural spline (ns, df = 4)",
        ylim = c(-0.4, 1))
abline(h = 0, col = "grey70", lty = 2)
legend("topright", legend = paste0("basis ", 1:4),
       col = c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3"),
       lty = 1, lwd = 2, bty = "n", cex = 0.8)

matplot(x_seq, B_bs, type = "l", lty = 1, lwd = 2,
        col = c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3"),
        xlab = "Standardized covariate", ylab = "Basis value",
        main = "B-spline (bs, df = 4, degree = 3)",
        ylim = c(-0.4, 1))
abline(h = 0, col = "grey70", lty = 2)
legend("topright", legend = paste0("basis ", 1:4),
       col = c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3"),
       lty = 1, lwd = 2, bty = "n", cex = 0.8)
```

Each basis column covers a different part of the covariate range. The optimizer
assigns a coefficient to each column; the weighted sum of all columns defines the
shape of the effect. A hump in the middle of the covariate range is captured by
upweighting the middle basis columns relative to the edge ones.

### Natural splines vs. B-splines

Both produce smooth curves, but they differ in behavior outside the observed
data range:

| Property | Natural spline (`ns`) | B-spline (`bs`) |
|---|---|---|
| Behavior at edges | **Linear** beyond boundary knots | Polynomial (degree-dependent); can extrapolate wildly |
| Partition of unity | No | Yes (basis sums to 1 everywhere) |
| Parameter count for `df = k` | `k` columns | `k` columns (knots placed to give `k`) |
| Default degree | N/A (cubic, fixed) | Cubic (`degree = 3`); user-adjustable |
| Best for | General use; safe extrapolation | When you know the curve may be non-linear at the edges too |

For most landscape-genetics applications, **natural splines are likely the safer
default**. The landscape feature you observe at sampling sites is what matters;
extrapolation is usually not meaningful for raster-based conductance models
because the graph only covers cells with observed covariate values. Still,
B-splines are useful when you have strong prior knowledge that the edge behavior
is non-linear, or when you want finer control over the spline degree.

### How many basis functions (df / k)?

More basis functions → more shape flexibility → more parameters to estimate.
With sparse spatial genetic data, overfitting is a real risk. Guidelines:

- **df = 3 or 4** is a good starting point for most covariates. This allows a
  single hump, valley, or threshold-like shape.
- **df = 2** is essentially a quadratic effect; it adds one parameter over the
  linear model and tests for a simple non-linear shape.
- **df ≥ 6** should be reserved for cases where both the data and a clear
  ecological hypothesis support it; compare against simpler models with AICc.
- Use **AICc** rather than AIC when sample sizes (number of population pairs)
  are small relative to the number of parameters.

---

## Data preparation

The rest of this vignette uses the bundled *Melipona* bee example with all 34
sampling sites.

```{r load-data}
library(terradish)
library(terra)

data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

cat("Number of sampling sites:", nrow(melip.coords), "\n")
cat("Dimension of Fst matrix: ", dim(melip.Fst), "\n")
```

Scale the raster covariates. This step is important for spline models because:
(1) it brings both layers to a comparable scale so spline knots are placed
sensibly relative to the data density, and (2) it helps the optimizer converge
from a common starting point.

```{r prepare-covariates}
covariates <- c(melip.altitude, melip.forestcover)
names(covariates) <- c("altitude", "forestcover")

# Center and scale each layer to mean 0, sd 1
covariates <- scale_covariates(covariates)

# Build the conductance surface graph
# saveStack = TRUE is required for conductance() and plot(type = "surface")
surface <- conductance_surface(
  covariates,
  melip.coords,
  directions = 8,
  saveStack  = TRUE
)

cat("Graph vertices:", nrow(surface$x), "\n")
head(surface$x)
```

---

## Checking settings before spline fitting

`terradish_assess_settings()` is worth calling before fitting spline conductance
models for an additional reason beyond raw graph size: **spline formulas expand
the conductance parameter count**. A formula with `s(altitude, df = 4)` has four
basis coefficients for altitude alone, plus any other terms. With enough
parameters, Newton-Raphson can become slower than BFGS per iteration. The
assessment handles this automatically; it probes the expanded model, not the
raw formula, and reports the smooth-term expansion in the `$profile` and in its
notes.

```{r assess-spline, eval = FALSE}
# Assessment for the altitude natural-spline model (df = 4)
# Uses the same surface built above (saveStack = TRUE required for solver probes)
assessment_spline <- terradish_assess_settings(
  melip.Fst ~ forestcover + s(altitude, df = 4),
  data              = surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  probe_maxit       = 2,
  verbose           = TRUE
)

assessment_spline
```

The printed summary will include a note such as:

```
Notes:
  - Smooth conductance formula expanded to 5 conductance parameters across
    1 smooth term (4 spline basis columns).
```

This confirms that the formula was evaluated on the expanded design matrix (one
linear forest-cover column plus four altitude basis columns = five parameters
total), and that probes were run with that full parameter count.

Spline expansions can make extreme starting values more risky because the
expanded linear predictor is exponentiated into conductance. If a spline fit
stops with a `CHOLMOD` warning or a `non-positive-definite reduced Laplacian`
message, first scale the input rasters, restart with coefficients closer to
zero, and check the retained graph cells for missing or infinite covariate
values. During diagnosis, rerun `terradish_assess_settings(..., verbose = TRUE)`
and consider `solver = "auto"` or `solver = "amg"` to separate direct-solver
factorization problems from the conductance specification.

### Passing `$recommended` to each candidate fit

When the assessment recommends a non-default optimizer (e.g. `bfgs` instead of
`newton` because the parameter count after expansion is high), apply it to all
candidate models to keep comparisons fair:

```{r assess-spline-use, eval = FALSE}
# Extract recommended optimizer and control once
rec <- assessment_spline$recommended

fit_ns_assessed <- terradish(
  melip.Fst ~ forestcover + s(altitude, df = 4),
  data              = surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control
)
```

For this vignette we run a lightweight assessment (`probe_maxit = 1`) and reuse
its recommendations across the candidate fits so the workflow is exercised in
real code while staying fast.

---

## Model fitting: from linear to spline

We fit four candidate models, moving from most constrained to most flexible:

1. **Linear baseline**: one parameter per covariate, standard log-linear
2. **Altitude spline (ns)**: altitude enters as a natural spline; forest cover
   remains linear
3. **Both splines (ns)**: both covariates enter as natural splines
4. **Altitude spline (bs)**: altitude as a B-spline; forest cover linear (for
   comparison with model 2)

We use a small `maxit` to keep the vignette fast; for real analyses, allow more
iterations.

```{r fit-models}
assessment_spline_light <- terradish_assess_settings(
  melip.Fst ~ forestcover + s(altitude, df = 4),
  data              = surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  probe_maxit       = 1,
  verbose           = FALSE
)
rec <- assessment_spline_light$recommended

# 1. Log-linear baseline (one parameter per covariate)
fit_linear <- terradish(
  melip.Fst ~ altitude + forestcover,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)

# 2. Natural spline on altitude, linear forest cover (df = 4)
fit_ns_alt <- terradish(
  melip.Fst ~ forestcover + s(altitude, df = 4),
  data              = surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)

# 3. Natural splines on both covariates (df = 4 each)
fit_ns_both <- terradish(
  melip.Fst ~ s(altitude, df = 4) + s(forestcover, df = 4),
  data              = surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)

# 4. Cubic B-spline on altitude, linear forest cover (k = 4 columns, degree 3)
fit_bs_alt <- terradish(
  melip.Fst ~ forestcover + s(altitude, k = 4, basis = "bs", degree = 3),
  data              = surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)
```

---

## Inspecting the design matrix

Before examining fits, it is instructive to look at what `smooth_loglinear_conductance()` produces:

```{r inspect-design}
# Build the smooth model object for the altitude-spline specification
smooth_spec <- smooth_loglinear_conductance(
  ~ forestcover + s(altitude, df = 4),
  surface$x
)

# Starting parameter vector (one entry per design column)
cat("Parameter names:\n")
names(attr(smooth_spec, "default"))

# Smooth term details
cat("\nSmooth term metadata:\n")
str(attr(smooth_spec, "smooth_loglinear_info"))
```

The `s(altitude).*` parameters are the basis coefficients for altitude. Taken
individually they are not interpretable. What matters is their combined effect
on the conductance surface, which is best visualized through marginal plots.

---

## Estimated conductance parameters

```{r show-coefs}
cat("--- Linear baseline ---\n")
print(round(coef(fit_linear), 4))

cat("\n--- Altitude ns spline ---\n")
print(round(coef(fit_ns_alt), 4))

cat("\n--- Both ns splines ---\n")
print(round(coef(fit_ns_both), 4))

cat("\n--- Altitude bs spline ---\n")
print(round(coef(fit_bs_alt), 4))
```

The spline coefficients for `s(altitude)` are not directly comparable to the
single altitude coefficient in the linear model. The linear model distills the
altitude effect to one number (positive = conductance increases with elevation,
negative = conductance decreases). The spline model allows the effect to change
direction across the altitude range. The marginal plots in the next sections
reveal the actual shape.

---

## Model comparison

Use AICc (small-sample-corrected AIC) to account for the different number of
parameters across models:

```{r aic-table}
aic_table(
  list(fit_linear, fit_ns_alt, fit_bs_alt, fit_ns_both),
  mod_names = c(
    "Linear",
    "Altitude ns (df=4)",
    "Altitude bs (k=4)",
    "Both ns (df=4 each)"
  ),
  AICc = TRUE
)
```

Interpret this table as a first-pass diagnostic, not a definitive verdict:

- **ΔAICc < 2** from the top model: models that are empirically indistinguishable
  given the data. Do not over-interpret small differences.
- **ΔAICc 2-7**: Some support for the simpler model; spline flexibility is not
  clearly warranted.
- **ΔAICc > 7**: Strong support for the better model.

An AICc improvement must be accompanied by an ecologically plausible marginal
effect curve. A spline with a lower AICc but a biologically implausible shape
(oscillating wildly, extreme at the edges) should be treated with skepticism.

---

## Marginal conductance effects

Marginal plots show what happens to conductance as one covariate changes while
all others are held at their mean. For spline models this is the primary
diagnostic tool, because the basis coefficients themselves are uninterpretable.

### Natural vs. B-spline altitude effect

Use `marginal_covariates = "altitude"` to extract only the altitude panel from
each model, then combine the ggplot objects side-by-side with `patchwork`:

```{r marginal-ns-vs-bs, fig.cap = "***Altitude marginal effect on the conductance scale.*** Left: natural spline (ns). Right: B-spline (bs). Both use df/k = 4 and degree 3.", fig.height = 4.5}
library(patchwork)
library(ggplot2)

p_ns <- plot(fit_ns_alt, type = "marginal", data = surface, n = 80,
             marginal_covariates = "altitude") +
  ggtitle("Altitude: ns (df = 4)")

p_bs <- plot(fit_bs_alt, type = "marginal", data = surface, n = 80,
             marginal_covariates = "altitude") +
  ggtitle("Altitude: bs (k = 4)")

p_ns | p_bs
```

Both spline families should produce broadly similar effect shapes when the signal
is genuine. Large discrepancies between ns and bs estimates at the same df/k are
a warning sign: the spline is fitting noise rather than signal, and a simpler
model may be preferable.

### All four model marginal effects (altitude)

Setting `marginal_covariates = "altitude"` keeps only the altitude panel from
each model (which may otherwise also show forest cover). This makes it easy to
arrange all four altitude curves in a single 2 × 2 grid for direct comparison:

```{r marginal-all, fig.cap = "***Altitude marginal effect across all four candidate models.*** The linear model constrains the effect to a straight line (on the log-conductance scale). Both spline models can represent non-monotone shapes.", fig.height = 6}
p_lin   <- plot(fit_linear,  type = "marginal", data = surface, n = 80,
                marginal_covariates = "altitude") +
  ggtitle("Linear baseline")

p_ns_a  <- plot(fit_ns_alt,  type = "marginal", data = surface, n = 80,
                marginal_covariates = "altitude") +
  ggtitle("Altitude ns (df = 4)")

p_ns_b  <- plot(fit_ns_both, type = "marginal", data = surface, n = 80,
                marginal_covariates = "altitude") +
  ggtitle("Both ns (df = 4): altitude panel")

p_bs_a  <- plot(fit_bs_alt,  type = "marginal", data = surface, n = 80,
                marginal_covariates = "altitude") +
  ggtitle("Altitude bs (k = 4)")

(p_lin | p_ns_a) / (p_ns_b | p_bs_a)
```

### Response-scale marginal effects

The `type = "marginal_response"` plot shows the same effect on the **response
scale** (genetic distance or log-likelihood units, depending on the measurement
model), together with predictive uncertainty bands. This is useful for
communicating results to an ecological audience that may not think in terms of
log-conductance.

Comparing the ns and bs altitude response-scale effects side-by-side gives an
additional consistency check: if the two families agree, the non-linear shape
has genuine support; if they diverge, the spline is likely fitting noise.

```{r marginal-response, fig.cap = "***Marginal response-scale effects from the altitude natural-spline model.*** Shaded bands are approximate pointwise 95% predictive intervals. Left: natural spline. Right: B-spline.", fig.height = 4.5}
p_resp_ns <- plot(fit_ns_alt, type = "marginal_response", data = surface,
                  n = 80, marginal_covariates = "altitude") +
  ggtitle("ns (df = 4): response scale")

p_resp_bs <- plot(fit_bs_alt, type = "marginal_response", data = surface,
                  n = 80, marginal_covariates = "altitude") +
  ggtitle("bs (k = 4): response scale")

p_resp_ns | p_resp_bs
```

### Robust tails with focal-support clamping

Spline models are especially sensitive to sparse support at covariate tails.
To keep spline marginals and predicted conductance maps anchored to focal-site
support, use `support = "focal"` with quantile bounds:

```{r spline-support-clamp, eval = FALSE}
plot(
  fit_ns_alt,
  type = "marginal",
  data = surface,
  n = 80,
  support = "focal",
  support_probs = c(0.01, 0.99),
  clamp_covariates = "altitude"
)

cond_ns_alt_focal <- conductance(
  surface,
  fit_ns_alt,
  support = "focal",
  support_probs = c(0.01, 0.99),
  clamp_covariates = c("altitude", "forestcover")
)
```

---

## Observed vs. fitted genetic distances

```{r obs-vs-fitted, fig.cap = "***Observed vs. fitted pairwise Fst for all candidate models.*** Each point is a pair of sampling sites. The diagonal (dashed) represents a perfect fit.", fig.height = 5}
p_fit_linear <- plot(fit_linear,  type = "fit", main = "Linear baseline")
p_fit_ns_alt <- plot(fit_ns_alt,  type = "fit", main = "Altitude ns (df = 4)")
p_fit_ns_both <- plot(fit_ns_both, type = "fit", main = "Both ns (df = 4)")
p_fit_bs_alt <- plot(fit_bs_alt,  type = "fit", main = "Altitude bs (k = 4)")

(p_fit_linear | p_fit_ns_alt) / (p_fit_ns_both | p_fit_bs_alt)
```

All four panels should show a similar cloud pattern because all models use the
same data and MLPE correlation structure. Systematic deviations from the diagonal
(e.g. all large-distance pairs systematically under-predicted) would indicate
model misspecification beyond the conductance formula.

---

## Estimated conductance surfaces

The conductance surface shows the per-cell conductance value across the
landscape. `conductance()` returns a three-layer SpatRaster with point estimate
and 95% confidence bounds (delta-method).

### Altitude natural-spline model

```{r surface-ns-alt, fig.cap = "***Estimated conductance surface from the altitude natural-spline model.*** The three panels show the point estimate and 95% CI bounds. Red points are sampling sites.", fig.height = 4}
cond_ns_alt <- conductance(surface, fit_ns_alt)

plot(cond_ns_alt,
     main = c("ns(altitude): estimate",
              "ns(altitude): lower 95%",
              "ns(altitude): upper 95%"))
points(melip.coords, pch = 19, cex = 0.6, col = "red")
```

### Comparison: linear vs. altitude spline

```{r surface-compare, fig.cap = "***Point-estimate conductance surfaces: linear model (left) vs. altitude natural-spline model (right).*** Differences in surface texture reflect the non-linear altitude effect captured by the spline.", fig.height = 4}
cond_linear <- conductance(surface, fit_linear)

par(mfrow = c(1, 2))
plot(cond_linear[[1]], main = "Linear baseline")
points(melip.coords, pch = 19, cex = 0.6, col = "red")

plot(cond_ns_alt[[1]], main = "Altitude ns (df = 4)")
points(melip.coords, pch = 19, cex = 0.6, col = "red")
```

Differences between the two surfaces are most informative at intermediate
elevations where the spline effect departs most from the linear extrapolation.

---

## Natural splines: varying degrees of freedom

A key modeling decision is how many basis functions to use. More df → more
flexibility → more parameters. This section illustrates the effect of df on the
fitted altitude curve using natural splines.

```{r ns-df-comparison, fig.cap = "***Natural spline altitude effect for df = 2, 3, 4, and 6.*** Higher df allows more shape complexity. Compare the smoothness and ecological plausibility of each curve before selecting a value.", fig.height = 6}
fit_ns2 <- terradish(
  melip.Fst ~ forestcover + s(altitude, df = 2),
  data              = surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)

fit_ns3 <- terradish(
  melip.Fst ~ forestcover + s(altitude, df = 3),
  data              = surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)

fit_ns6 <- terradish(
  melip.Fst ~ forestcover + s(altitude, df = 6),
  data              = surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)

# Use marginal_covariates = "altitude" to pull only the altitude panel from
# each fit, then tile the four ggplot objects with patchwork.
p2 <- plot(fit_ns2,    type = "marginal", data = surface, n = 80,
           marginal_covariates = "altitude") +
  ggtitle("ns, df = 2 (quadratic-like)")

p3 <- plot(fit_ns3,    type = "marginal", data = surface, n = 80,
           marginal_covariates = "altitude") +
  ggtitle("ns, df = 3")

p4 <- plot(fit_ns_alt, type = "marginal", data = surface, n = 80,
           marginal_covariates = "altitude") +
  ggtitle("ns, df = 4")

p6 <- plot(fit_ns6,    type = "marginal", data = surface, n = 80,
           marginal_covariates = "altitude") +
  ggtitle("ns, df = 6")

(p2 | p3) / (p4 | p6)
```

```{r ns-df-aic}
aic_table(
  list(fit_linear, fit_ns2, fit_ns3, fit_ns_alt, fit_ns6),
  mod_names = c("Linear", "ns df=2", "ns df=3", "ns df=4", "ns df=6"),
  AICc = TRUE
)
```

**Reading the df comparison**: Start from the most constrained model (df = 2)
and add flexibility only if the AICc improves meaningfully *and* the shape is
ecologically plausible. Oscillating or extreme-at-boundary curves with high df
are likely fitting noise.

---

## B-spline degree exploration

B-splines allow control over the polynomial degree separately from the number of
basis columns. Degree 1 is piecewise linear, degree 2 is quadratic, and degree 3
(the default) is cubic. Higher degrees produce smoother but more globally
coupled curves.

```{r bs-degree-comparison, fig.cap = "***B-spline altitude effect for degree 1, 2, and 3 (k = 4 basis columns each).*** Degree 1 is piecewise linear; degree 3 is cubic. Smoother curves do not always fit better.", fig.height = 4}
fit_bs1 <- terradish(
  melip.Fst ~ forestcover + s(altitude, k = 4, basis = "bs", degree = 1),
  data              = surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)

fit_bs2 <- terradish(
  melip.Fst ~ forestcover + s(altitude, k = 4, basis = "bs", degree = 2),
  data              = surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)

pd1 <- plot(fit_bs1,    type = "marginal", data = surface, n = 80,
            marginal_covariates = "altitude") +
  ggtitle("bs, degree = 1 (piecewise linear)")

pd2 <- plot(fit_bs2,    type = "marginal", data = surface, n = 80,
            marginal_covariates = "altitude") +
  ggtitle("bs, degree = 2 (quadratic)")

pd3 <- plot(fit_bs_alt, type = "marginal", data = surface, n = 80,
            marginal_covariates = "altitude") +
  ggtitle("bs, degree = 3 (cubic)")

pd1 | pd2 | pd3
```

For most conductance problems, degree 3 (cubic) B-splines are a good default
when you choose B-splines. They produce smooth second derivatives, which is
appropriate for a surface that represents gradual changes in habitat quality. Use
degree 1 when you expect abrupt thresholds (e.g. an ecological edge) rather than
smooth transitions.

---

## Mixing linear and spline terms

You can freely mix linear terms and spline terms in the same formula. This is
often the most useful specification: use a spline for the covariate you have
prior reason to believe has a non-linear effect, and keep all others linear to
avoid over-parameterization.

```{r mixed-terms}
# Forest cover linear, altitude as natural spline
fit_mixed <- terradish(
  melip.Fst ~ forestcover + s(altitude, df = 3),
  data              = surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)

coef(fit_mixed)
```

The `forestcover` coefficient is interpreted exactly as in the linear model: its
sign indicates whether forest cover increases or decreases conductance, and its
magnitude reflects the strength of that effect (on the log-conductance scale).
The `s(altitude).*` coefficients together define the altitude shape.

When a model contains both linear and spline terms, the default marginal plot
produces one panel per covariate. Use `marginal_covariates` to view a single
panel in isolation, or to place the two panels side-by-side rather than
stacked:

```{r mixed-marginal, fig.cap = "***Marginal effects from the mixed model (linear forest cover + altitude natural spline, df = 3).*** Left: the linear forest-cover effect (log-conductance scale). Right: the non-linear altitude effect (smooth curve).", fig.height = 4}
p_fc  <- plot(fit_mixed, type = "marginal", data = surface, n = 80,
              marginal_covariates = "forestcover") +
  ggtitle("Forest cover (linear term)")

p_alt <- plot(fit_mixed, type = "marginal", data = surface, n = 80,
              marginal_covariates = "altitude") +
  ggtitle("Altitude: ns (df = 3)")

p_fc | p_alt
```

---

## Summary table: all models

```{r final-aic}
aic_table(
  list(fit_linear, fit_ns2, fit_ns3, fit_ns_alt, fit_ns6,
       fit_bs_alt, fit_ns_both, fit_mixed),
  mod_names = c(
    "Linear",
    "Altitude ns (df=2)",
    "Altitude ns (df=3)",
    "Altitude ns (df=4)",
    "Altitude ns (df=6)",
    "Altitude bs (k=4, deg=3)",
    "Both ns (df=4 each)",
    "Altitude ns (df=3) + linear forestcover"
  ),
  AICc = TRUE
)
```

---

## Practical recommendations

### When to use spline conductance

Use `smooth_loglinear_conductance()` when:

1. **There is a prior ecological reason** to expect a non-linear landscape effect, e.g. an intermediate optimum, a threshold, or a well-documented hump-shaped
   response in the literature.
2. **The linear model's marginal effect is implausible**: e.g. a strong
   negative altitude coefficient that would predict near-zero conductance at
   high elevations where sampling actually occurs.
3. **The data are dense enough** to support the extra parameters. As a rough
   guide, you need at least 5-10 site-pairs per estimated parameter. With 34
   sites (561 pairs) and df = 4, the spline adds 3 parameters over the linear
   model, a modest increase that is usually well-supported.

### When to stay with the linear model

- Small datasets (fewer than ~20 sampling sites).
- No prior hypothesis about a non-linear effect; treat splines as exploratory
  rather than confirmatory.
- The spline and linear models give similar AICc; parsimony favors the simpler
  model.

### Suggested workflow

```{r workflow, eval = FALSE}
# Step 1: Fit a linear baseline: this is your reference
assessment <- terradish_assess_settings(
  S ~ covar2 + s(covar1, df = 3),
  data              = surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  probe_maxit       = 1,
  verbose           = FALSE
)
rec <- assessment$recommended

fit_linear <- terradish(
  S ~ covar1 + covar2,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)

# Step 2: Inspect the linear marginal effects.
# Which covariate might plausibly have a non-linear effect?
# Use marginal_covariates to view one covariate at a time.
plot(fit_linear, type = "marginal", data = surface,
     marginal_covariates = "covar1")

# Step 3: Add one spline term with small df first
fit_spline <- terradish(
  S ~ covar2 + s(covar1, df = 3),
  data              = surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)

# Step 4: Compare with AICc
aic_table(list(fit_linear, fit_spline),
          mod_names = c("Linear", "Spline df=3"),
          AICc = TRUE)

# Step 5: Inspect the spline marginal effect
plot(fit_spline, type = "marginal", data = surface)

# Step 6: If AICc improved and the curve is plausible, try more flexibility
fit_spline4 <- terradish(
  S ~ covar2 + s(covar1, df = 4),
  data              = surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  optimizer         = rec$optimizer,
  control           = rec$control,
  solver            = rec$solver,
  solver_control    = rec$solver_control,
  approximation     = rec$approximation,
  approximation_control = rec$approximation_control
)

aic_table(list(fit_linear, fit_spline, fit_spline4),
          mod_names = c("Linear", "df=3", "df=4"),
          AICc = TRUE)

# Step 7: Visualize the selected model
best_fit <- fit_spline  # or whichever won
plot(best_fit, type = "fit",             data = surface)
plot(best_fit, type = "surface",         data = surface)
plot(best_fit, type = "marginal",        data = surface)
plot(best_fit, type = "marginal_response", data = surface)
```

### Checking for overfitting

Even when AICc improves, inspect these warning signs:

- **Wild oscillation** in the marginal curve, especially with sign reversals at
  the edges of the covariate range.
- **Very large basis coefficients** relative to their standard errors.
- **Near-singular design matrix** or optimizer warnings about convergence.
- **Large discrepancy between ns and bs** at the same df/k: if the two spline
  families give qualitatively different shapes, neither has strong support.

```{r overfit-check, eval = FALSE}
# Quick eigenvalue check: are estimated parameters degenerate?
hess <- fit_ns_alt$fit$hessian
ev <- eigen(hess, symmetric = TRUE, only.values = TRUE)$values
min(ev)   # Should be clearly positive; near-zero → numerical problems

# Compare ns and bs marginal shapes side by side
par(mfrow = c(1, 2))
plot(fit_ns_alt, type = "marginal", data = surface, main = "ns")
plot(fit_bs_alt, type = "marginal", data = surface, main = "bs")
```

---

## Recoverability example

The package includes an example script that simulates a covariance response from
a known non-linear conductance surface and then checks whether spline models
recover it:

```{r recovery-path}
system.file("examples", "spline-recovery-melip.R", package = "terradish")
```

Run it interactively:

```{r recovery-script, eval = FALSE}
source(system.file("examples", "spline-recovery-melip.R",
                   package = "terradish"))

# Simulate a hump-shaped altitude response and check recovery
recovery <- run_spline_recovery_example()
recovery$model_comparison   # AICc table
recovery$parameter_recovery # true vs estimated theta

# Full melip worked example
melip_spline <- fit_melip_spline_example()
melip_spline$model_comparison
```

This script is the right tool when you want to verify that a candidate spline
specification is identifiable for your particular landscape and sampling design
before committing to it in the real analysis.

---

## Quick-reference: the full spline workflow

```{r quickref, eval = FALSE}
library(terradish)
library(terra)

# ---- 1. Prepare the landscape ----
data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

covariates <- c(melip.altitude, melip.forestcover)
names(covariates) <- c("altitude", "forestcover")
covariates <- scale_covariates(covariates)   # Always scale before fitting
surface <- conductance_surface(covariates, melip.coords,
                               directions = 8, saveStack = TRUE)

# ---- 2. Assess once and reuse recommended settings ----
assessment <- terradish_assess_settings(
  melip.Fst ~ forestcover + s(altitude, df = 4),
  data              = surface,
  conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  probe_maxit       = 1,
  verbose           = FALSE
)
rec <- assessment$recommended

# ---- 3. Fit candidate models ----
# Linear baseline
fit_lin <- terradish(
  melip.Fst ~ altitude + forestcover,
  data = surface, conductance_model = loglinear_conductance,
  measurement_model = mlpe,
  optimizer = rec$optimizer,
  control = rec$control,
  solver = rec$solver,
  solver_control = rec$solver_control,
  approximation = rec$approximation,
  approximation_control = rec$approximation_control
)

# Natural spline on altitude (df = 4)
fit_ns <- terradish(
  melip.Fst ~ forestcover + s(altitude, df = 4),
  data = surface, conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  optimizer = rec$optimizer,
  control = rec$control,
  solver = rec$solver,
  solver_control = rec$solver_control,
  approximation = rec$approximation,
  approximation_control = rec$approximation_control
)

# B-spline on altitude (k = 4, cubic)
fit_bs <- terradish(
  melip.Fst ~ forestcover + s(altitude, k = 4, basis = "bs", degree = 3),
  data = surface, conductance_model = smooth_loglinear_conductance,
  measurement_model = mlpe,
  optimizer = rec$optimizer,
  control = rec$control,
  solver = rec$solver,
  solver_control = rec$solver_control,
  approximation = rec$approximation,
  approximation_control = rec$approximation_control
)

# ---- 4. Compare ----
aic_table(list(fit_lin, fit_ns, fit_bs),
          mod_names = c("Linear", "ns df=4", "bs k=4"),
          AICc = TRUE)

# ---- 5. Visualize the best model ----
best <- fit_ns   # replace with whichever model won

plot(best, type = "fit")                       # Observed vs fitted
plot(best, type = "surface",  data = surface)  # Conductance raster + CI

# Marginal plots: all covariates together, or one at a time
plot(best, type = "marginal", data = surface)
plot(best, type = "marginal", data = surface,
     marginal_covariates = "altitude")          # altitude panel only

# Side-by-side comparison of two models' altitude effects with patchwork
library(patchwork)
p1 <- plot(fit_lin,  type = "marginal", data = surface,
           marginal_covariates = "altitude") + ggtitle("Linear")
p2 <- plot(fit_ns,   type = "marginal", data = surface,
           marginal_covariates = "altitude") + ggtitle("ns df=4")
p1 | p2

plot(best, type = "marginal_response",         # Response-scale effects
     data = surface, marginal_covariates = "altitude")

# Extract the conductance surface as a SpatRaster
cond_raster <- conductance(surface, best)
# Layer names: "est", "lower95", "upper95"
plot(cond_raster)
```
