---
title: "Covariance-Based IBR with wishart_covariance"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Covariance-Based IBR with wishart_covariance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4.5,
  warning  = FALSE,
  message  = FALSE
)

```

## Overview

The standard `terradish` workflow uses a pre-computed **distance matrix** (e.g.
F~ST~) as the genetic response. Two measurement models, `leastsquares` and
`mlpe`, accept such matrices and differ only in how they handle the
non-independence of pairwise observations.

When raw individual-level genotype data are available, a more principled route
exists. Instead of collapsing genotypes to pairwise distances and then fitting a
distance model, you can:

1. Summarize the genetic data as a **covariance matrix** between populations.
2. Fit the resistance model with `wishart_covariance`, which evaluates the full
   Wishart likelihood on that covariance matrix using the effective marker
   degrees of freedom as the parameter `nu`.

This approach retains more information than a distance matrix: every off-diagonal
element of the covariance matrix, together with the diagonal, contributes to the
likelihood. It is the recommended path when you have access to the original
genotype data.

| Model | Input `S` | Requires `nu` | Correlation structure |
|---|---|---|---|
| `leastsquares` | distance matrix | no | none |
| `mlpe` | distance matrix | no | shared-site |
| `generalized_wishart` | distance matrix | yes | Wishart |
| `wishart_covariance` | **covariance matrix** | yes | Wishart |

This vignette covers:

- Converting different types of raw genetic data to a covariance matrix with
  `cov_from_genetic_data()` (and `cov_from_biallelic()` for biallelic SNPs).
- Setting up the landscape and conductance surface with the bundled *Melipona*
  data.
- Fitting an IBR model with `wishart_covariance`.
- Interpreting and comparing the fitted model.

---

## Preparing genetic covariance data

`terradish` provides two functions for turning raw genotype data into a
covariance matrix:

- **`cov_from_biallelic(Y, N)`**: fast path for biallelic SNP count matrices.
- **`cov_from_genetic_data(x, ...)`**: flexible path that handles biallelic
  dosage, multiallelic SNPs, microsatellite allele calls, PCA scores, or any
  other numeric encoding.

Both functions return a square, symmetric covariance matrix (one row/column per
population) that can be passed directly to `terradish` as the response `S`.

### Biallelic SNPs: `cov_from_biallelic()`

If your data are derived-allele counts at biallelic loci, use
`cov_from_biallelic()`:

```{r biallelic-example}
library(terradish)

# Simulated counts: 4 populations, 5 biallelic loci
# Y[i, l] = number of derived alleles in population i at locus l
# N[i, l] = total haploid chromosomes sampled (here, 20 diploids → N = 40)
set.seed(42)
n_pop <- 4
n_loci <- 5
Y <- matrix(sample(0:40, n_pop * n_loci, replace = TRUE),
            nrow = n_pop, ncol = n_loci,
            dimnames = list(paste0("pop", seq_len(n_pop)),
                            paste0("snp", seq_len(n_loci))))
N <- matrix(40L, nrow = n_pop, ncol = n_loci)

S_biallelic <- cov_from_biallelic(Y, N)
round(S_biallelic, 3)
```

### Numeric feature matrix: `cov_from_genetic_data()`

For dosage-coded SNPs, PCA scores, or any numeric feature matrix, use
`cov_from_genetic_data()` with `input = "features"` (the default).
The function centers and scales each feature, then applies Gower
double-centering to produce a covariance matrix:

```{r feature-example}
# Individual-level PCA scores: 12 individuals across 4 populations, 3 PCs
set.seed(7)
pca_scores <- matrix(rnorm(12 * 3), nrow = 12, ncol = 3,
                     dimnames = list(NULL, c("PC1", "PC2", "PC3")))
groups <- rep(c("pop1", "pop2", "pop3", "pop4"), each = 3)

S_pca <- cov_from_genetic_data(pca_scores, groups = groups)
round(S_pca, 3)
```

Setting `normalize = "features"` divides the covariance by the number of
retained features; this puts the scale on a per-feature basis rather than
summed across all features, which can make `tau` estimates easier to interpret
across datasets with different numbers of markers. This normalization changes
the scale of `S`; it does not automatically determine the best `nu` for every
genotyping system.

### Microsatellite allele calls: `cov_from_genetic_data()` with `input = "allele_calls"`

For diploid microsatellite data stored as two allele-call columns per locus,
set `input = "allele_calls"` and supply the `loci` argument to identify which
column belongs to which locus. The function expands each locus into per-allele
dosage columns before computing covariance:

```{r microsatellite-example}
# 9 individuals, 2 microsatellite loci, each stored in two allele-call columns
alleles <- data.frame(
  loc1_a = c(100, 100, 102, 102, 104, 104, 100, 102, 104),
  loc1_b = c(100, 102, 102, 104, 104, 100, 104, 100, 102),
  loc2_a = c(200, 202, 200, 202, 204, 204, 200, 204, 202),
  loc2_b = c(202, 202, 204, 204, 204, 200, 202, 200, 204)
)
groups_ms <- rep(c("popA", "popB", "popC"), each = 3)
loci_ms <- c("loc1", "loc1", "loc2", "loc2")

S_ms <- cov_from_genetic_data(
  alleles,
  groups  = groups_ms,
  input   = "allele_calls",
  loci    = loci_ms
)
round(S_ms, 3)
```

## Choosing `nu`

Wishart models treat `nu` as the degrees of freedom in the empirical covariance
matrix. It is fixed by the genetic data used to build `S`, and it is not
estimated by `terradish`.

For biallelic SNPs, the biological marker count and the statistical degrees of
freedom are usually the same: each retained SNP contributes one independent
standardized allele-frequency column. In that case, use the number of retained
polymorphic SNPs (reduced if markers are in linkage disequilibrium).

For microsatellites, **use the number of loci as the conservative default.**
Allele frequencies within a locus are correlated because they sum to a constant
equal to the ploidy, so the locus, not the individual allele, is the natural unit of
information in population genetics. A 10-locus panel contributes at most 10
independent observations in this sense, regardless of how many alleles each
locus has.

A common alternative formula, `sum(K_l - 1)` where `K_l` is the number of
observed alleles at locus `l`, is an **upper bound** that would only be correct
if the alleles within each locus were mutually independent after removing the
sum constraint. In practice they are not: populations that are high in one allele
tend to be predictably low in others through drift and selection. Using
`sum(K_l - 1)` as `nu` can produce confidence intervals that are too narrow and
model-selection statistics that are inflated.

The true effective `nu` for microsatellite data is likely somewhere between the
locus count `L` and `sum(K_l - 1)`. You can compute the upper bound to bracket
the plausible range:

```{r microsatellite-nu}
# Upper bound: sum(K_l - 1) across loci
msat_nu_upper <- sum(vapply(split(seq_along(loci_ms), loci_ms), function(j) {
  alleles_l <- unlist(alleles[, j, drop = FALSE], use.names = FALSE)
  length(unique(alleles_l[!is.na(alleles_l)])) - 1L
}, integer(1)))

n_loci <- length(unique(loci_ms))   # conservative lower bound: number of loci

cat("Conservative lower bound (number of loci):  nu =", n_loci, "\n")
cat("Upper bound (sum of alleles - 1 per locus): nu =", msat_nu_upper, "\n")
```

**Recommended practice:**

1. Set `nu = n_loci` for your primary analysis.
2. Report this choice in your methods section.
3. Repeat the key inference steps (confidence intervals and model comparison)
   with `nu = msat_nu_upper` to show how sensitive your conclusions are to the
   choice.
4. If the model rankings and qualitative conclusions are stable across the range,
   you can report that the results are robust to the specific choice of `nu`.

If only a published F~ST~ matrix is available and allele counts by locus are
unavailable, the locus count is the only reasonable option and is straightforwardly
conservative.

### Key arguments

| Argument | Default | Effect |
|---|---|---|
| `groups` | `NULL` | Population vector; if `NULL`, individual-level covariance is returned |
| `input` | `"features"` | `"features"` for numeric matrices; `"allele_calls"` for allele-call columns |
| `loci` | `NULL` | Required for `"allele_calls"`: maps each column to its locus |
| `center` | `TRUE` | Centers each feature by its global mean before covariance calculation |
| `scale` | `TRUE` | Scales each feature by its global standard deviation; drops constant features |
| `diagonal` | `"auto"` | `"within"` replaces the Gower diagonal with within-population variance (Dyer-style); `"gower"` leaves it as-is |
| `normalize` | `"none"` | `"features"` divides covariance by number of retained features |

The `diagonal = "within"` setting is useful for Dyer-style population-graph
workflows. Note, however, that replacing the diagonal can produce matrices that
are not positive definite for all datasets; check eigenvalues when in doubt:

```{r check-eigenvalues}
ev <- eigen(S_ms, symmetric = TRUE, only.values = TRUE)$values
all(ev > 0)  # TRUE → positive definite; FALSE → semi-definite or indefinite
```

---

## Set up the landscape

The rest of this vignette uses the bundled *Melipona* bee data from the
getting-started vignette. Steps are identical to those already covered there:
see `vignette("getting-started", package = "terradish")` for explanations.

```{r landscape-setup}
library(terra)

data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

covariates <- c(melip.altitude, melip.forestcover)
names(covariates) <- c("altitude", "forestcover")
covariates <- scale_covariates(covariates)

surface <- conductance_surface(covariates, melip.coords, directions = 8)

n_sites <- nrow(melip.coords)
cat("Number of sampling sites:", n_sites, "\n")
```

---

## Generate a covariance response for the demonstration

The bundled *Melipona* data ship with a pairwise F~ST~ matrix (`melip.Fst`), not
raw genotypes. To produce a realistic **covariance** matrix that reflects the
same landscape, we use `simulate_covariance_response()`. This function:

1. Computes the conductance-implied covariance matrix `E(theta)` at specified
   parameter values.
2. Forms the Wishart scale matrix `Sigma = tau * E + sigma * I`.
3. Draws `S ~ Wishart(nu, Sigma) / nu`.

```{r simulate-covariance}
# Use conductance parameters roughly consistent with the melip fit in the
# getting-started vignette (forestcover positive, altitude negative).
# nu = 1000 mimics a typical medium-density SNP panel.
sim <- simulate_covariance_response(
  theta             = c(forestcover = 1.0, altitude = -0.5),
  formula           = ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  tau               = 1.0,
  sigma             = 0.5,
  nu                = 1000,
  seed              = 123
)

# This is what would come from cov_from_genetic_data() applied to real genotypes.
S_cov <- sim$covariance
dim(S_cov)
round(S_cov[1:5, 1:5], 4)
```

In a real analysis you would replace `S_cov` with the output of
`cov_from_genetic_data()` applied to your actual genotype data, for example:

```{r real-data-sketch, eval = FALSE}
# With biallelic SNP count matrices (Y = derived allele counts, N = haploid sample sizes)
S_cov <- cov_from_biallelic(Y, N)

# With individual dosage / PCA scores grouped by population
S_cov <- cov_from_genetic_data(dosage_matrix, groups = pop_vector,
                                normalize = "features")

# With microsatellite allele calls
S_cov <- cov_from_genetic_data(allele_df, groups = pop_vector,
                                input = "allele_calls",
                                loci  = locus_id_vector)
```

---

## Fit the model with `wishart_covariance`

The only differences from a standard `terradish` call are:

- Pass the **covariance matrix** `S_cov` as the left-hand side of the formula.
- Set `measurement_model = wishart_covariance`.
- Supply `nu` (the effective Wishart degrees of freedom for `S_cov`).

```{r fit-wishart-covariance}
fit_wc <- terradish(
  S_cov ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = wishart_covariance,
  nu                = 1000
)

summary(fit_wc)
```

### Understanding the nuisance parameters

The `wishart_covariance` measurement model fits two nuisance parameters (`phi`):

- **`tau`** (≥ 0): a nonnegative scale applied to the conductance-implied
  covariance `E`. If `tau` ≈ 0, there is no detectable resistance signal in the
  covariance data. If `tau` is large and well-estimated, resistance distance
  strongly predicts genetic covariance.
- **`sigma`** (log-scale): a scalar nugget added as `exp(sigma) * I` to the
  fitted covariance matrix. This absorbs genetic variation that is not explained
  by landscape resistance, e.g. genetic drift, population-size differences, or
  loci under local selection.

The fitted covariance model is:
> **Sigma = tau · E(theta) + exp(sigma) · I**

where `E(theta)` is the generalized inverse of the graph Laplacian evaluated at
the estimated conductance parameters `theta`.

---

## The role of `nu`: an effective sample size

`nu` is the most consequential setting in a covariance-based fit, and it
behaves in a way that surprises most users. The entire Wishart negative
log-likelihood is multiplied by `nu / 2`, so `nu` is a global multiplier of the
objective surface. That single fact has three separate consequences.

**1. Point estimates do not depend on `nu`.** Scaling an objective by a positive
constant does not move its optimum, so the estimated conductance parameters
`theta` (and the nuisance parameters `tau`, `sigma`) are *identical* for any
`nu`. We can verify this directly:

```{r nu-invariance}
fit_nu <- function(nu) {
  f <- terradish(
    S_cov ~ forestcover + altitude,
    data              = surface,
    conductance_model = loglinear_conductance,
    measurement_model = wishart_covariance,
    nu                = nu
  )
  list(theta = coef(f),
       se    = sqrt(diag(solve(f$fit$hessian))))
}

f100   <- fit_nu(100)
f1000  <- fit_nu(1000)
f10000 <- fit_nu(10000)

# Point estimates are byte-for-byte identical across three orders of magnitude
rbind(`nu=100`   = f100$theta,
      `nu=1000`  = f1000$theta,
      `nu=10000` = f10000$theta)
```

**2. Standard errors scale as 1 / sqrt(`nu`).** The Hessian scales with `nu`, so
the asymptotic covariance scales with 1 / `nu` and every standard error scales
with 1 / sqrt(`nu`). More markers carry more information and produce tighter
intervals, exactly as they should:

```{r nu-se}
rbind(`nu=100`   = f100$se,
      `nu=1000`  = f1000$se,
      `nu=10000` = f10000$se)

# Ratio confirms the 1/sqrt(nu) scaling: a 10x increase in nu
# shrinks SEs by sqrt(10) ~ 3.16
f100$se / f1000$se
```

**3. Model selection and hypothesis tests depend strongly on `nu`.** The
log-likelihood scales linearly with `nu`, but the AIC/BIC complexity penalty
(and the chi-squared reference distribution for a likelihood-ratio test) does
not. With large `nu` the penalty becomes negligible and extra covariates are
almost always retained; with small `nu`, parsimony dominates. The *same data*
can favor a simpler or a more complex model purely through the choice of `nu`.

The practical upshot: **`nu` is an effective sample size.** It does not change
*what* the data say about the shape of the conductance surface, but it controls
*how strongly* the data speak. Choose it carefully:

- For biallelic SNPs, use the number of retained (polymorphic) SNPs.
- For microsatellites, use approximately the number of independent allele
  frequencies, sum over loci of (alleles - 1), not the raw number of loci.
- Markers in strong linkage disequilibrium are not independent, so the
  effective `nu` is **smaller** than the raw marker count. Thin to roughly
  unlinked loci, or use a deliberately conservative value.
- When `nu` is uncertain, report a brief sensitivity analysis: refit across a
  plausible range and check whether your conclusions (significance, model
  ranking) are stable.

---

## Recover the true parameters

Because we simulated these data with known parameters, we can check how well the
model recovers them. In practice this comparison is obviously not available.

```{r recovery}
cat("True conductance parameters:\n")
cat("  forestcover:", 1.0, "  altitude:", -0.5, "\n\n")

cat("Estimated conductance parameters:\n")
print(round(coef(fit_wc), 3))

cat("\nTrue nuisance parameters (on the variance scale):\n")
cat("  tau   =", sim$tau, "\n")
cat("  sigma =", sim$sigma, " (nugget variance; fitted model stores log(sigma))\n\n")

cat("Estimated nuisance parameters (phi vector, log scale for sigma):\n")
phi_est <- round(fit_wc$fit$phi, 3)
print(phi_est)

cat("\nFor comparison, exp(phi['sigma']) =",
    round(exp(fit_wc$fit$phi["sigma"]), 3),
    " vs. true sigma =", sim$sigma, "\n")
```

---

## Visualize the fit

### Observed vs. fitted covariance

```{r plot-fit, fig.cap = "***Observed vs. fitted genetic covariance.***"}
plot(fit_wc, type = "fit")
```

Each point represents a pair of sites. The x-axis is the fitted covariance from
the Wishart model; the y-axis is the observed covariance in `S_cov`. Points
near the diagonal indicate a good fit.

### Estimated conductance surface

```{r plot-surface, fig.cap = "***Estimated conductance surface with 95% CI bounds.***"}
plot(fit_wc, type = "surface", data = surface)
```

### Marginal effects on conductance

```{r plot-marginal, fig.cap = "***Marginal effects on the fitted conductance scale.***"}
plot(fit_wc, type = "marginal", data = surface)
```

Response-scale marginal plots are currently defined for regression-style
measurement models such as `mlpe`. For covariance-response models, inspect the
conductance-scale marginal effects alongside the observed-vs-fitted covariance
plot above.

If one covariate has weak focal support at its tails, apply focal-support
clamping during plotting and surface prediction:

```{r wishart-support-clamp, eval = FALSE}
plot(
  fit_wc,
  type = "marginal",
  data = surface,
  support = "focal",
  support_probs = c(0.01, 0.99),
  clamp_covariates = c("forestcover", "altitude")
)

cond_wc_focal <- conductance(
  surface,
  fit_wc,
  support = "focal",
  support_probs = c(0.01, 0.99),
  clamp_covariates = c("forestcover", "altitude")
)
```

---

## Compare with `generalized_wishart`

`generalized_wishart` fits the same Wishart distribution but works on the
**distance** representation of the same genetic data. The distance matrix is
obtained from the covariance matrix via `dist_from_cov()`:

```{r compare-gw}
# Convert the simulated covariance to a distance matrix
S_dist <- dist_from_cov(S_cov)

fit_gw <- terradish(
  S_dist ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = generalized_wishart,
  nu                = 1000
)

# Side-by-side parameter estimates
cat("wishart_covariance estimates:\n")
print(round(coef(fit_wc), 3))

cat("\ngeneralized_wishart estimates:\n")
print(round(coef(fit_gw), 3))
```

Both models use the same Wishart distribution, so when the covariance and
distance representations contain the same information the conductance estimates
should agree closely. The AIC comparison is not meaningful across the two models
here because the response is different (`S_cov` vs. `S_dist`), but the
parameter estimates and their standard errors are directly comparable.

---

## Choosing between covariance and distance input

| Consideration | `wishart_covariance` | `generalized_wishart` |
|---|---|---|
| Raw genotype data available | **Preferred**: use `cov_from_genetic_data()` | Requires an extra distance-conversion step |
| Only a pre-computed distance matrix is available | Requires converting to covariance (not always reversible) | **Preferred** |
| Population-size differences matter | `diagonal = "within"` can encode within-population spread | Not directly available |
| Goal is a Dyer-style population graph | Diagonal from within-population variance is natural | Less natural |
| `nu` is known | Both models require `nu` | Both models require `nu` |

If you only have a distance matrix (e.g. F~ST~) and no individual-level data,
use `generalized_wishart`. If you have raw genotypes, start with
`cov_from_genetic_data()` and `wishart_covariance`.

---

## Quick-reference: the full workflow

```{r full-workflow, eval = FALSE}
library(terradish)
library(terra)

# 1. Load landscape data and build conductance surface
data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

covariates <- c(melip.altitude, melip.forestcover)
names(covariates) <- c("altitude", "forestcover")
covariates <- scale_covariates(covariates)
surface <- conductance_surface(covariates, melip.coords, directions = 8)

# 2. Prepare the genetic covariance matrix from raw genotype data
#    (replace with your actual data and effective nu)

# Option A: biallelic SNPs (Y = derived allele counts, N = haploid totals)
S <- cov_from_biallelic(Y, N)

# Option B: individual-level numeric features grouped by population
S <- cov_from_genetic_data(dosage_matrix, groups = pop_vector,
                            normalize = "features")

# Option C: microsatellite allele calls
S <- cov_from_genetic_data(allele_df, groups = pop_vector,
                            input = "allele_calls",
                            loci  = locus_id_vector)

nu <- 500  # effective Wishart degrees of freedom for S

# 3. Fit with wishart_covariance
fit <- terradish(
  S ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = wishart_covariance,
  nu                = nu
)

# 4. Summarize and visualize
summary(fit)
coef(fit)

plot(fit, type = "fit")
plot(fit, type = "surface", data = surface)
plot(fit, type = "marginal", data = surface)   # conductance-scale effects
```
