---
title: "Covariance-Based IBR with wishart_covariance"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Covariance-Based IBR with wishart_covariance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4.5,
  warning  = FALSE,
  message  = FALSE
)

```

## Overview

The standard `terradish` workflow uses a pre-computed **distance
matrix** (e.g. F~ST~) as the genetic response. Two measurement models,
`leastsquares` and `mlpe`, accept such matrices and differ only in how
they handle the non-independence of pairwise observations.

When raw individual-level genotype data are available, you can retain a
population covariance summary rather than beginning from a precomputed
distance matrix:

1.  Summarize the genetic data as a **covariance matrix** between
    populations.
2. Fit the resistance model with `wishart_covariance`, using a declared
   effective degrees of freedom, `nu`.

Both Wishart interfaces use the **contrast likelihood**. For an orthonormal
site-contrast matrix `Q`, the covariance likelihood uses `Q' S Q` and the
corresponding fitted contrast covariance. Components in the constant-site
direction are excluded. Centering `S` therefore leaves this likelihood
unchanged. The previous full-rank formulation was removed because that
constant direction depends on arbitrary centering conventions and does not
provide the intended resistance information.

`dist_from_cov(S)` retains the covariance information used by this likelihood.
Covariance and matching squared-distance inputs are two representations of
the same contrast model. Neither is inherently more informative. Begin with
`vignette("getting-started", package = "terradish")` for the graph workflow.

| Model | Input `S` | Requires `nu` | Correlation structure |
|----|----|----|----|
| `leastsquares` | distance matrix | no | none |
| `mlpe` | distance matrix | no | shared-site |
| `generalized_wishart` | admissible squared-distance matrix | yes | generalized Wishart |
| `wishart_covariance` | **covariance matrix** | yes | Wishart |

This vignette covers:

- Converting different types of raw genetic data to a covariance matrix
  with `cov_from_genetic_data()` (and `cov_from_biallelic()` for
  biallelic SNPs).
- Setting up the landscape and conductance surface with the bundled
  *Melipona* data.
- Fitting an IBR model with `wishart_covariance`.
- Interpreting and comparing the fitted model.

------------------------------------------------------------------------

## Preparing genetic covariance data

`terradish` provides two functions for turning raw genotype data into a
covariance matrix:

- **`cov_from_biallelic(Y, N)`**: fast path for biallelic SNP count
  matrices.
- **`cov_from_genetic_data(x, ...)`**: flexible path that handles
  biallelic dosage, multiallelic SNPs, microsatellite allele calls, PCA
  scores, or any other numeric encoding.

Both functions return a square, symmetric covariance matrix (one
row/column per population). Check the construction, sampling design, site
order, and covariance geometry before using the matrix as response `S`.
The helpers are provided by the companion `landgraph` package.

### Biallelic SNPs: `cov_from_biallelic()`

If your data are derived-allele counts at biallelic loci, use
`cov_from_biallelic()`:

```{r biallelic-example}
library(terradish)

# Simulated counts: 4 populations, 5 biallelic loci
# Y[i, l] = number of derived alleles in population i at locus l
# N[i, l] = total haploid chromosomes sampled (here, 20 diploids → N = 40)
set.seed(42)
n_pop <- 4
n_loci <- 5
Y <- matrix(sample(0:40, n_pop * n_loci, replace = TRUE),
            nrow = n_pop, ncol = n_loci,
            dimnames = list(paste0("pop", seq_len(n_pop)),
                            paste0("snp", seq_len(n_loci))))
N <- matrix(40L, nrow = n_pop, ncol = n_loci)

S_biallelic <- cov_from_biallelic(Y, N)
round(S_biallelic, 3)
```

Biallelic standardization can give rare variants large influence. Apply and
report a defensible minor-allele-frequency filter, then examine sensitivity
to that filter. Unequal chromosome counts also produce unequal sampling
variance across populations; a common nugget does not remove that concern.
Covariance metadata describe the construction, not proof of Wishart sampling.

### Numeric feature matrix: `cov_from_genetic_data()`

For dosage-coded SNPs, PCA scores, or any numeric feature matrix, use
`cov_from_genetic_data()` with `input = "features"` (the default). The
function centers and scales each feature, then applies Gower
double-centering to produce a covariance matrix:

```{r feature-example}
# Individual-level PCA scores: 12 individuals across 4 populations, 3 PCs
set.seed(7)
pca_scores <- matrix(rnorm(12 * 3), nrow = 12, ncol = 3,
                     dimnames = list(NULL, c("PC1", "PC2", "PC3")))
groups <- rep(c("pop1", "pop2", "pop3", "pop4"), each = 3)

S_pca <- cov_from_genetic_data(pca_scores, groups = groups)
round(S_pca, 3)
```

Setting `normalize = "features"` divides the covariance by the number of
retained features; this puts the scale on a per-feature basis rather
than summed across all features, which can make `tau` estimates easier
to interpret across datasets with different numbers of markers. This
normalization changes the scale of `S`; it does not automatically
determine the best `nu` for every genotyping system.

### Microsatellite allele calls: `cov_from_genetic_data()` with `input = "allele_calls"`

For diploid microsatellite data stored as two allele-call columns per
locus, set `input = "allele_calls"` and supply the `loci` argument to
identify which column belongs to which locus. The function expands each
locus into per-allele dosage columns before computing covariance:

```{r microsatellite-example}
# 9 individuals, 2 microsatellite loci, each stored in two allele-call columns
alleles <- data.frame(
  loc1_a = c(100, 100, 102, 102, 104, 104, 100, 102, 104),
  loc1_b = c(100, 102, 102, 104, 104, 100, 104, 100, 102),
  loc2_a = c(200, 202, 200, 202, 204, 204, 200, 204, 202),
  loc2_b = c(202, 202, 204, 204, 204, 200, 202, 200, 204)
)
groups_ms <- rep(c("popA", "popB", "popC"), each = 3)
loci_ms <- c("loc1", "loc1", "loc2", "loc2")

S_ms <- cov_from_genetic_data(
  alleles,
  groups  = groups_ms,
  input   = "allele_calls",
  loci    = loci_ms
)
round(S_ms, 3)
```

## Choosing `nu`

`nu` is a positive effective degrees-of-freedom value supplied by the analyst.
It is not estimated by `terradish`, and it is not fixed automatically by the
number of markers used to build `S`. Linkage creates dependence among nearby
markers; shared population history can create dependence even among unlinked
markers. Counting SNPs, loci, or within-locus allele contrasts does not by
itself establish effective replication.

A block jackknife can assess how marker blocks contribute to variability.
Because it need not capture all dependence due to shared history, regard its
implied information as an upper-bound guide rather than a complete calibration.
Describe the blocks and the uncertainty sources the procedure represents.

Declare a primary value with its rationale and a range of plausible values.
Use `terradish_rescale_nu()` for sensitivity of intervals and likelihood-based
comparisons. Use fixed-domain cross-validation to assess conductance terms
and shape. A stable result over a declared range is evidence of sensitivity
within that range, not proof that the primary value is correct.

If only a published F~ST~ matrix is available, a reported locus count does not
make that response Wishart. F~ST~ is a ratio estimator; consider MLPE for such
dissimilarities and evaluate its shared-site assumptions. See
`vignette("model-comparison", package = "terradish")` for predictive comparisons.

### Key arguments

| Argument | Default | Effect |
|----|----|----|
| `groups` | `NULL` | Population vector; if `NULL`, individual-level covariance is returned |
| `input` | `"features"` | `"features"` for numeric matrices; `"allele_calls"` for allele-call columns |
| `loci` | `NULL` | Required for `"allele_calls"`: maps each column to its locus |
| `center` | `TRUE` | Centers each feature by its global mean before covariance calculation |
| `scale` | `TRUE` | Scales each feature by its global standard deviation; drops constant features |
| `diagonal` | `"auto"` | `"auto"` selects `"gower"`; `"within"` replaces the diagonal with within-population variance |
| `normalize` | `"none"` | `"features"` divides covariance by number of retained features |

The default grouped-data construction uses `diagonal = "gower"`. Replacing
that diagonal with `diagonal = "within"` changes the covariance estimand and
can change fitted signs; the supplied release simulations recovered the
expected signs with Gower diagonals but not with the within-population
replacement. Retain `within` only for a separately justified construction,
and compare its effect explicitly. It may also produce an indefinite matrix.

```{r check-eigenvalues}
ev <- eigen(S_ms, symmetric = TRUE, only.values = TRUE)$values
range(ev)
```

**How to read the output.** A centered covariance has a zero eigenvalue in
the constant direction, so strict positive definiteness of the full matrix
is not the target. Tiny negative values may reflect rounding. Substantive
negative contrast eigenvalues require investigation; changing the diagonal
is not a neutral repair.

------------------------------------------------------------------------

## Set up the landscape

The rest of this vignette uses the bundled *Melipona* bee data from the
getting-started vignette. Steps are identical to those already covered
there: see `vignette("getting-started", package = "terradish")` for
explanations.

```{r landscape-setup}
library(terra)

data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

covariates <- c(melip.altitude, melip.forestcover)
names(covariates) <- c("altitude", "forestcover")
covariates <- scale_covariates(covariates)

surface <- conductance_surface(covariates, melip.coords, directions = 8)

n_sites <- nrow(melip.coords)
cat("Number of sampling sites:", n_sites, "\n")
```

------------------------------------------------------------------------

## Generate a covariance response for the demonstration

The bundled *Melipona* data ship with a pairwise F~ST~ matrix
(`melip.Fst`), not raw genotypes. To produce a realistic **covariance**
matrix that reflects the same landscape, we use
`simulate_covariance_response()`. This function:

1.  Computes the conductance-implied covariance matrix `E(theta)` at
    specified parameter values.
2.  Forms the Wishart scale matrix `Sigma = tau * E + sigma * I`.
3.  Draws `S ~ Wishart(nu, Sigma) / nu`.

```{r simulate-covariance}
# Use conductance parameters roughly consistent with the melip fit in the
# getting-started vignette (forestcover positive, altitude negative).
# nu = 1000 is an illustrative simulation parameter, not a marker-count rule.
sim <- simulate_covariance_response(
  theta             = c(forestcover = 1.0, altitude = -0.5),
  formula           = ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  tau               = 1.0,
  sigma             = 0.5,
  nu                = 1000,
  seed              = 123
)

# This draw follows the fitted statistical model; real genotype summaries may not.
S_cov <- sim$covariance
dim(S_cov)
round(S_cov[1:5, 1:5], 4)
```

In a real analysis you would replace `S_cov` with the output of
`cov_from_genetic_data()` applied to your actual genotype data, for
example:

```{r real-data-sketch, eval = FALSE}
# With biallelic SNP count matrices (Y = derived allele counts, N = haploid sample sizes)
S_cov <- cov_from_biallelic(Y, N)

# With individual dosage / PCA scores grouped by population
S_cov <- cov_from_genetic_data(dosage_matrix, groups = pop_vector,
                                normalize = "features")

# With microsatellite allele calls
S_cov <- cov_from_genetic_data(allele_df, groups = pop_vector,
                                input = "allele_calls",
                                loci  = locus_id_vector)
```

------------------------------------------------------------------------

## Fit the model with `wishart_covariance`

The only differences from a standard `terradish` call are:

- Pass the **covariance matrix** `S_cov` as the left-hand side of the
  formula.
- Set `measurement_model = wishart_covariance`.
- Supply `nu` (the effective Wishart degrees of freedom for `S_cov`).

```{r fit-wishart-covariance}
fit_wc <- terradish(
  S_cov ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = wishart_covariance,
  nu                = 1000
)

summary(fit_wc)
```

### Understanding the nuisance parameters

The `wishart_covariance` measurement model fits two nuisance parameters
(`phi`):

- **`tau`** (≥ 0): a nonnegative weight applied to the
  conductance-implied covariance `E`. If `tau` ≈ 0, there is no
  detectable graph-covariance signal under this response and model. Its
  magnitude is not comparable to the Gaussian distance-model `tau`,
  which is a log precision.
- **`sigma`** (log-scale): a scalar nugget added as `exp(sigma) * I` to
  the fitted covariance matrix. It captures diagonal variation not
  represented by the graph kernel, including sampling variation and
  model mismatch. It is not the Gaussian raster-smoothing scale and is
  not a direct estimate of drift or population size.

The covariance kernel is `Sigma = tau * E(theta) + exp(sigma) * I`.
The likelihood evaluates its projection into the site-contrast subspace.

where `E(theta)` is the generalized inverse of the graph Laplacian
evaluated at the estimated conductance parameters `theta`.

------------------------------------------------------------------------

## The role of `nu`: an effective sample size

`nu` is the most consequential setting in a covariance-based fit. The
parameter-dependent Wishart objective is multiplied by `nu / 2`, so
`nu` is a global multiplier of the objective surface, which has three
separate consequences.

**1. The likelihood optimum does not depend on `nu`.** Scaling an
objective by a positive constant does not move its optimum, so the
theoretical point estimates of `theta`, `tau`, and `sigma` are invariant
to `nu`. Separate numerical fits should agree up to optimization
tolerance:

```{r nu-invariance}
# Reuse the fitted optimum; rescale its likelihood and uncertainty.
fit_nu <- function(nu) {
  f <- terradish_rescale_nu(fit_wc, nu = nu)
  list(theta = coef(f), se = sqrt(diag(vcov(f))))
}
f100   <- fit_nu(100)
f1000  <- fit_nu(1000)
f10000 <- fit_nu(10000)

# Point estimates agree across this 100-fold range
rbind(`nu=100`   = f100$theta,
      `nu=1000`  = f1000$theta,
      `nu=10000` = f10000$theta)
```

**2. Standard errors scale as 1 / sqrt(`nu`).** The Hessian scales with
`nu`, so the asymptotic covariance scales with 1 / `nu` and every
interior standard error scales with 1 / sqrt(`nu`). Boundary inference
requires separate care. Under the model, more
approximately independent marker information produces tighter intervals:

```{r nu-se}
rbind(`nu=100`   = f100$se,
      `nu=1000`  = f1000$se,
      `nu=10000` = f10000$se)

# Ratio confirms the 1/sqrt(nu) scaling: a 10x increase in nu
# shrinks SEs by sqrt(10) ~ 3.16
f100$se / f1000$se
```

**3. Model selection and hypothesis tests depend strongly on `nu`.** The
log-likelihood scales linearly with `nu`, but the AIC/BIC complexity
penalty (and the chi-squared reference distribution for a
likelihood-ratio test) does not. With large `nu` the penalty becomes
negligible and extra covariates are almost always retained; with small
`nu`, parsimony dominates. The *same data* can favor a simpler or a more
complex model purely through the choice of `nu`.

Report how intervals and model comparisons depend on `nu`. Rescaling
does not re-estimate conductance, so this sensitivity is inexpensive and
separates the fitted surface from the assumed precision. Use the same `nu`
for every candidate in a comparison. Neither a high AIC weight nor a small
nominal LRT p-value establishes that `nu` represents independent information.

------------------------------------------------------------------------

## Recover the true parameters

Because we simulated these data with known parameters, we can check how
well the model recovers them. In practice this comparison is obviously
not available.

```{r recovery}
cat("True conductance parameters:\n")
cat("  forestcover:", 1.0, "  altitude:", -0.5, "\n\n")

cat("Estimated conductance parameters:\n")
print(round(coef(fit_wc), 3))

cat("\nTrue nuisance parameters (on the variance scale):\n")
cat("  tau   =", sim$tau, "\n")
cat("  sigma =", sim$sigma, " (nugget variance; fitted model stores log(sigma))\n\n")

cat("Estimated nuisance parameters (phi vector, log scale for sigma):\n")
phi_est <- round(fit_wc$fit$phi, 3)
print(phi_est)

cat("\nFor comparison, exp(phi['sigma']) =",
    round(exp(fit_wc$fit$phi["sigma"]), 3),
    " vs. true sigma =", sim$sigma, "\n")
```

------------------------------------------------------------------------

## Visualize the fit

### Observed vs. fitted covariance

```{r plot-fit, fig.cap = "***Observed and fitted covariance in the site-centered representation.***"}
plot(fit_wc, type = "fit")
```

Each point represents a pair of sites. The x-axis is the fitted
covariance from the Wishart model; the y-axis is the observed covariance
in the centered representation used by the contrast fit. Points near the
diagonal indicate agreement for that summary, not calibrated sampling uncertainty.

### Estimated conductance surface

```{r plot-surface, fig.cap = "***Estimated conductance surface with 95% CI bounds.***"}
plot(fit_wc, type = "surface", data = surface)
```

### Marginal associations on conductance

```{r plot-marginal, fig.cap = "***Marginal associations on the fitted conductance scale.***"}
plot(fit_wc, type = "marginal", data = surface)
```

Response-scale marginal plots are currently defined for regression-style
measurement models such as `mlpe`. For covariance-response models,
inspect the conductance-scale marginal associations alongside the
observed-vs-fitted covariance plot above.

If one covariate has weak focal support at its tails, apply
focal-support clamping during plotting and surface prediction:

```{r wishart-support-clamp, eval = FALSE}
plot(
  fit_wc,
  type = "marginal",
  data = surface,
  support = "focal",
  support_probs = c(0.01, 0.99),
  clamp_covariates = c("forestcover", "altitude")
)

cond_wc_focal <- conductance(
  surface,
  fit_wc,
  support = "focal",
  support_probs = c(0.01, 0.99),
  clamp_covariates = c("forestcover", "altitude")
)
```

------------------------------------------------------------------------

## Preparing a generalized-Wishart squared-distance response

The generalized-Wishart likelihood expects a squared-distance response
${D = d^2}$, not an arbitrary symmetric dissimilarity matrix. For ${n}$
sites, `terradish` forms the centered Gram matrix

$$B = -\frac{1}{2} H D H, \qquad
H = I - \frac{1}{n}\mathbf{1}\mathbf{1}^{\mathsf T},$$

and requires ${B}$ to be positive semidefinite within a relative
numerical tolerance. This geometry is the mathematical support condition
for the likelihood. It is separate from whether the selected genetic
summary is biologically appropriate for the sampling design and
scientific question.

Use `check_distance_response()` on the matrix that you actually intend
to fit:

```{r check-gw-response}
S_dist <- dist_from_cov(S_cov)
gw_check <- check_distance_response(S_dist)
gw_check[c("admissible", "relative_min_eigenvalue", "n_negative")]
```

**How to read the diagnostic:** `admissible = TRUE` means the implied
centered Gram matrix is positive semidefinite within tolerance. A very
small negative `relative_min_eigenvalue` can reflect floating-point
rounding. `n_negative` counts only eigenvalues that are more negative
than the scaled tolerance. `terradish()` performs this check
automatically before any generalized-Wishart fit and stops on
substantive violations.

The checker does not select a genetic distance for you. Choose the
response based on the biology, marker type, sampling design, and
intended interpretation, then test the realized matrix. The name of a
distance measure is not enough because scaling, locus aggregation,
transformations, and missing-data handling can change the resulting
geometry.

### Potentially relevant distance measures available in R

The table is a navigation aid, not a list of recommended or
interchangeable responses. Functions can differ in their scaling, input
class, treatment of missing data, and whether they operate on
individuals or populations. Confirm what the selected function returns.
If it returns ordinary Euclidean distances ${d}$, square the matrix to
obtain ${D = d^2}$ before calling `check_distance_response()` or fitting
`generalized_wishart`.

| Candidate measure | Example R implementations | Guidance for generalized Wishart |
|----|----|----|
| Edwards/angular and Cavalli-Sforza-Edwards chord distances | `adegenet::dist.genpop(method = 2)`, `poppr::edwards.dist()`, `hierfstat::genet.dist(method = "Dch")` | The `adegenet` manual labels method 2 Euclidean. Cavalli-Sforza-Edwards chord distance is one potentially relevant candidate, not a required or preferred response. Verify the exact implementation and square ordinary distances before checking. |
| Reynolds distance | `adegenet::dist.genpop(method = 3)`, `poppr::reynolds.dist()` | The `adegenet` manual labels its implementation Euclidean. Verify the realized matrix and its scaling. |
| Rogers distance | `adegenet::dist.genpop(method = 4)`, `poppr::rogers.dist()`, `hierfstat::genet.dist(method = "Dr")` | The `adegenet` manual labels method 4 Euclidean. Square ordinary distances and check the resulting matrix. |
| Nei distances | `adegenet::dist.genpop(method = 1)`, `poppr::nei.dist()`, variants from `hierfstat::genet.dist()` | These names cover different formulas. The `adegenet` manual labels method 1 non-Euclidean, so do not assume admissibility. |
| Prevosti distance | `adegenet::dist.genpop(method = 5)`, `poppr::prevosti.dist()`, `hierfstat::genet.dist(method = "Cp")` | The `adegenet` manual labels method 5 non-Euclidean. Do not assume admissibility. |
| Pairwise F~ST~ or F~ST~/(1 - F~ST~) | `hierfstat::pairwise.WCfst()`, `hierfstat::pairwise.neifst()` | These are common landscape-genetic responses, but they have no general squared-Euclidean guarantee. F~ST~ is a ratio estimator without a Wishart sampling justification; passing a geometry check is insufficient. Consider `mlpe` and assess its assumptions. |
| Euclidean, chord, or Hellinger distances on transformed numeric features | `stats::dist()`, `vegan::vegdist(method = "euclidean")`, `vegan::vegdist(method = "chord")`, `vegan::vegdist(method = "hellinger")` | Geometry depends on the feature representation and transformation. Scientific appropriateness remains study-specific. |

`ade4::is.euclid()` provides an optional independent diagnostic for
ordinary distances. Because it squares its input internally, the
equivalent call for an already squared generalized-Wishart response `D`
is:

```{r ade4-cross-check, eval = FALSE}
# D is already squared, so take its square root for ade4::is.euclid().
ade4::is.euclid(as.dist(sqrt(D)))
```

Do not call `ade4::is.euclid(as.dist(D))` for this purpose because that
tests whether ${D^2}$, rather than ${D}$, has the required geometry.
Functions such as `ade4::cailliez()` and `ade4::lingoes()` can make an
ordinary distance matrix Euclidean for ordination, but `terradish` does
not apply them automatically. They change the genetic response and can
change its biological interpretation.

------------------------------------------------------------------------

## Compare with `generalized_wishart`

`generalized_wishart` uses the corresponding generalized Wishart
formulation for an admissible **squared-distance** representation of the
same genetic data. Here the response was obtained from the covariance
matrix with `dist_from_cov()` and checked above:

```{r compare-gw}
fit_gw <- terradish(
  S_dist ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = generalized_wishart,
  nu                = 1000
)

# Side-by-side parameter estimates
cat("wishart_covariance estimates:\n")
print(round(coef(fit_wc), 3))

cat("\ngeneralized_wishart estimates:\n")
print(round(coef(fit_gw), 3))
```

When the covariance and admissible squared-distance representations
contain the same information under identical centering, scaling, sites,
and `nu`, their fitted conductance parameters should agree up to
numerical tolerance. The AIC comparison is not meaningful across the two
models because the responses differ. Standard errors should only be
compared after confirming that the transformation is lossless and the
effective information assumptions are identical.

------------------------------------------------------------------------

## Choosing between covariance and distance input

| Consideration | Guidance |
|----|----|
| Raw genotype data available | Construct and document an appropriate covariance summary; assess filtering and unequal sampling. |
| Covariance and its matching squared distances available | Both Wishart interfaces use the same contrast likelihood. Choose the representation that makes preprocessing clearest. |
| Only a precomputed dissimilarity is available | Geometry is necessary but insufficient for Wishart sampling; consider MLPE for generic genetic distances. |
| Site-specific sampling variance matters | A common nugget does not fully address unequal sampling; examine design and sensitivity. |
| Effective information is uncertain | Declare a range of `nu`, rescale inference, and assess conductance complexity with held-out prediction. |

### Environmental differences between sites

`pairwise_covariates()` combines geographic distances and endpoint environmental
differences in one design. Both `mlpe_covariates()` and `wishart_covariates()`
accept that design, while linking it to different response models. For Wishart,
the nonnegative environmental kernel weights describe conditional contributions
to covariance structure. They do not identify a particular biological
mechanism. See `vignette("ibe-ibr-workflow", package = "terradish")` for a
worked comparison, conditional conductance estimates, and IBE:IBR ratios.

------------------------------------------------------------------------

## Quick-reference: the full workflow

```{r full-workflow, eval = FALSE}
library(terradish)
library(terra)

# 1. Load landscape data and build conductance surface
data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

covariates <- c(melip.altitude, melip.forestcover)
names(covariates) <- c("altitude", "forestcover")
covariates <- scale_covariates(covariates)
surface <- conductance_surface(covariates, melip.coords, directions = 8)

# 2. Prepare the genetic covariance matrix from raw genotype data
#    (replace with your actual data and effective nu)

# Option A: biallelic SNPs (Y = derived allele counts, N = haploid totals)
S <- cov_from_biallelic(Y, N)

# Option B: individual-level numeric features grouped by population
S <- cov_from_genetic_data(dosage_matrix, groups = pop_vector,
                            normalize = "features")

# Option C: microsatellite allele calls
S <- cov_from_genetic_data(allele_df, groups = pop_vector,
                            input = "allele_calls",
                            loci  = locus_id_vector)

nu <- effective_marker_df  # justify, report, and evaluate sensitivity

# 3. Fit with wishart_covariance
fit <- terradish(
  S ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = wishart_covariance,
  nu                = nu
)

# 4. Summarize and visualize
summary(fit)
coef(fit)

plot(fit, type = "fit")
plot(fit, type = "surface", data = surface)
plot(fit, type = "marginal", data = surface)   # conductance-scale associations
```

## Summary of key functions

| Function | Purpose |
|----|----|
| `cov_from_biallelic()` | Genetic covariance from SNP allele counts and sample sizes |
| `cov_from_genetic_data()` | Genetic covariance from dosages, PCA scores, or allele calls |
| `dist_from_cov()` | Convert a covariance matrix to the matching distance matrix |
| `wishart_covariance()` | Measurement model for a covariance response; needs `nu` |
| `check_distance_response()` | Diagnose squared-Euclidean admissibility of a proposed generalized-Wishart response |
| `generalized_wishart()` | Generalized Wishart likelihood for an admissible squared-distance response |
| `simulate_covariance_response()` | Simulate a covariance matrix from known parameters |
| `terradish()` | Fit the model and return coefficients, standard errors, and AIC |

## See also

- **`vignette("getting-started", package = "terradish")`**, the core
  workflow using a distance response.
- **`vignette("simulation-design", package = "terradish")`**, how `nu`
  and sample size translate into statistical power before you collect
  data.
- **`vignette("ibe-ibr-workflow", package = "terradish")`**, adding
  environmental kernels to the Wishart likelihood with
  `wishart_covariates()`.
