---
title: "Covariance-Based IBR with wishart_covariance"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Covariance-Based IBR with wishart_covariance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4.5,
  warning  = FALSE,
  message  = FALSE
)

```

## Overview

The standard `terradish` workflow uses a pre-computed **distance matrix** (e.g.
F~ST~) as the genetic response. Two measurement models, `leastsquares` and
`mlpe`, accept such matrices and differ only in how they handle the
non-independence of pairwise observations.

When raw individual-level genotype data are available, you can retain a
population covariance summary rather than beginning from a precomputed distance
matrix:

1. Summarize the genetic data as a **covariance matrix** between populations.
2. Fit the resistance model with `wishart_covariance`, which evaluates the full
   Wishart likelihood on that covariance matrix using the effective marker
   degrees of freedom as the parameter `nu`.

Every off-diagonal element and the diagonal contribute to the covariance
likelihood. A lossless, consistently centered distance transformation can carry
the same information, so covariance input is not inherently superior. It is the
natural choice when the scientific summary is covariance and its diagonal is
meaningful.

| Model | Input `S` | Requires `nu` | Correlation structure |
|---|---|---|---|
| `leastsquares` | distance matrix | no | none |
| `mlpe` | distance matrix | no | shared-site |
| `generalized_wishart` | admissible squared-distance matrix | yes | generalized Wishart |
| `wishart_covariance` | **covariance matrix** | yes | Wishart |

This vignette covers:

- Converting different types of raw genetic data to a covariance matrix with
  `cov_from_genetic_data()` (and `cov_from_biallelic()` for biallelic SNPs).
- Setting up the landscape and conductance surface with the bundled *Melipona*
  data.
- Fitting an IBR model with `wishart_covariance`.
- Interpreting and comparing the fitted model.

---

## Preparing genetic covariance data

`terradish` provides two functions for turning raw genotype data into a
covariance matrix:

- **`cov_from_biallelic(Y, N)`**: fast path for biallelic SNP count matrices.
- **`cov_from_genetic_data(x, ...)`**: flexible path that handles biallelic
  dosage, multiallelic SNPs, microsatellite allele calls, PCA scores, or any
  other numeric encoding.

Both functions return a square, symmetric covariance matrix (one row/column per
population) that can be passed directly to `terradish` as the response `S`.

### Biallelic SNPs: `cov_from_biallelic()`

If your data are derived-allele counts at biallelic loci, use
`cov_from_biallelic()`:

```{r biallelic-example}
library(terradish)

# Simulated counts: 4 populations, 5 biallelic loci
# Y[i, l] = number of derived alleles in population i at locus l
# N[i, l] = total haploid chromosomes sampled (here, 20 diploids → N = 40)
set.seed(42)
n_pop <- 4
n_loci <- 5
Y <- matrix(sample(0:40, n_pop * n_loci, replace = TRUE),
            nrow = n_pop, ncol = n_loci,
            dimnames = list(paste0("pop", seq_len(n_pop)),
                            paste0("snp", seq_len(n_loci))))
N <- matrix(40L, nrow = n_pop, ncol = n_loci)

S_biallelic <- cov_from_biallelic(Y, N)
round(S_biallelic, 3)
```

### Numeric feature matrix: `cov_from_genetic_data()`

For dosage-coded SNPs, PCA scores, or any numeric feature matrix, use
`cov_from_genetic_data()` with `input = "features"` (the default).
The function centers and scales each feature, then applies Gower
double-centering to produce a covariance matrix:

```{r feature-example}
# Individual-level PCA scores: 12 individuals across 4 populations, 3 PCs
set.seed(7)
pca_scores <- matrix(rnorm(12 * 3), nrow = 12, ncol = 3,
                     dimnames = list(NULL, c("PC1", "PC2", "PC3")))
groups <- rep(c("pop1", "pop2", "pop3", "pop4"), each = 3)

S_pca <- cov_from_genetic_data(pca_scores, groups = groups)
round(S_pca, 3)
```

Setting `normalize = "features"` divides the covariance by the number of
retained features; this puts the scale on a per-feature basis rather than
summed across all features, which can make `tau` estimates easier to interpret
across datasets with different numbers of markers. This normalization changes
the scale of `S`; it does not automatically determine the best `nu` for every
genotyping system.

### Microsatellite allele calls: `cov_from_genetic_data()` with `input = "allele_calls"`

For diploid microsatellite data stored as two allele-call columns per locus,
set `input = "allele_calls"` and supply the `loci` argument to identify which
column belongs to which locus. The function expands each locus into per-allele
dosage columns before computing covariance:

```{r microsatellite-example}
# 9 individuals, 2 microsatellite loci, each stored in two allele-call columns
alleles <- data.frame(
  loc1_a = c(100, 100, 102, 102, 104, 104, 100, 102, 104),
  loc1_b = c(100, 102, 102, 104, 104, 100, 104, 100, 102),
  loc2_a = c(200, 202, 200, 202, 204, 204, 200, 204, 202),
  loc2_b = c(202, 202, 204, 204, 204, 200, 202, 200, 204)
)
groups_ms <- rep(c("popA", "popB", "popC"), each = 3)
loci_ms <- c("loc1", "loc1", "loc2", "loc2")

S_ms <- cov_from_genetic_data(
  alleles,
  groups  = groups_ms,
  input   = "allele_calls",
  loci    = loci_ms
)
round(S_ms, 3)
```

## Choosing `nu`

Wishart models treat `nu` as the degrees of freedom in the empirical covariance
matrix. It is fixed by the genetic data used to build `S`, and it is not
estimated by `terradish`.

For biallelic SNPs, the biological marker count and the statistical degrees of
freedom are usually the same: each retained SNP contributes one independent
standardized allele-frequency column. In that case, use the number of retained
polymorphic SNPs (reduced if markers are in linkage disequilibrium).

For microsatellites, **use the number of loci as the conservative primary value.**
Allele frequencies within a locus are correlated because they sum to a constant
equal to the ploidy, so the locus, not the individual allele, is the natural unit of
information for a conservative analysis.

A common alternative formula, `sum(K_l - 1)` where `K_l` is the number of
observed alleles at locus `l`, counts within-locus allele-frequency dimensions
more aggressively. Those dimensions are correlated. Treat this formula as a
larger sensitivity value, not as a known independent-marker count. Using it as
the primary `nu` can produce confidence intervals that are too narrow and
model-selection statistics that are inflated.

You can compute the larger value for a transparent sensitivity analysis:

```{r microsatellite-nu}
# Larger sensitivity value: sum(K_l - 1) across loci
msat_nu_upper <- sum(vapply(split(seq_along(loci_ms), loci_ms), function(j) {
  alleles_l <- unlist(alleles[, j, drop = FALSE], use.names = FALSE)
  length(unique(alleles_l[!is.na(alleles_l)])) - 1L
}, integer(1)))

n_loci <- length(unique(loci_ms))   # conservative primary value

cat("Primary value (number of loci):             nu =", n_loci, "\n")
cat("Larger sensitivity value (sum(K_l - 1)):   nu =", msat_nu_upper, "\n")
```

**Recommended practice:**

1. Set `nu = n_loci` for your primary analysis.
2. Report this choice in your methods section.
3. Repeat the key inference steps (confidence intervals and model comparison)
   with `nu = msat_nu_upper` to show how sensitive your conclusions are to the
   choice.
4. If the model rankings and qualitative conclusions are stable across the range,
   you can report that the results are robust to the specific choice of `nu`.

If only a published F~ST~ matrix is available and allele counts by locus are
unavailable, the locus count is the only reasonable option and is straightforwardly
conservative.

### Key arguments

| Argument | Default | Effect |
|---|---|---|
| `groups` | `NULL` | Population vector; if `NULL`, individual-level covariance is returned |
| `input` | `"features"` | `"features"` for numeric matrices; `"allele_calls"` for allele-call columns |
| `loci` | `NULL` | Required for `"allele_calls"`: maps each column to its locus |
| `center` | `TRUE` | Centers each feature by its global mean before covariance calculation |
| `scale` | `TRUE` | Scales each feature by its global standard deviation; drops constant features |
| `diagonal` | `"auto"` | `"within"` replaces the Gower diagonal with within-population variance (Dyer-style); `"gower"` leaves it as-is |
| `normalize` | `"none"` | `"features"` divides covariance by number of retained features |

The `diagonal = "within"` setting is useful for Dyer-style population-graph
workflows. Note, however, that replacing the diagonal can produce matrices that
are not positive definite for all datasets; check eigenvalues when in doubt:

```{r check-eigenvalues}
ev <- eigen(S_ms, symmetric = TRUE, only.values = TRUE)$values
all(ev > 0)  # TRUE → positive definite; FALSE → semi-definite or indefinite
```

---

## Set up the landscape

The rest of this vignette uses the bundled *Melipona* bee data from the
getting-started vignette. Steps are identical to those already covered there:
see `vignette("getting-started", package = "terradish")` for explanations.

```{r landscape-setup}
library(terra)

data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

covariates <- c(melip.altitude, melip.forestcover)
names(covariates) <- c("altitude", "forestcover")
covariates <- scale_covariates(covariates)

surface <- conductance_surface(covariates, melip.coords, directions = 8)

n_sites <- nrow(melip.coords)
cat("Number of sampling sites:", n_sites, "\n")
```

---

## Generate a covariance response for the demonstration

The bundled *Melipona* data ship with a pairwise F~ST~ matrix (`melip.Fst`), not
raw genotypes. To produce a realistic **covariance** matrix that reflects the
same landscape, we use `simulate_covariance_response()`. This function:

1. Computes the conductance-implied covariance matrix `E(theta)` at specified
   parameter values.
2. Forms the Wishart scale matrix `Sigma = tau * E + sigma * I`.
3. Draws `S ~ Wishart(nu, Sigma) / nu`.

```{r simulate-covariance}
# Use conductance parameters roughly consistent with the melip fit in the
# getting-started vignette (forestcover positive, altitude negative).
# nu = 1000 is an illustrative effective marker count for this simulation.
sim <- simulate_covariance_response(
  theta             = c(forestcover = 1.0, altitude = -0.5),
  formula           = ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  tau               = 1.0,
  sigma             = 0.5,
  nu                = 1000,
  seed              = 123
)

# This is what would come from cov_from_genetic_data() applied to real genotypes.
S_cov <- sim$covariance
dim(S_cov)
round(S_cov[1:5, 1:5], 4)
```

In a real analysis you would replace `S_cov` with the output of
`cov_from_genetic_data()` applied to your actual genotype data, for example:

```{r real-data-sketch, eval = FALSE}
# With biallelic SNP count matrices (Y = derived allele counts, N = haploid sample sizes)
S_cov <- cov_from_biallelic(Y, N)

# With individual dosage / PCA scores grouped by population
S_cov <- cov_from_genetic_data(dosage_matrix, groups = pop_vector,
                                normalize = "features")

# With microsatellite allele calls
S_cov <- cov_from_genetic_data(allele_df, groups = pop_vector,
                                input = "allele_calls",
                                loci  = locus_id_vector)
```

---

## Fit the model with `wishart_covariance`

The only differences from a standard `terradish` call are:

- Pass the **covariance matrix** `S_cov` as the left-hand side of the formula.
- Set `measurement_model = wishart_covariance`.
- Supply `nu` (the effective Wishart degrees of freedom for `S_cov`).

```{r fit-wishart-covariance}
fit_wc <- terradish(
  S_cov ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = wishart_covariance,
  nu                = 1000
)

summary(fit_wc)
```

### Understanding the nuisance parameters

The `wishart_covariance` measurement model fits two nuisance parameters (`phi`):

- **`tau`** (≥ 0): a nonnegative weight applied to the conductance-implied
  covariance `E`. If `tau` ≈ 0, there is no detectable graph-covariance signal
  under this response and model. Its magnitude is not comparable to the
  Gaussian distance-model `tau`, which is a log precision.
- **`sigma`** (log-scale): a scalar nugget added as `exp(sigma) * I` to the
  fitted covariance matrix. It captures diagonal variation not represented by
  the graph kernel, including sampling variation and model mismatch. It is not
  the Gaussian raster-smoothing scale and is not a direct estimate of drift or
  population size.

The fitted covariance model is:
> **Sigma = tau · E(theta) + exp(sigma) · I**

where `E(theta)` is the generalized inverse of the graph Laplacian evaluated at
the estimated conductance parameters `theta`.

---

## The role of `nu`: an effective sample size

`nu` is the most consequential setting in a covariance-based fit, and it
behaves in a way that surprises most users. The entire Wishart negative
log-likelihood is multiplied by `nu / 2`, so `nu` is a global multiplier of the
objective surface. That single fact has three separate consequences.

**1. The likelihood optimum does not depend on `nu`.** Scaling an objective by a
positive constant does not move its optimum, so the theoretical point estimates
of `theta`, `tau`, and `sigma` are invariant to `nu`. Separate numerical fits
should agree up to optimization tolerance:

```{r nu-invariance}
fit_nu <- function(nu) {
  f <- terradish(
    S_cov ~ forestcover + altitude,
    data              = surface,
    conductance_model = loglinear_conductance,
    measurement_model = wishart_covariance,
    nu                = nu
  )
  list(theta = coef(f),
       se    = sqrt(diag(solve(f$fit$hessian))))
}

f100   <- fit_nu(100)
f1000  <- fit_nu(1000)
f10000 <- fit_nu(10000)

# Point estimates agree across three orders of magnitude
rbind(`nu=100`   = f100$theta,
      `nu=1000`  = f1000$theta,
      `nu=10000` = f10000$theta)
```

**2. Standard errors scale as 1 / sqrt(`nu`).** The Hessian scales with `nu`, so
the asymptotic covariance scales with 1 / `nu` and every standard error scales
with 1 / sqrt(`nu`). Under the model, more approximately independent marker
information produces tighter intervals:

```{r nu-se}
rbind(`nu=100`   = f100$se,
      `nu=1000`  = f1000$se,
      `nu=10000` = f10000$se)

# Ratio confirms the 1/sqrt(nu) scaling: a 10x increase in nu
# shrinks SEs by sqrt(10) ~ 3.16
f100$se / f1000$se
```

**3. Model selection and hypothesis tests depend strongly on `nu`.** The
log-likelihood scales linearly with `nu`, but the AIC/BIC complexity penalty
(and the chi-squared reference distribution for a likelihood-ratio test) does
not. With large `nu` the penalty becomes negligible and extra covariates are
almost always retained; with small `nu`, parsimony dominates. The *same data*
can favor a simpler or a more complex model purely through the choice of `nu`.

The practical upshot: **`nu` is an effective sample size.** It does not change
*what* the data say about the shape of the conductance surface, but it controls
*how strongly* the data speak. Choose it carefully:

- For biallelic SNPs, use the number of retained (polymorphic) SNPs.
- For microsatellites, use the **number of loci** as a conservative default.
  Alleles within a locus are correlated (they sum to a constant), so
  `sum(K_l - 1)` over loci is a larger sensitivity value that counts
  within-locus dimensions more aggressively (see the "Choosing `nu`" section
  above).
- Markers in strong linkage disequilibrium are not independent, so the
  effective `nu` is **smaller** than the raw marker count. Thin to roughly
  unlinked loci, or use a deliberately conservative value.
- When `nu` is uncertain, report a brief sensitivity analysis: refit across a
  plausible range and check whether your intervals, likelihood evidence, and model
  rankings are stable.

---

## Recover the true parameters

Because we simulated these data with known parameters, we can check how well the
model recovers them. In practice this comparison is obviously not available.

```{r recovery}
cat("True conductance parameters:\n")
cat("  forestcover:", 1.0, "  altitude:", -0.5, "\n\n")

cat("Estimated conductance parameters:\n")
print(round(coef(fit_wc), 3))

cat("\nTrue nuisance parameters (on the variance scale):\n")
cat("  tau   =", sim$tau, "\n")
cat("  sigma =", sim$sigma, " (nugget variance; fitted model stores log(sigma))\n\n")

cat("Estimated nuisance parameters (phi vector, log scale for sigma):\n")
phi_est <- round(fit_wc$fit$phi, 3)
print(phi_est)

cat("\nFor comparison, exp(phi['sigma']) =",
    round(exp(fit_wc$fit$phi["sigma"]), 3),
    " vs. true sigma =", sim$sigma, "\n")
```

---

## Visualize the fit

### Observed vs. fitted covariance

```{r plot-fit, fig.cap = "***Observed vs. fitted genetic covariance.***"}
plot(fit_wc, type = "fit")
```

Each point represents a pair of sites. The x-axis is the fitted covariance from
the Wishart model; the y-axis is the observed covariance in `S_cov`. Points
near the diagonal indicate a good fit.

### Estimated conductance surface

```{r plot-surface, fig.cap = "***Estimated conductance surface with 95% CI bounds.***"}
plot(fit_wc, type = "surface", data = surface)
```

### Marginal associations on conductance

```{r plot-marginal, fig.cap = "***Marginal associations on the fitted conductance scale.***"}
plot(fit_wc, type = "marginal", data = surface)
```

Response-scale marginal plots are currently defined for regression-style
measurement models such as `mlpe`. For covariance-response models, inspect the
conductance-scale marginal associations alongside the observed-vs-fitted covariance
plot above.

If one covariate has weak focal support at its tails, apply focal-support
clamping during plotting and surface prediction:

```{r wishart-support-clamp, eval = FALSE}
plot(
  fit_wc,
  type = "marginal",
  data = surface,
  support = "focal",
  support_probs = c(0.01, 0.99),
  clamp_covariates = c("forestcover", "altitude")
)

cond_wc_focal <- conductance(
  surface,
  fit_wc,
  support = "focal",
  support_probs = c(0.01, 0.99),
  clamp_covariates = c("forestcover", "altitude")
)
```

---

## Modeling site-specific diagonal variance

The `sigma` nugget above is a **single scalar**: every site receives the same
diagonal variance. If the response and study design support a hypothesis about
site-specific diagonal heterogeneity, a covariate model can replace that common
nugget. This extension does not, by itself, separate movement, drift, sampling
variation, density, or effective size.

`wishart_drift_covariates()` replaces the scalar nugget with a **per-site
diagonal parameterized by site-level covariates**, producing a site-specific
diagonal-variance surface:

> **Sigma = tau · E(theta) + diag(exp(Z gamma))**

where `Z` is a site design matrix with an intercept and one column per supplied
covariate. The graph kernel and diagonal terms are estimated jointly. The
off-diagonal term is not a direct movement estimate, and the diagonal term is
not a direct estimate of local drift, effective size, abundance, or density.
Because the nugget is `exp(Z gamma)`, it is always positive and `Sigma` stays
positive definite. With no drift covariates (intercept only) the model reduces
**exactly** to `wishart_covariance`.

To demonstrate the parameterization, we simulate a covariance matrix whose
diagonal variance declines along a site covariate. The generating log-diagonal
slope is negative.

```{r drift-simulate}
# An illustrative site-level covariate for diagonal variance.
n_site <- length(surface$demes)
site_variance <- scale(seq_len(n_site))[, 1]

# True conductance-implied covariance E at the conductance truth used above.
E_true <- as.matrix(terradish_algorithm(
  loglinear_conductance(~ forestcover + altitude, surface$x),
  leastsquares, surface, S = diag(n_site),
  theta = c(forestcover = 1.0, altitude = -0.5),
  objective = FALSE, gradient = FALSE, hessian = FALSE, partial = FALSE
)$covariance)

# Per-site nugget: baseline exp(log(0.2)), slope -0.6 on the (centered) covariate.
Z <- cbind(1, scale(site_variance, scale = FALSE))
gamma_true <- c(log(0.2), -0.6)
nugget     <- as.vector(exp(Z %*% gamma_true))
Sigma_true <- 1.0 * E_true + diag(nugget)

set.seed(123)
nu_demo <- 500
S_drift <- rWishart(1, df = nu_demo, Sigma = Sigma_true / nu_demo)[, , 1]
```

Fit the diagonal-variance extension and, for comparison, the scalar-nugget model:

```{r drift-fit}
g_drift <- wishart_drift_covariates(site_variance, model = "wishart_covariance")

fit_drift <- terradish(
  S_drift ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = g_drift,
  nu                = nu_demo
)

fit_scalar <- terradish(
  S_drift ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = wishart_covariance,
  nu                = nu_demo
)

summary(fit_drift)
```

The `phi` table now contains `tau`, `sigma` (the baseline log-nugget, identical
in meaning to the scalar model's `sigma`), and `gamma_var1`, the slope of the
log-nugget on the site covariate.

```{r drift-interpret}
# Recovered log-diagonal slope (true value -0.6)
fit_drift$fit$phi["gamma_var1", 1]

# Compare only because the response, sites, graph, likelihood, and nu match.
aic_table(
  list(fit_drift, fit_scalar),
  mod_names = c("diagonal covariate", "scalar nugget")
)
```

A negative `gamma_var1` means that larger covariate values are associated with a
smaller fitted diagonal variance, conditional on the graph kernel and other
terms. A drift or effective-size interpretation would require an external model
that identifies how this genetic summary maps to those demographic quantities.

A few practical notes:

- The diagonal covariate is **site-level** (one value per focal site), supplied as
  a vector, data frame, or `terra::SpatRaster` plus `coords`, just like
  `wishart_covariates()`.
- `gamma` is on the **log-nugget** scale and is unconstrained in sign; only
  `tau` is constrained to be nonnegative.
- Adding diagonal covariates can change `theta` because the terms are estimated
  jointly. It does not guarantee protection from spatial or demographic
  confounding.
- This is the diagonal (within-site) companion to `wishart_covariates()`, which
  adds off-diagonal isolation-by-environment kernels. The two address different
  parts of `Sigma` and can be motivated by the same data.

---

## Preparing a generalized-Wishart squared-distance response

The generalized-Wishart likelihood expects a squared-distance response
${D = d^2}$, not an arbitrary symmetric dissimilarity matrix. For ${n}$ sites,
`terradish` forms the centered Gram matrix

$$B = -\frac{1}{2} H D H, \qquad
H = I - \frac{1}{n}\mathbf{1}\mathbf{1}^{\mathsf T},$$

and requires ${B}$ to be positive semidefinite within a relative numerical
tolerance. This geometry is the mathematical support condition for the
likelihood. It is separate from whether the selected genetic summary is
biologically appropriate for the sampling design and scientific question.

Use `check_distance_response()` on the matrix that you actually intend to fit:

```{r check-gw-response}
S_dist <- dist_from_cov(S_cov)
gw_check <- check_distance_response(S_dist)
gw_check[c("admissible", "relative_min_eigenvalue", "n_negative")]
```

**How to read the diagnostic:** `admissible = TRUE` means the implied centered
Gram matrix is positive semidefinite within tolerance. A very small negative
`relative_min_eigenvalue` can reflect floating-point rounding. `n_negative`
counts only eigenvalues that are more negative than the scaled tolerance.
`terradish()` performs this check automatically before any generalized-Wishart
fit and stops on substantive violations.

The checker does not select a genetic distance for you. Choose the response
based on the biology, marker type, sampling design, and intended interpretation,
then test the realized matrix. The name of a distance measure is not enough
because scaling, locus aggregation, transformations, and missing-data handling
can change the resulting geometry.

### Potentially relevant distance measures available in R

The table is a navigation aid, not a list of recommended or interchangeable
responses. Functions can differ in their scaling, input class, treatment of
missing data, and whether they operate on individuals or populations. Confirm
what the selected function returns. If it returns ordinary Euclidean distances
${d}$, square the matrix to obtain ${D = d^2}$ before calling
`check_distance_response()` or fitting `generalized_wishart`.

| Candidate measure | Example R implementations | Guidance for generalized Wishart |
|---|---|---|
| Edwards/angular and Cavalli-Sforza-Edwards chord distances | `adegenet::dist.genpop(method = 2)`, `poppr::edwards.dist()`, `hierfstat::genet.dist(method = "Dch")` | The `adegenet` manual labels method 2 Euclidean. Cavalli-Sforza-Edwards chord distance is one potentially relevant candidate, not a required or preferred response. Verify the exact implementation and square ordinary distances before checking. |
| Reynolds distance | `adegenet::dist.genpop(method = 3)`, `poppr::reynolds.dist()` | The `adegenet` manual labels its implementation Euclidean. Verify the realized matrix and its scaling. |
| Rogers distance | `adegenet::dist.genpop(method = 4)`, `poppr::rogers.dist()`, `hierfstat::genet.dist(method = "Dr")` | The `adegenet` manual labels method 4 Euclidean. Square ordinary distances and check the resulting matrix. |
| Nei distances | `adegenet::dist.genpop(method = 1)`, `poppr::nei.dist()`, variants from `hierfstat::genet.dist()` | These names cover different formulas. The `adegenet` manual labels method 1 non-Euclidean, so do not assume admissibility. |
| Prevosti distance | `adegenet::dist.genpop(method = 5)`, `poppr::prevosti.dist()`, `hierfstat::genet.dist(method = "Cp")` | The `adegenet` manual labels method 5 non-Euclidean. Do not assume admissibility. |
| Pairwise F~ST~ or F~ST~/(1 - F~ST~) | `hierfstat::pairwise.WCfst()`, `hierfstat::pairwise.neifst()` | These are common landscape-genetic responses, but they have no general squared-Euclidean guarantee. Check the realized matrix or use a Gaussian distance likelihood such as `mlpe`. |
| Euclidean, chord, or Hellinger distances on transformed numeric features | `stats::dist()`, `vegan::vegdist(method = "euclidean")`, `vegan::vegdist(method = "chord")`, `vegan::vegdist(method = "hellinger")` | Geometry depends on the feature representation and transformation. Scientific appropriateness remains study-specific. |

`ade4::is.euclid()` provides an optional independent diagnostic for ordinary
distances. Because it squares its input internally, the equivalent call for an
already squared generalized-Wishart response `D` is:

```{r ade4-cross-check, eval = FALSE}
# D is already squared, so take its square root for ade4::is.euclid().
ade4::is.euclid(as.dist(sqrt(D)))
```

Do not call `ade4::is.euclid(as.dist(D))` for this purpose because that tests
whether ${D^2}$, rather than ${D}$, has the required geometry. Functions such
as `ade4::cailliez()` and `ade4::lingoes()` can make an ordinary distance matrix
Euclidean for ordination, but `terradish` does not apply them automatically.
They change the genetic response and can change its biological interpretation.

---

## Compare with `generalized_wishart`

`generalized_wishart` uses the corresponding generalized Wishart formulation for
an admissible **squared-distance** representation of the same genetic data.
Here the response was obtained from the covariance matrix with
`dist_from_cov()` and checked above:

```{r compare-gw}
fit_gw <- terradish(
  S_dist ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = generalized_wishart,
  nu                = 1000
)

# Side-by-side parameter estimates
cat("wishart_covariance estimates:\n")
print(round(coef(fit_wc), 3))

cat("\ngeneralized_wishart estimates:\n")
print(round(coef(fit_gw), 3))
```

When the covariance and admissible squared-distance representations contain the same
information under identical centering, scaling, sites, and `nu`, their fitted
conductance parameters should agree up to numerical tolerance. The AIC
comparison is not meaningful across the two models because the responses differ.
Standard errors should only be compared after confirming that the transformation
is lossless and the effective information assumptions are identical.

---

## Choosing between covariance and distance input

| Consideration | `wishart_covariance` | `generalized_wishart` |
|---|---|---|
| Raw genotype data available | **Preferred**: use `cov_from_genetic_data()` | Requires an extra distance-conversion step |
| Only a pre-computed distance matrix is available | Requires converting to covariance (not always reversible) | Use only if the squared-distance response is admissible; otherwise use MLPE |
| Site-specific diagonal variance matters | Can retain and model the diagonal, subject to positive-semidefinite checks | Encoded only through an admissible squared-distance representation |
| Goal is a Dyer-style population graph | Diagonal from within-population variance is natural | Less natural |
| `nu` is known | Both models require `nu` | Both models require `nu` |

If you only have a distance matrix and no individual-level data, first verify
that it is admissible for the generalized Wishart representation. If that
condition is not defensible and effective `nu` is unavailable, use a Gaussian
distance likelihood such as `mlpe`. If you have raw genotypes and covariance is
the intended summary, start with `cov_from_genetic_data()` and
`wishart_covariance`.

---

## Quick-reference: the full workflow

```{r full-workflow, eval = FALSE}
library(terradish)
library(terra)

# 1. Load landscape data and build conductance surface
data(melip)
melip.altitude    <- terra::unwrap(melip.altitude)
melip.forestcover <- terra::unwrap(melip.forestcover)
melip.coords      <- terra::unwrap(melip.coords)

covariates <- c(melip.altitude, melip.forestcover)
names(covariates) <- c("altitude", "forestcover")
covariates <- scale_covariates(covariates)
surface <- conductance_surface(covariates, melip.coords, directions = 8)

# 2. Prepare the genetic covariance matrix from raw genotype data
#    (replace with your actual data and effective nu)

# Option A: biallelic SNPs (Y = derived allele counts, N = haploid totals)
S <- cov_from_biallelic(Y, N)

# Option B: individual-level numeric features grouped by population
S <- cov_from_genetic_data(dosage_matrix, groups = pop_vector,
                            normalize = "features")

# Option C: microsatellite allele calls
S <- cov_from_genetic_data(allele_df, groups = pop_vector,
                            input = "allele_calls",
                            loci  = locus_id_vector)

nu <- effective_marker_df  # justify, report, and evaluate sensitivity

# 3. Fit with wishart_covariance
fit <- terradish(
  S ~ forestcover + altitude,
  data              = surface,
  conductance_model = loglinear_conductance,
  measurement_model = wishart_covariance,
  nu                = nu
)

# 4. Summarize and visualize
summary(fit)
coef(fit)

plot(fit, type = "fit")
plot(fit, type = "surface", data = surface)
plot(fit, type = "marginal", data = surface)   # conductance-scale associations
```

## Summary of key functions

| Function | Purpose |
|----------|---------|
| `cov_from_biallelic()` | Genetic covariance from SNP allele counts and sample sizes |
| `cov_from_genetic_data()` | Genetic covariance from dosages, PCA scores, or allele calls |
| `dist_from_cov()` | Convert a covariance matrix to the matching distance matrix |
| `wishart_covariance()` | Measurement model for a covariance response; needs `nu` |
| `check_distance_response()` | Diagnose squared-Euclidean admissibility of a proposed generalized-Wishart response |
| `generalized_wishart()` | Generalized Wishart likelihood for an admissible squared-distance response |
| `wishart_drift_covariates()` | Add site-specific diagonal variance covariates |
| `simulate_covariance_response()` | Simulate a covariance matrix from known parameters |
| `terradish()` | Fit the model and return coefficients, standard errors, and AIC |

## See also

- **`vignette("getting-started", package = "terradish")`**, the core workflow
  using a distance response.
- **`vignette("simulation-design", package = "terradish")`**, how `nu` and
  sample size translate into statistical power before you collect data.
- **`vignette("ibe-ibr-workflow", package = "terradish")`**, adding
  environmental kernels to the Wishart likelihood with `wishart_covariates()`.
- **`vignette("hierarchical-conductance", package = "terradish")`**, adding a
  diagnostic residual field and evaluating sensitivity to spatial confounding.
